---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Python interoperability {#sec-bkg-python-interoperability}

## Introduction

Ideally, users would be able to leverage the full range of available tools for their analysis. Methods should be selected based on scientific merit (ideally demonstrated by neutral benchmarks), and independent of having been implemented in a given programming language (R or Python) or framework (e.g. Bioconductor or `Seurat`).

[R's application to statistical analysis of spatial data dates back decades, primarily in epidemiological and geospatial research. As a result, various tools for spatial analyses have been established. For example, `r BiocStyle::CRANpkg("sp")` provides a "coherent set of classes and methods for \[...\] points, lines, polygons, and grids"; `r BiocStyle::CRANpkg("spatstat")` and, more recently, `r BiocStyle::CRANpkg("sf")` provide tools for spatial point pattern and vector data, respectively.]{.aside} For single-cell and spatial omics data analysis, being able to leverage different ecosystems is especially powerful. On the one hand, Python offers superb infrastructure for image analysis and machine learning-based approaches. On the other hand, the R programming language has been historically dedicated to statistical computing; as a result, many modern R methods for spatial omics data build on a solid foundation of tools for spatial statistics and statistical modeling in general.

Different data structures, although standardized within a given framework, make switching between languages and tools somewhat cumbersome. In the realm of single-cell and spatial omics, all Bioconductor tools are built around `r BiocStyle::Biocpkg("SummarizedExperiment")`-derived classes, while `r BiocStyle::CRANpkg("Seurat")`, `r BiocStyle::CRANpkg("Giotto")` and `r BiocStyle::Githubpkg("BIMSBbioinfo/VoltRon")` each rely on their own object definitions. In Python, [`scanpy`](https://scanpy.readthedocs.io/) [@Wolf2018-Scanpy] and [`squidpy`](https://squidpy.readthedocs.io/) [@Palla2022-Squidpy] use [`AnnData`](https://anndata.readthedocs.io/). Attempts to alleviate the problem are being made -- e.g. `r BiocStyle::Biocpkg("zellkonverter")` and functions from `r BiocStyle::CRANpkg("Seurat")` allow for conversion between Python's `AnnData`, and R/Bioconductor's `r BiocStyle::Biocpkg("SingleCellExperiment")` or `r BiocStyle::Biocpkg("SpatialExperiment")`.

[[Quarto](https://quarto.org/) is the successor to [R Markdown](https://rmarkdown.rstudio.com/) (by Posit, formerly known as RStudio). Similar to *.Rmd* files, *.qmd* files can include scientific content (e.g. cross-referencing, LaTeX-based equations), and can be published in multiple output formats (HTML, PDF, etc.). This book is built using Quarto.]{.aside} On a higher level, tools that enable interoperability between programming languages have become available. For example, the `r BiocStyle::Biocpkg("reticulate")` package provides an R interface to Python, including support to translate between objects from both languages; `r BiocStyle::Biocpkg("basilisk")` facilitates Python environment management within the Bioconductor ecosystem, and can also be interfaced with `reticulate`; and Quarto can generate dynamic reports from code in different languages.


## Dependencies

```{r libraries, message=FALSE, warning=FALSE}
#| code-fold: false
library(OSTA.data)
library(SpatialExperiment)
library(VisiumIO)
library(zellkonverter)
library(jsonlite)
library(reticulate)
library(basilisk)
```


## Configuring Python

We first create an environment with Python modules necessary for the examples in this chapter. For that, we first define a `r BiocStyle::Biocpkg("basilisk")` virtual environment and use the resulting environment with `r BiocStyle::Biocpkg("reticulate")`.

```{r env, warning=FALSE, results="hide"}
#| code-fold: false
env <- BasiliskEnvironment(
  envname = "bkg-interop", 
  pkgname = "base", 
  channels = c("conda-forge", "bioconda", "pytorch"), 
  packages = c("python=3.11", "Dask=2024.12.1"), 
  pip = c("ome-zarr==0.10.2", "zarr==2.18.7","squidpy==1.6.2")
)


# use_virtualenv(obtainEnvironmentPath(env))
use_condaenv(obtainEnvironmentPath(env))
```


## SingleCellExperiment

### Calling Python

After configuring the Python environment, R commands can now be `reticulate`d as follows: [Note that running code in this way comes with a small overhead of starting up a Python session in the background. But this is typically small compared to the runtime required to run computation-heavy methods, or when analyzing large-scale single-cell and spatial data (hundreds of thousands of cells or more).]{.aside}

```{r load-sce}
#| code-fold: false
anndata <- reticulate::import("anndata")
example_h5ad <- system.file("extdata", "krumsiek11.h5ad", 
                            package = "zellkonverter")
(ad <- anndata$read_h5ad(example_h5ad))
```

### Continuing in R

We can access *any* of the variables above in R. For basic outputs, this works out of the box:

```{r uniq-celltype}
#| code-fold: false
unique(ad$obs$cell_type)
```

`reticulate` also supports a few direct [type conversions](https://rstudio.github.io/reticulate/#type-conversions) (e.g. dictionary $\leftrightarrow$ named list). In the example demonstrated here, we use `zellkonverter` to convert from `AnnData` to `SingleCellExperiment`:

```{r anndata2sce, warning=FALSE}
#| code-fold: false
(sce <- AnnData2SCE(ad))
```

### Back to Python

We can also do the reverse, i.e. go from R's `SingleCellExperiment` to Python's `AnnData`:

```{r sce2anndata}
#| code-fold: false
(ad <- SCE2AnnData(sce, X_name = "X"))
```


## SpatialExperiment

Since the `SpatialExperiment` class extends `SingleCellExperiment` (see @sec-bkg-data-infrastructure), conversion operations that we discussed above are also applicable to `SpatialExperiment`. However, to accomplish a full conversion from the `AnnData` object, we need to manually insert the spatial information using `reticulate` directly. 

### Starting with R

For this use case with `SpatialExperiment`, we will use the dataset from @Janesick2023-high-res, which includes Visium measurements on human breast cancer tissue.

```{r load-spe, message=FALSE, warning=FALSE, results="hide"}
#| code-fold: false
id <- "Visium_HumanBreast_Janesick"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir = td)
obj <- TENxVisium(
  spacerangerOut = td, 
  format = "h5", 
  images = "lowres")
(spe <- VisiumIO::import(obj))
```

We also need to parse the original scaling information (i.e. scale factor) for spots and images available in the standard Visium output. We will use this later during conversion.

```{r scalefactors, message=FALSE, results="hide"}
#| code-fold: false
scalefactors <- jsonlite::read_json(file.path(td, "outs/spatial/scalefactors_json.json"))
```

We again use the `SCE2AnnData` function from the previous example, and convert the `SingleCellExperiment`-relevant components of the `SpatialExperiment` object to an `AnnData` object.

```{r spe2anndata, message=FALSE}
#| code-fold: false
(ad <- SCE2AnnData(spe, X_name = "counts"))
```

We can now populate the `uns` and `obsm` components of the `AnnData` object with spatial coordinates and images. We start with the coordinates.

```{r coords}
#| code-fold: false
coords <- spatialCoords(spe)
colnames(coords) <- c("x", "y")
obsm <- list(spatial = coords)
```

Now let's create the `uns` component. The list of `uns` should compose of as many samples as the images in the `SpatialExperiment` object. Also, each sample entry in the list should have two elements, one for the image and the other for the scaling information.

```{r uns}
#| code-fold: false
# get image metadata
imgdata <- imgData(spe)

# get image
img <- imgRaster(spe)
img <- apply(img, c(1, 2), \(x) col2rgb(x))
img <- aperm(img, perm = c(2, 3, 1))
img <- img / 255

# create uns
uns <- list(
  spatial = setNames(list(NULL), imgdata$sample_id)
)
uns[["spatial"]][[imgdata$sample_id]] <- 
  list(images = list(lowres = img), 
       scalefactors = scalefactors)
```

Now let's insert the components to the `AnnData` object, and write back to an h5ad file.

```{r writeh5ad}
#| code-fold: false
ad$obs$library_id <- imgdata$sample_id
ad$obsm <- obsm
ad$uns <- uns
ad$write_h5ad("spe.h5ad")
```

### Calling Python

Now that we have converted the `SpatialExperiment` object to `AnnData` format, we visualize the Visium data using, for example, the `squidpy` module.

```{python modules}
#| code-fold: false
#| output: false
import squidpy as sq
import anndata as ad
import matplotlib.pyplot as plt
```

Now, we read `spe.h5ad` and visualize features or gene expression (e.g. ERBB2).

```{python spatial-scatter}
#| code-fold: false
#| include: true
#| warning: false
#| message: false
#| error: false
#| output: true
adata = ad.read_h5ad("spe.h5ad")
adata.var_names = adata.var['Symbol'].astype(str)
sq.pl.spatial_scatter(adata, 
                      color = ["ERBB2"], 
                      img_res_key = "lowres")

# plt.show()  # display plot if running code interactively
```

```{r, include=FALSE}
file.remove("spe.h5ad")
```


## Appendix

### References {.unnumbered}
