# Python interoperability {#sec-bkg-python-interoperability}

{{< include _start.qmd >}}

## Introduction

Methods discussed in this book are usually available as either R or Python packages. Ideally, users should be able to leverage the full range of available tools for their analyses. Methods should be selected based on scientific merit (ideally demonstrated by neutral benchmarks), and independent of having been implemented in a given programming language (R or Python) or framework (e.g. Bioconductor or `Seurat`).

[R's application to statistical analysis of spatial data dates back decades, primarily in epidemiological and geospatial research. As a result, various tools for spatial analyses have been established. For example, `r BiocStyle::CRANpkg("sp")` provides a "coherent set of classes and methods for \[...\] points, lines, polygons, and grids"; `r BiocStyle::CRANpkg("spatstat")` and, more recently, `r BiocStyle::CRANpkg("sf")` provide tools for spatial point pattern and vector data, respectively.]{.aside} For single-cell and spatial omics data analysis, being able to leverage different ecosystems is especially powerful. On the one hand, Python offers superb infrastructure for image analysis and machine learning-based approaches. On the other hand, the R programming language has been historically dedicated to statistical computing; as a result, many modern R methods for spatial omics data build on a solid foundation of tools for spatial statistics and statistical modeling in general.

Different data structures, although standardized within a given framework, make switching between languages and tools somewhat cumbersome. In the realm of single-cell and spatial omics, all Bioconductor tools are built around `r BiocStyle::Biocpkg("SummarizedExperiment")`-derived classes, while [Seurat](https://satijalab.org/seurat/) [@Hao2023-Seurat], [Giotto](https://giottosuite.com/) [@Chen2025-Giotto-Suite], and `r BiocStyle::Githubpkg("BIMSBbioinfo/VoltRon")` each rely on their own object definitions. In Python, [Scanpy](https://scanpy.readthedocs.io/) [@Wolf2018-Scanpy] and [Squidpy](http://squidpy.readthedocs.io/) [@Palla2022-Squidpy] use [AnnData](https://anndata.readthedocs.io/). Attempts to alleviate the problem are being made -- e.g. `r BiocStyle::Biocpkg("zellkonverter")`, [anndataR](https://github.com/scverse/anndataR) [@Deconinck2025-anndataR], and functions from `Seurat` allow for conversion between Python's `AnnData` and R/Bioconductor's `r BiocStyle::Biocpkg("SingleCellExperiment")` or `r BiocStyle::Biocpkg("SpatialExperiment")`.

[[Quarto](https://quarto.org/) is the successor to [R Markdown](https://rmarkdown.rstudio.com/) (by Posit, formerly known as RStudio). Similar to *.Rmd* files, *.qmd* files can include scientific content (e.g. cross-referencing, LaTeX-based equations), and can be published in multiple output formats (HTML, PDF, etc.). This book is built using Quarto.]{.aside} On a higher level, tools that enable interoperability between programming languages have become available. For example, [reticulate](https://rstudio.github.io/reticulate/) provides an R interface to Python, including support to translate between objects from both languages; `r BiocStyle::Biocpkg("basilisk")` facilitates Python environment management within the Bioconductor ecosystem, and can also be interfaced with `reticulate`; and Quarto can generate dynamic reports from code in different languages.

In this chapter, we demonstrate examples showing how to set up Python environments and interact with `anndata` objects in R, using either a Conda environment or virtual environment, together with [reticulate](https://rstudio.github.io/reticulate/), `r BiocStyle::Biocpkg("basilisk")`, and/or `r BiocStyle::Biocpkg("zellkonverter")`.


## Dependencies

```{r libraries, message=FALSE, warning=FALSE}
#| code-fold: false
library(SpatialExperiment)
library(VisiumIO)
library(OSTA.data)
library(jsonlite)
library(reticulate)
library(basilisk)
library(zellkonverter)
```


## Conda environment and reticulate

In the first example, we set up a Conda environment and use [reticulate](https://rstudio.github.io/reticulate/) to interact with Python `anndata` objects in R. This is the simplest option, and is the recommended starting point for most users. Note that we are using functions from the `reticulate` package in the code below.

To set up the Conda environment, we also need to specify which Python installation to use. In order to run this example programmatically to compile this book chapter, we install a new Python installation with [Miniconda](https://www.anaconda.com/docs/getting-started/miniconda/main). However, if you are following this example on a laptop or another system, it is easier to specify an existing Python installation to use. This will help avoid cluttering your system with additional Python installations. This option is noted in the code comments below.


### Install Miniconda

_Note: skip this section if you are using an existing Python installation instead._

In order to build this book chapter programmatically, we install a new Python installation with [Miniconda](https://www.anaconda.com/docs/getting-started/miniconda/main), which is a free, lightweight installation of Python, Conda, and a small number of packages, provided by Anaconda. Note that we also need to set some conda settings using environment variables to run this code programmatically, which can also be skipped if you are installing Miniconda interactively.

```{r conda-settings, warning=FALSE, message=FALSE, results="hide"}
# use Miniforge (conda-forge only) instead of Anaconda's Miniconda
Sys.setenv(
  RETICULATE_MINICONDA_URL = 
    "https://github.com/conda-forge/miniforge/releases/latest/download/Miniforge3-Linux-x86_64.sh"
)

# set conda channels using temporary .condarc file
condarc_txt <- "
channels:
  - conda-forge
  - bioconda
channel_priority: strict
"
condarc_tmp <- tempfile("condarc-")
writeLines(trimws(condarc_txt, which="left"), condarc_tmp)
Sys.setenv(CONDARC=condarc_tmp)

# conda settings for terms of service
Sys.setenv(CONDA_PLUGINS_AUTO_ACCEPT_TOS="yes")
Sys.setenv(CI="true")
```

```{r install-miniconda, warning=FALSE, message=FALSE, results="hide"}
# install Miniconda (or update if already installed)
flag <- dir.exists(miniconda_path())
if (flag) miniconda_update() else install_miniconda()
```


### Set up Conda environment

Next, we create a Conda environment and install some key dependencies.

If you are using an existing Python installation on your system (instead of installing Miniconda), this can be specified with the argument `conda = ...` -- i.e. specify the path to the `conda` binary from the existing Python installation to use.

```{r conda-create, warning=FALSE, message=FALSE, results="hide"}
# create conda environment
env <- "py-interop"
# specify 'conda=...' in 'conda_create()'
# to use an existing Python installation
conda_create(env, packages="python=3.11")  
use_condaenv(env, required=TRUE)
```

```{r conda-list}
# list conda environments
conda_list()$name
```

The following dependencies are needed in some of the following chapters, and need to be installed from `conda-forge`. This can also be skipped if you are only running the examples in this chapter. We install these dependencies here so that we can re-use this Conda environment again in later chapters.

```{r deps-conda-forge, warning=FALSE, message=FALSE, results="hide"}
# install dependencies that need to be 
# installed from conda-forge (not PyPI)
conda_install(
    envname=env, 
    channel=c("conda-forge", "nodefaults"),
    packages=c("libcxx", "llvm-openmp", "llvmlite", "numba", 
               "proj>=9.4", "pyproj>=3.7"))
```

The following dependencies are necessary for the examples in this chapter. These are installed from PyPI.

```{r deps-pypi, warning=FALSE, message=FALSE, results="hide"}
# install Python packages from PyPI
conda_install(
    pip=TRUE, 
    envname=env, 
    packages=c("Dask==2024.12.1", "zarr==2.18.7", "squidpy==1.6.2"))
```

The following option sets Python code chunks to run via `reticulate`, allowing this chapter to be compiled programmatically.

```{r python-reticulate}
# set to run {python} code chunks via reticulate
knitr::opts_chunk$set(python.reticulate = TRUE)
```


### SingleCellExperiment {#sec-bkg-python-interop-sce}

#### Calling Python

After configuring the Python environment, R commands can now be run using `reticulate` as follows. For more details on the syntax used, see the [reticulate](https://rstudio.github.io/reticulate/) documentation. [Note that running code in this way comes with a small overhead of starting up a Python session in the background. But this is typically small compared to the runtime required to run computation-heavy methods, or when analyzing large-scale single-cell and spatial data (hundreds of thousands of cells or more).]{.aside}

```{r load-sce}
h5ad <- system.file("extdata", "krumsiek11.h5ad", package="zellkonverter")
anndata <- import("anndata")
(ad <- anndata$read_h5ad(h5ad))
```

#### Continuing in R

We can access any of the variables above in R. For basic outputs, this works out of the box:

```{r uniq-celltype}
unique(ad$obs$cell_type)
```

`reticulate` also supports a few direct [type conversions](https://rstudio.github.io/reticulate/#type-conversions) (e.g. dictionary $\leftrightarrow$ named list). In the example demonstrated here, we use `zellkonverter` to convert from `AnnData` to `SingleCellExperiment`:

```{r anndata2sce, message=FALSE}
(sce <- AnnData2SCE(ad))
```

#### Back to Python

We can also do the reverse, i.e. go from R's `SingleCellExperiment` to Python's `AnnData`:

```{r sce2anndata}
(ad <- SCE2AnnData(sce, X_name="X"))
```


### SpatialExperiment

Since the `SpatialExperiment` class extends `SingleCellExperiment` (see @sec-bkg-infrastructure), conversion operations that we discussed above are also applicable to `SpatialExperiment`. However, to accomplish a full conversion from the `AnnData` object, we need to manually insert the spatial information using `reticulate` directly.

#### Starting with R

For this use case with `SpatialExperiment`, we will use the dataset from @Janesick2023-high-res, which includes Visium measurements on human breast cancer tissue.

```{r load-spe, message=FALSE, warning=FALSE, results="hide"}
id <- "Visium_HumanBreast_Janesick"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir = td)
obj <- TENxVisium(
    spacerangerOut=file.path(td, "outs"), 
    format="h5", 
    images="lowres")
(spe <- VisiumIO::import(obj))
```

We also need to parse the original scaling information (i.e. scale factor) for spots and images available in the standard Visium output. We will use this later during conversion.

```{r scalefactors, message=FALSE, results="hide"}
sfs <- jsonlite::read_json(file.path(td, "outs/spatial/scalefactors_json.json"))
```

We again use the `SCE2AnnData` function from `zellkonverter` from the previous example, and convert the `SingleCellExperiment`-relevant components of the `SpatialExperiment` object to an `AnnData` object.

```{r spe2anndata, message=FALSE}
(ad <- SCE2AnnData(spe, X_name="counts"))
```

We can now populate the `uns` and `obsm` components of the `AnnData` object with spatial coordinates and images. We start with the coordinates.

```{r coords}
spatialCoordsNames(spe) <- c("x", "y")
obsm <- list(spatial=spatialCoords(spe))
```

Now let's create the `uns` component. The list of `uns` should be composed of as many samples as the images in the `SpatialExperiment` object. Also, each sample entry in the list should have two elements, one for the image and the other for the scaling information.

```{r uns}
# get image metadata
imgdata <- imgData(spe)

# get image
img <- imgRaster(spe)
img <- apply(img, c(1, 2), \(x) col2rgb(x))
img <- aperm(img, perm=c(2, 3, 1))
img <- img / 255

# create uns
uns <- list(images=list(lowres=img), scalefactors=sfs)
uns <- list(spatial=setNames(list(uns), imgdata$sample_id))
```

Now let's insert the components to the `AnnData` object, and write back to an `.h5ad` file.

```{r writeh5ad}
ad$obs$library_id <- imgdata$sample_id
ad$obsm <- obsm
ad$uns <- uns
ad$write_h5ad("spe.h5ad")
```

#### Calling Python

Now that we have converted the `SpatialExperiment` object to `AnnData` format, we can run Python code using the `AnnData` object. Here, for example, we visualize the Visium data using the `squidpy` module.

```{python modules}
#| output: false
import squidpy as sq
import anndata as ad
import matplotlib.pyplot as plt
```

Read in `spe.h5ad` and visualize features or gene expression (e.g. ERBB2):

```{python spatial-scatter}
#| include: true
#| warning: false
#| message: false
#| error: false
#| output: true
adata = ad.read_h5ad("spe.h5ad")
adata.var_names = adata.var['Symbol'].astype(str)
sq.pl.spatial_scatter(adata, 
                      color = ["ERBB2"], 
                      img_res_key = "lowres")
# plt.show() # needed to display plot if running code interactively
```

```{r, include=FALSE}
# clean up by removing the `spe.h5ad` file
file.remove("spe.h5ad")
```


## basilisk

Alternatively, we can use `r BiocStyle::Biocpkg("basilisk")` to install and manage Python environments. This is an alternative to setting up a Conda environment with an existing Python installation or a new installation of Miniconda, as demonstrated above. Using `basilisk` provides a self-contained environment, which helps avoid problems due to interactions with other environments on your system, and provides a higher level of reproducibility. This is also especially useful for package development by advanced users. However, setup may be more challenging in some cases.


### Configuring Python

In this example, we use `r BiocStyle::Biocpkg("basilisk")` to create an environment containing a Python installation and all necessary dependencies for this chapter, and then use the resulting environment with `reticulate`, as above.

```{r env, warning=FALSE, message=FALSE, results="hide", eval=FALSE}
# set up environment using basilisk
env_basilisk <- BasiliskEnvironment(
    pkgname="base", 
    envname="env-basilisk", 
    pip=c("zarr==2.18.7", "squidpy==1.6.2"),
    packages=c("python=3.11", "Dask=2024.12.1")) 

# use virtual environment
use_virtualenv(obtainEnvironmentPath(env_basilisk))
```


### SingleCellExperiment

Now that we have set up the Python environment using `basilisk`, we can run R commands using `reticulate` in the same way as in the previous example. For example, using the same code as above:

```{r load-sce-basilisk, eval=FALSE}
h5ad <- system.file("extdata", "krumsiek11.h5ad", package="zellkonverter")
anndata <- import("anndata")
(ad <- anndata$read_h5ad(h5ad))
```

To continue the example, please follow the code from the previous section above, starting from @sec-bkg-python-interop-sce.


## Appendix

Following are links to several key packages and tools relating to 
R-Python interoperability, which were mentioned in the sections above:

- `r BiocStyle::Githubpkg("scverse/anndataR")` [@Deconinck2025-anndataR]: 
R package and community project to work with `AnnData` objects in R, including 
conversion to and from `SingleCell/SpatialExperiment` and `Seurat` objects

- `r BiocStyle::Biocpkg("zellkonverter")` [@Zappia2020-zellkonverter]: 
R package to convert between `AnnData` and `SingleCellExperiment` objects, 
as well as reading from and writing to H5AD

- [reticulate](https://rstudio.github.io/reticulate/) [@Ushey2017-reticulate]: 
R package and framework to call Python and run Python code in R, translate 
between R and Python objects, and manage virtual and Conda environments

- `r BiocStyle::Biocpkg("basilisk")` [@Lun2022-basilisk]: 
R package to install and manage Python environments in R packages and sessions


### References {.unnumbered}

{{< include _stop.qmd >}}
