# Importing data {#sec-bkg-importing}

## Flat file structure

At present, the file structure and formats of data from spatial transcriptomics platforms varies between commercial providers.
Nevertheless, all data are similar in their essence, e.g.: sequencing-based data include spatial locations of array spots and a count matrix; imaging-based data include transcript locations (from spot-calling), polygon boundaries (from segmentation), and a count matrix (from allocating transcripts to cells), etc.


### Visium (10x Genomics)

Running `spaceranger` on Visium data creates a set of standardized output files. These comprise raw measurement data (similar to scRNA-seq but including, e.g., spot-level coordinates, potential images), as well as results from a 'button-push' analysis pipeline that includes standard quality control, dimension reduction (PCA, t-SNE, and UMAP), graph-based clustering, etc. 
The resulting outputs are well described [here](https://www.10xgenomics.com/support/software/space-ranger/latest/analysis/outputs/output-overview); briefly:

```{r, eval=FALSE}
Visium
  └── outs
    ├── spatial
      ├── tissue_positions_list.csv # spot locations
      └── tissue_lowres_image.png   # same-section H&E
    └── filtered_feature_bc_matrix  # in-tissue matrix files
    └── raw_feature_bc_matrix       # unfiltered matrix files
         ├── barcodes.tsv # spot barcodes (i.e., sequences)
         ├── features.tsv # gene metadata (e.g., ensembl IDs)
         └── matrix.mtx   # (gene x spot) count matrix
```


### Visium HD (10x Genomics)

Running `spaceranger` on Visium HD data generates outputs similar to the above, but by default includes outputs binned at a resolution of 2, 8, and 16 µm. Thus, outputs have a hierarchical structure where each `binned_outputs/` subdirectory contains files analogous to the `outs/` directory for Visium; e.g.:

```{r, eval=FALSE}
VisiumHD
  └── binned_outputs
    └─── square_002um
      └── filtered_feature_bc_matrix.h5
      └── filtered_feature_bc_matrix
        └── barcodes.tsv.gz
        └── features.tsv.gz
        └── matrix.mtx.gz
      └── raw_feature_bc_matrix.h5
      └── raw_feature_bc_matrix
        └── ...
      └── spatial
        └── tissue_positions.parquet
        └── ...
    └── square_*
```


### Xenium (10x Genomics)

Running `xeniumranger` reduces raw data generated from Xenium runs to an output bundle with standardized file structure and contents. Notably, 10x Genomics provides a variety of output file formats (for example, both *.csv* and *.parquet* for gene/cell metadata), facilitating interoperability with a variety of frameworks.
All outputs are described in detail [here](https://www.10xgenomics.com/support/software/xenium-onboard-analysis/latest/analysis/xoa-output-understanding-outputs); briefly:

```{r, eval=FALSE}
Xenium
  └── outs 
    ├── cells.parquet          # cell metadata (e.g., area)
    ├── cell_feature_matrix.h5 # compressed format of the below 
    └── cell_feature_matrix    # segmentation-derived matrix files
      ├── barcodes.tsv # cell barcodes (i.e., sequences)
      ├── features.tsv # gene metadata (e.g., target type)
      └── matrix.mtx   # (gene x cell) count matrix
    ├── transcripts.parquet        # molecule locations
    ├── cell_boundaries.parquet    # membrane segmentation
    ├── nucleus_boundaries.parquet # nuclear segmentation 
    └── experiment.xenium # experiment-wide metadata (in .json format)
```


### CosMx (NanoString)

Through [custom module scripts](https://github.com/Nanostring-Biostats/CosMxDACustomModules), the [AtoMx](https://nanostring.com/products/atomx-spatial-informatics-platform/atomx-sip-overview/) Spatial Informatics Portal (SIP) allows exporting different types of objects and, importantly, 'flat' (human-readable) file formats. Unlike raw data (e.g. images prior to spot-calling), these represent processed outputs (e.g. spatial locations of segmentation boundary vertices and molecules, segmentation-derived count matrix, etc.). A detailed description of these files is given [here](https://nanostring-biostats.github.io/CosMx-Analysis-Scratch-Space/posts/flat-file-exports/flat-files-compare.html); briefly:

```{r, eval=FALSE}
CosMx 
  ├── exprMat_file.csv       # (gene x cell) counts
  ├── fov_positions_file.csv # FOV corner positions
  ├── metadata_file.csv      # cell-level metadata 
  ├── polygons.csv           # segmentation boundaries
  └── tx_file.csv            # molecule locations
```


## Reading into R

Reader functions from several packages can be used to import data from raw files into a `SpatialExperiment` object (or derivatives thereof), including:

- `r BiocStyle::Biocpkg("VisiumIO")` provides readers for spatial data from 10x Genomics's Space Ranger pipeline, i.e. Visium and Visium HD. This includes support for *.mtx*, *.tar.gz* and *.h5* file formats, and for reading in multiple samples at once. Data are read into a `SpatialExperiment` object.

- `r BiocStyle::Biocpkg("XeniumIO")` provides functions to import 10x Genomics Xenium data into R. Notably, there is support for multiple file formats (e.g. *.h5* and *.mtx* for count data, *.parquet* and *.csv* for polygons and molecules, etc.), as well as automated distinction between RNA targets and other barcodes (e.g. negative probes, blank codes, etc.).

- `r BiocStyle::Githubpkg("SpatialExperimentIO")` provides readers for a variety of imaging-based spatial transcriptomics platforms, including CosMx (Bruker), Xenium (10x Genomics), MERSCOPE (Vizgen), and seqFISH (Spatial Genomics). Data may be read into a `SingleCellExperiment` or `SpatialExperiment`.

- `r BiocStyle::Biocpkg("SpatialFeatureExperiment")` provides functions to read CosMx, Xenium, MERSCOPE and Visium(HD) as `SpatialFeatureExperiment` objects.


## Appendix

### References {.unnumbered}
