# Interoperability {#sec-bkg-interoperability}

## Introduction

Ideally, users would be able to leverage the full range of available tools for their analysis. Methods should be selected based on scientific merit (as demonstrated by third-party benchmarks), and independent of having been implemented in a given programming language (R or Python) or framework (say, Bioconductor or `Seurat`).

[R's application to spatial data dates back decades, primarily in epidemiological and geospatial research. As a result, various tools for spatial analyses have been established; for example, `r BiocStyle::CRANpkg("sp")` provides a "coherent set of classes and methods for \[...\] points, lines, polygons, and grids"; `r BiocStyle::CRANpkg("spatstat")` and, more recently, `r BiocStyle::CRANpkg("sf")` provide tools for spatial point pattern and vector data, respectively.]{.aside} For single-cell and spatial omics data analysis, being able to leverage different ecosystems is especially powerful. On the one hand, Python offers superb infrastructure for image analysis and machine learning-based approaches. On the other hand, the R programming language has been historically dedicated to statistical computing; as a result, many a modern R method for spatial omics data builds on a solid foundation of tools for spatial statistics and statistical modeling in general.

Different data structures, although standardized within a given framework, make switching between languages and tools especially cumbersome. In the realm of single-cell and spatial omics, Bioconductor tools build around `r BiocStyle::Biocpkg("SummarizedExperiment")`-derived classes, while `r BiocStyle::CRANpkg("Seurat")`, `r BiocStyle::CRANpkg("Giotto")` and `r BiocStyle::Githubpkg("BIMSBbioinfo/VoltRon")` rely on objects built in-house; and Python's [`scanpy`](https://scanpy.readthedocs.io/) [@Wolf2018-scanpy] and [`squidpy`](https://squidpy.readthedocs.io/) [@Palla2022-squidpy] use [`AnnData`](https://anndata.readthedocs.io/). Attempts to alleviate the problem are being made; e.g. `r BiocStyle::Biocpkg("zellkonverter")` and functions from `r BiocStyle::CRANpkg("Seurat")` allow for conversion between Python's `AnnData` and R/Bioconductor's `r BiocStyle::Biocpkg("SingleCellExperiment")`.

[[Quarto](https://quarto.org) is the successor of [R Markdown](https://rmarkdown.rstudio.com) by Posit (formerly known as RStudio). Similar to *.rmd* files, *.qmd* files can include scientific content (e.g. cross-referencing, LaTeX-based equations), and can be published in multiple output formats (HTML, PDF, etc.). -- This book uses Quarto.]{.aside} On a higher level, tools that enable interoperability between programming languages have become available. For example, the `r BiocStyle::Biocpkg("reticulate")` package provides an R interface to Python, including support to translate between objects from both languages; `r BiocStyle::Biocpkg("basilisk")` facilitates Python environment management within the Bioconductor ecosystems, but can be interfaced with `reticulate` as well; and, Quarto can generate dynamic reports from code in different languages.

```{=html}
<!--
Nevertheless, future efforts should be made on the developers' side to consider existing frameworks in the first place (during implementation), and to provide users with documentation (e.g., code demos) and the necessary methods (e.g., conversion functions) to use other tools, if they so wish, in multi-tool and potentially multilingual analysis pipelines.

A standardized file format (FF) that can be written/exported and read/imported anywhere, while tracking all relevant (meta)data, is crucial to (F)AIR (i.e., accessible, interoperable, reusable) data and software interoperability. 
A promising development towards such a standard for imaging-based spatial omics data (and, in turn, `simpler' sequencing-based spatial and single-cell omics data) is the Open Microscopy Environment's next-generation FF (OME-NGFF)\cite{Moore2021-OME-NGFF} OME-Zarr\cite{Moore2023-OME-Zarr}.
Developed for cloud-storage of large high-dimensional arrays, Python's Zarr\cite{Miles2020-Zarr} provides a standardized FF that has already been adopted by many scientific communities.
The `r BiocStyle::Biocpkg("Rarr")` package, which has recently been submitted to Bioconductor, aims to provide an R interface to Zarr.
Together, these developments have the potential to lay a foundation for more seamless interoperability of computational tools to analyze single-cell and spatial omics data.
-->
```

## Dependencies

```{r}
# retrieve dataset from OSF repo
library(OSTA.data)
library(SpatialExperiment)
library(VisiumIO)
library(zellkonverter)
library(jsonlite)
library(reticulate)
library(basilisk)
```

## Configuring Python

We first create an environment with Python modules necessary the examples in this chapter. For that, we first define a `r BiocStyle::Biocpkg("basilisk")` environment and use the resulting conda environment with `r BiocStyle::Biocpkg("reticulate")`. 

```{r env, warning=FALSE}
env <- BasiliskEnvironment(
  envname="bkg-interop", 
  pkgname="base",
  channels=c("conda-forge", "bioconda", "pytorch"),
  packages=c(
    "python=3.11",
    "Dask=2024.12.1"),
  pip=c(
    "scvi-tools==1.3.0",
    # "torchmetrics==0.6.0",
    "squidpy==1.6.2"))
# activate underlying conda environment
use_condaenv(obtainEnvironmentPath(env))
```

## SingleCellExperiment

### Calling Python

After configuring the python environment, R commands will now be `reticulate`d; e.g.: [Note that running code in this way comes with the overhead of starting up a Python session in the background. But this is typically negligible compared to the runtime required to run computation-heavy methods, or when analyzing large-scale single-cell and spatial data (100,000s of cells).]{.aside}

```{r eval=FALSE}
#| code-fold: false
scvi <- reticulate::import("scvi")
dir.create(td <- tempfile())
(ad <- scvi$data$cortex(save_path=td))
# AnnData object with n_obs Ă— n_vars = 3005 Ă— 19972
#     obs: 'labels', 'precise_labels', 'cell_type'
```

### Continuing in R

We can access *any* of the variables above in R. For basic outputs, this works out of the box:

```{r eval=FALSE}
#| code-fold: false
unique(ad$obs$cell_type)
# [1] "interneurons"         "pyramidal SS" "pyramidal CA1"       
# [4] "oligodendrocytes"     "microglia"    "endothelial-mural"   
# [7] "astrocytes_ependymal"
```

Then again, `reticulate` supports few direct [type conversions](https://rstudio.github.io/reticulate/#type-conversions) (e.g. dictionary $\leftrightarrow$ named list). In the example demonstrated here, we can use `zellkonverter` to go from `AnnData` to `SingleCellExperiment`:

```{r eval=FALSE}
#| code-fold: false
(sce <- AnnData2SCE(ad))
# class: SingleCellExperiment 
# dim: 19972 3005 
# metadata(0):
# assays(1): X
# rownames(19972): Tspan12 Tshz1 ... Gm21943_loc3 Gm20738_loc3
# rowData names(0):
# colnames(3005): 0 1 ... 3003 3004
# colData names(3): labels precise_labels cell_type
# reducedDimNames(0):
# mainExpName: NULL
# altExpNames(0):
```

### Back to Python

We can also do the reverse, i.e. go from R's `SingleCellExperiment` to Python's `AnnData`:

```{r eval=FALSE}
#| code-fold: false
(ad <- SCE2AnnData(sce, X_name="X"))
# AnnData object with n_obs Ă— n_vars = 3005 Ă— 19972
#     obs: 'labels', 'precise_labels', 'cell_type'
#     uns: 'X_name'
```

## SpatialExperiment

### Starting with R

For the example with `SpatialExperiment`, we will use the dataset from @Janesick2023, which includes Visium measurements on human breast cancer tissue.

```{r load-spe, message=FALSE, results="hide"}
# retrieve dataset from OSF repo
id <- "Visium_HumanBreast_Janesick"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
obj <- TENxVisium(
    spacerangerOut=td, 
    format="h5", 
    images="lowres")
(spe <- VisiumIO::import(obj))
```

We also have to parse the original scaling information for spots and images available in standard Visium output. We will use this later during conversion.

```{r scalefactors, message=FALSE, results="hide"}
# retrieve scalefactors for Visium assay
scalefactors <- jsonlite::read_json(file.path(td, "outs/spatial/scalefactors_json.json"))
```

We can also use `SCE2AnnData` function from the previous example and convert the `SingleCellExperiment` relevant information of the `SpatialExperiment` object to an `AnnData`.

```{r}
(ad <- SCE2AnnData(spe, X_name="counts"))
# AnnData object with n_obs Ă— n_vars = 4269 Ă— 18085
#     obs: 'in_tissue', 'array_row', 'array_col', 'sample_id'
#     var: 'ID', 'Symbol', 'Type'
#     uns: 'X_name', 'resources', 'spatialList'
```

We should now populate `uns` and `obsm` components of `AnnData` with spatial coordinates and images. We start with the coordinates.

```{r}
coords <- spatialCoords(spe)
colnames(coords) <- c("x", "y")
obsm <- list(spatial = coords)
```

Now lets create the `uns` component. The list of `uns` should compose of as many samples as the images in the `SpatialExperiment` object. Also, each sample entry in the list should have two elements, one for the image and the other for scaling information.

```{r}
# get image metadata
imgdata <- imgData(spe)

# get image
img <- imgRaster(spe)
img <- apply(img, c(1,2), \(x) col2rgb(x))
img <- aperm(img, perm = c(2,3,1))
img <- img/255

# create uns
uns <- list(
  spatial = setNames(list(NULL), imgdata$sample_id)
)
uns[["spatial"]][[imgdata$sample_id]] <-
  list(images = list(lowres = img),
       scalefactors = scalefactors)
```

Now lets insert the components back to the `AnnData`.

```{r}
ad$library_id <- imgdata$sample_id
ad$obsm <- obsm
ad$uns <- uns
ad$write_h5ad("spe.h5ad")
```

### Calling Python

Now that we have converted our `SpatialExperiment` object to `AnnData` format, we can use `squidpy` module to visualize the Visium data. Since we wrote the `spe.h5ad` to disk, we can as well use python directly.

```{python warning=FALSE, fig.keep='all', fig.show='hold'}
#| code-fold: false
import squidpy as sq
import anndata as ad
import matplotlib.pyplot as plt
adata = ad.read_h5ad("spe.h5ad")
sq.pl.spatial_scatter(adata,
                      color=["array_row"], 
                      img_res_key = "lowres")
```

## Appendix

### References {.unnumbered}
