# Datasets {#sec-bkg-datasets}

Throughout this book, we will rely on a set of publicly available datasets 
that cover different sequencing- and imaging-based platforms, namely: 
Visium, Visium HD, Xenium (10x Genomics) and CosMx (NanoString). 

## Reading

These have been deposited in an Open Storage Framework (OSF) repository
[here](https://osf.io/5n4q3), and can be easily queried and downloaded 
using functions from the `r BiocStyle::CRANpkg("osfr")` package.

For convenience, we have implemented `OSTA.data` to:

- retrieve a dataset from our OSF node
- cache it as a *.zip* archive using `r BiocStyle::Biocpkg("BiocFileCache")`
- unpack and read it into a `SpatialExperiment`  
using `r BiocStyle::Biocpkg("VisiumIO")` for Visium(HD),  
or otherwise `r BiocStyle::Biocpkg("SpatialExperimentIO")`

```{r osta-data, eval=FALSE}
if (!require("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("estellad/OSTA.data")
```

The following datasets are currently available:

```{r list-data, message=FALSE, warning=FALSE}
OSTA.data::.list_data()
```

[Note that this might take a while, as data needs to be downloaded and zipped; 
it will go much quicker the second time (for a given dataset identifier).]{.aside}
Any of the above can be retrieved via:

```{r read-data, message=FALSE, warning=FALSE}
(spe <- OSTA.data::.read_data("Xenium_HumanColon_Oliveria"))
```

## Details

Below, we briefly summarize the characteristics of each dataset,
and note across which parts of the book each dataset is being used. <!--TODO-->

**HumanBreast1/2_Janesick**

- @Janesick2023
- **Visium**
  - source: [GSM7782699](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM7782699)
  - 4,992 spots x 18,085 genes
- **Xenium**
  - source: [GSM7780153](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM7780153), [GSM7780154](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM7780154)
- **Chromium**
  - source: [GSM7782698](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM7782698)
  - 37,143 genes x 708,252 cells

**HumanColon_Oliveira**

- @Oliveira2024
- source: [10x Genomics](https://www.10xgenomics.com/products/visium-hd-spatial-gene-expression/dataset-human-crc)
- **Visium**
  - 18,085 genes x 4,269 spots
- **VisiumHD**
  - 18,085 genes x
    - 545,913 bins (8um)
    - 137,051 bins (16um)
- **Chromium**
  - 18,082 genes x 279,609 cells
  - 8 samples across 5 patients (P1-5)
    - normal adjacent tissue (NAT):<br>32,801 cells (P2,3,5)
    - colorectal carcinoma (CRC):<br>41,933 cells (P1-5)

**CosMx1k_MouseBrain1/2**

- source: [nanostring.com](https://nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/)

**CosMx6k_MouseBrain**

- source: [nanostring.com](https://nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/human-frontal-cortex-ffpe-dataset/)

## References {.unnumbered}
