# Spatially variable genes {#sec-crs-spatially-variable}

## Preamble

### Introduction

Methods to identify spatially variable genes (SVGs) aim to identify genes whose expression is not uniform, and is instead characterized by strong spatial structures. These methods can be used as a spatially-aware alternative to feature selection methods described in @sec-seq-norm-feat-dimred, or to identify potentially biologically informative genes that can serve as candidates for further experimental follow-up.

Here, we define SVGs as any genes with **non-uniform spatial expression patterns** in a tissue.

SVGs are usually identified integrating gene expression measurements with spatial coordinates, either with or without predefined spatial domains. SVGs approaches can be broadly categorized into three types: overall SVGs, spatial-domain-marker SVGs, and cell-type-specific SVGs [@Yan2024]. 

The detection of overall SVGs is sometimes used as a feature selection step for further downstream analyses such as spatially-aware clustering (see @sec-seq-clustering). Several methods have been proposed for detecting overall SVGs; below, we report some few notable examples:

- [spatialDE](https://www.nature.com/articles/nmeth.4636) and [nnSVG](https://www.nature.com/articles/s41467-023-39748-z#Bib1), which are based on  Gaussian process model;
- [Moran's I](https://en.wikipedia.org/wiki/Moran%27s_I) and [Geary's C](https://en.wikipedia.org/wiki/Geary%27s_C), that rank genes according to their observed spatial autocorrelation (see @sec-crs-spat-stat);
- [SPARK-X](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-021-02404-0), that uses a non-parametric test of the covariance matrices of the spatial expression data.

Spatial-domain-specific SVGs target changes in gene expression between spatial domains, which are used to summarize the whole spatial information. Genes displaying expression changes across spatial clusters indicate SVGs. Spatial domains can be predefined based on morphology knowledge or identified through spatially resolved clustering algorithms such as `r BiocStyle::Biocpkg("BayesSpace")` [@Zhao2021-BayesSpace] and `r BiocStyle::Biocpkg("Banksy")` [@Singhal2024-BANKSY]. Methods that belong to this category include [SpaGCN](https://www.nature.com/articles/s41592-021-01255-8) and [DESpace](https://academic.oup.com/bioinformatics/article/40/2/btae027/7577855).

Cell-type-specific SVG methods leverage external cell-type annotations to identify SVGs within cell types. They analyze interaction effects between cell types and spatial coordinates [@Yan2024]. Example methods include [CTSV](https://academic.oup.com/bioinformatics/article/38/17/4135/6632658), [C-SIDE](https://www.nature.com/articles/s41592-022-01575-3), and [spVC](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-024-03245-3).

The following is a list of methods mentioned above:

- Python (pip): `r BiocStyle::Githubpkg("Teichlab/SpatialDE")` [@Svensson2018-SpatialDE]
- R (Bioc): `r BiocStyle::Biocpkg("nnSVG")` [@Weber2023-nnSVG]
- R (GitHub): `r BiocStyle::Githubpkg("xzhoulab/SPARK")` [@Sun2020-SPARK] and `SPARK-X` [@Zhu2021-SPARK-X]
- Python (pip): `r BiocStyle::Githubpkg("jianhuupenn/SpaGCN")` [@Hu2021-SpaGCN]
- R (Bioc): `r BiocStyle::Biocpkg("DESpace")` [@Cai2024-DESpace]
- R (GitHub): `r BiocStyle::Githubpkg("shanyu-stat/spVC")` [@Yu2024-spVC]
- R (Bioc): `r BiocStyle::Biocpkg("CTSV")` [@Yu2022-CTSV]
- R (Bioc): `r BiocStyle::Githubpkg("dmcable/spacexr")` [@Cable2022-spacexr]

Below, we provide scripts for implementing some analysis frameworks for detecting SVGs.

### Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(SpatialExperiment)
library(nnSVG)
library(DESpace)
library(scran)
library(scater)
library(ggspavis)
library(patchwork)
library(dplyr)
library(tidyr)
library(ggrepel)
(spe <- readRDS("seq-spe_clustering.rds"))
```

## Analysis

### Spatially aware analysis frameworks

Here, we demonstrate brief examples of how to identify a set of top SVGs using (`r BiocStyle::Biocpkg("nnSVG")` [@Weber2023-nnSVG] and `r BiocStyle::Biocpkg("DESpace")` [@Cai2024-DESpace]). These methods are available through Bioconductor and can be easily integrated into Bioconductor-based workflows. 

#### nnSVG

In this example, we use a small subset of the dataset for faster runtime. We select a subset of the data, by subsampling the set of spots and including stringent filtering for lowly expressed genes. A full analysis with `nnSVG` using all spots for this dataset and default filtering parameters for an individual Visium sample from human brain tissue (available via [spatialLIBD](http://bioconductor.org/packages/spatialLIBD) Bioconductor R package) takes around 45 minutes on a laptop.

```{r sub, message=FALSE}
# # set random seed for number generation
# # in order to make results reproducible
# set.seed(123)
# # sample 100 spots to decrease runtime in this demo
# # (note: skip this step in full analysis)
# n <- 100
# sub <- spe[, sample(ncol(spe), 100)]
# # filter lowly expressed genes using stringent
# # criteria to decrease runtime in this demo
# # (note: use default criteria in full analysis)
# sub <- filter_genes(sub,     # filter for genes with...
#     filter_genes_ncounts=10, # at least 10 counts in
#     filter_genes_pcspots=3)  # at least 3% of spots
# # re-normalize counts post-filtering
# sub <- logNormCounts(sub)
```

```{r nnSVG, cache=TRUE}
# # run nnSVG
# set.seed(123)
# sub <- nnSVG(sub)
```

```{r res}
# # extract gene-level results
# res_nnSVG <- rowData(sub)
# # show results
# head(res_nnSVG, 3)
# # count significant SVGs (at 5% FDR significance level)
# table(res_nnSVG$padj <= 0.05)
# # identify the top-ranked SVGs
# res_nnSVG$gene_name[res_nnSVG$rank == 1]
# # store the top 7 SVGs
# top_nnSVG <- res_nnSVG$gene_name[res_nnSVG$rank < 7.5]
```


#### DESpace

`DESapce` relies on pre-computed spatial domains to summarize the primary spatial structures of the data. See @sec-seq-clustering and @sec-img-clustering for clustering methods.

```{r plt-BayesSpace, message=FALSE, fig.width=4, fig.height=5.5}
#| code-fold: true
# Visualization of spatial domains detected by BayesSpace
plotSpots(spe, annotate = "BayesSpace", legend_position = "none") + ggtitle("Saptial domains detected by BayesSpace")
```

```{r DESpace}
# run DESpace
set.seed(123)
res <- DESpace_test(spe, spatial_cluster = "BayesSpace")
res_DESpace <- res$gene_results
# show results
head(res_DESpace, 3)
# count significant SVGs (at 5% FDR significance level)
table(res_DESpace$FDR <= 0.05)
# store the top 7 SVGs
top_DESpace <- res_DESpace$gene_id[seq_len(7)]
```

### Non-spatial analysis frameworks

We can also use methods originally developed for bulk and single-cell transcriptomics, such as those for detecting highly variable genes (HVGs) and differentially expressed genes (DEGs), on spatial omics data. Note that HVGs are defined only based on molecular features (i.e., gene expression), and do not incorporate spatial information. In the context of single-cell and spatial transcriptomics, differentially expressed (DE) may refer to differences between groups of samples, or to differences between clusters (e.g., cell types, or spatial domains); below, we consider changes across spatial structures.


#### Highly variable genes (HVGs)


For a comprehensive discussion on HVG selection, we refer readers to [OSCA](https://bioconductor.org/books/3.19/OSCA.basic/feature-selection.html#hvg-selection). Below, we outline a standard approach for selecting HVGs. For more details and examples, see @sec-seq-norm-feat-dimred.

```{r hvg}
# fit mean-variance relationship
dec <- modelGeneVar(spe)
# select the top 10% of HVGs
prop_hvgs <- getTopHVGs(dec, prop = 0.1)
# select genes based on the FDR of the test to identify HVGs 
fdr_hvgs <- getTopHVGs(dec, fdr.threshold = 0.05)
length(prop_hvgs); length(fdr_hvgs)
top_hvgs <- prop_hvgs[seq_len(7)]
```


#### Differentially expressed genes (DEGs)

For details on identifying DE genes between groups of cells from single-cell RNA-seq data, we refer readers to [OSCA](https://bioconductor.org/books/3.19/OSCA.basic/marker-detection.html). Below, we show a potential approach, using marker detection methods on spatial transcriptomics data, to identify DE genes between spatial clusters.

```{r deg, fig.width=10, fig.height=3}
# differential gene expression analyses
mgs <- findMarkers(spe, groups=spe$BayesSpace, direction="up")
# select for a few markers per cluster
deg <- lapply(mgs, \(df) rownames(df)[df$Top <= 3])
length(deg <- unique(unlist(deg)))
# select one top marker for each cluster
top_deg <- lapply(mgs, \(df) rownames(head(df, 1))) 
top_deg <- unique(unlist(top_deg))
```

::: {.callout-note collapse="true" title="Downstream analyses"}

## Downstream analyses 

The set of top SVGs may be further investigated, e.g. by plotting the spatial expression of several top genes and via gene pathway analyses (i.e., comparing significant SVGs with known gene sets associated with specific biological functions). 

## Visualization

To visualize the expression levels of selected genes in spatial coordinates on the tissue slide, we can use plotting functions from the `r BiocStyle::Biocpkg("ggspavis")` package.

```{r plt-xy, message=FALSE, fig.width=10, fig.height=7}
#| code-fold: true
# selected genes for 4 methods
# method_names <- c("nnSVG", "DESpace", "HVGs", "DEGs")
# gs <- c(top_DESpace, top_hvgs, top_deg)
# # convert ensembl gene id to gene name
# gs <- rowData(spe)$gene_name[match(gs, rowData(spe)$gene_id)]
# gs <- c(top_nnSVG, gs)
# # expression plots for each top gene
# ps <- lapply(gs, \(.) 
#     plotSpots(spe, 
#         annotate=., 
#         assay_name="logcounts", 
#         feature_names="gene_name")
# )
# # seven top genes for each method
# ranges <- split(seq_along(ps), ceiling(seq_along(ps) / 7))
# # wrap plots
# plots <- mapply(\(idx, name) {
#     wrap_plots(ps[idx], nrow = 1) & 
#         plot_annotation(title = name) & 
#         theme(
#             legend.key.width = unit(0.4, "lines"), 
#             legend.key.height = unit(0.8, "lines")
#         ) &
#     scale_color_gradientn(colors = pals::parula())
# }, ranges, method_names, SIMPLIFY = FALSE)
# # to keep titles
# final_plot <- wrap_plots(
#     lapply(seq_along(plots), \(i) 
#         wrap_elements(plots[[i]])), 
#     ncol = 1
# )
# final_plot
```


## Downstream comparison

We can compare the ranks of the genes detected by each method and compute pairwise correlations, highlighting two known cortical layer-associated SVGs: *MOBP* and *SNAP25*.
While all four methods—nnSVG, DESpace, HVGs, and DEGs—identify similar gene ranks, each method captures distinct aspects of gene expression, with varying pairwise correlations between them.


```{r plt-data, message=FALSE}
#| code-fold: true
# subset data for the 113 genes nnSVG is based on
# sub_gene <- rowData(sub)$gene_id
# sub_DESpace <- res_DESpace[sub_gene, ] 
# sub_HVG <- as.data.frame(dec[sub_gene, ])
# 
# # aggregate DEGs for each spatial domain
# subset_mgs <- lapply(mgs, \(x) x[sub_gene, 1:3] |> as.data.frame())
# sub_DEG <- do.call(cbind, subset_mgs) 
# top_cols <- sub_DEG |> select(ends_with(".Top"))
# sub_DEG$Top <- do.call(pmin, c(top_cols, na.rm = TRUE))
# 
# # compute ranks 
# sub_DESpace$rank <- rank(sub_DESpace$FDR, ties.method = "first")
# sub_HVG$rank <- rank(-1 * sub_HVG$bio, ties.method = "first")
# sub_DEG$rank <- rank(sub_DEG$Top, ties.method = "first")
# 
# # combine 'rank' column for each method
# res_all <- list(
#     nnSVG = as.data.frame(res_nnSVG),
#     DESpace = sub_DESpace, HVGs = sub_HVG, DEGs = sub_DEG
# )
# rank_all <- do.call(cbind, lapply(res_all, `[[`, "rank"))
# rank_all <- data.frame(gene_id = sub_gene, rank_all)
```

```{r plt-scatter-nnsvg, message=FALSE, fig.width=10, fig.height=7}
#| code-fold: true
# known SVGs for this data
# known_genes <- c("MOBP", "SNAP25")
# # convert data structure
# rank_long <- rank_all |>
#     left_join(data.frame(rowData(sub)[, c("gene_id", "gene_name")]), by = "gene_id") |>
#     select(all_of(c("gene_name", method_names))) |>
#     pivot_longer(cols = c(nnSVG, DESpace, HVGs, DEGs), 
#                  names_to = "method", 
#                  values_to = "rank")
# # all pairwise comparisons
# df_pairs <- t(combn(method_names, 2)) |> as.data.frame()
# # function to plot each pairwise comparison
# plot_pairwise_comparison <- function(m1, m2, df) {
#   # filter the data for the two methods being compared
#   df <- df |>
#       filter(method %in% c(m1, m2)) |>
#       pivot_wider(names_from = method, values_from = rank)
#   # compute pearson correlation between methods
#   cor_val <- cor(df[[m1]], df[[m2]], method = "spearman")
#   # plot
#   ggplot(df, aes(x = .data[[m1]], y = .data[[m2]])) +
#       geom_point() + 
#       geom_text_repel(data = df %>% filter(gene_name %in% known_genes), 
#                       aes(label = gene_name), color = "red", size = 3.25, 
#                       nudge_x = 50, nudge_y = 10, box.padding = 0.5) +
#       labs(x = paste(m1, "rank"), y = paste(m2, "rank"),
#            title = paste(m2, "vs.", m1, ": Cor = ", round(cor_val, 2))) +
#       #scale_color_manual(values = c("darkorange", "firebrick3", "deepskyblue2")) +
#       theme_bw() + coord_fixed() + 
#       xlim(c(0, 120)) + ylim(c(0, 120))
# }
# 
# # generate and display all pairwise comparison plots
# plots <- lapply(seq_len(nrow(df_pairs)), \(i) {
#     plot_pairwise_comparison(df_pairs[i, 2], df_pairs[i, 1],rank_long)
# })
# 
# wrap_plots(plots, ncol = 3)
```

:::

## Appendix

### Benchmarks and reviews {.unnumbered}

* https://pmc.ncbi.nlm.nih.gov/articles/pmid/38370977/
* https://arxiv.org/abs/2405.18779
* https://genomebiology.biomedcentral.com/articles/10.1186/s13059-023-03145-y

### References {.unnumbered}
