# Imputation

## Introduction

The analyses demonstrated here will use 313-plex Xenium data (10x Genomics) of a human breast cancer biopsy section [@Janesick2023] and the 1k-plex CosMx data (Nanostring) of a mouse brain sample. 

## Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(scran)
library(scater)
library(harmony)
library(OSTA.data)
library(SpatialExperiment)
library(SpatialExperimentIO)
library(DropletUtils)
library(Matrix)
library(data.table)
library(RANN)
library(igraph)
library(patchwork)
```

## Data import

`r BiocStyle::Biocpkg("SpatialExperimentIO")` provides functions to import CosMx, Xenium, and MERSCOPE datasets in Bioconductor, starting from the output provided by the three platforms.

### Xenium

We now load Xenium data:

```{r, message=FALSE, warning=FALSE}
id <- "Xenium_HumanBreast1_Janesick"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
spe <- readXeniumSXE(td)
spe$sample_id <- "JanesickBC_Xenium1"
spe
```

### Single Cell

We now load the single cell data:

```{r, message=FALSE, warning=FALSE}
id <- "Chromium_HumanBreast_Janesick"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
h5 <- file.path(td, "filtered_feature_bc_matrix.h5")
sce <- read10xCounts(h5, col.names=TRUE)
metadata(sce) <- list()
sce$sample_id <- "Janesick_Chromium"
sce

# set gene symbols as feature names
rownames(sce) <- make.unique(rowData(sce)$Symbol)

# count the number of reads per cell
sce$Count <- colSums(assay(sce))

# retrieve cell type labels
fs <- list.files(td, full.names=TRUE)
csv <- grep("csv$", fs, value=TRUE)
cd <- read.csv(csv, row.names=1)
colData(sce)[names(cd)] <- cd[colnames(sce), ]
```
## scRNA Marker Analysis

```{r markers}
# # normalize
# sce <- logNormCounts(sce)
# 
# # test for marker genes
# colLabels(sce) <- sce$Annotation
# markers <- findMarkers(sce, test="wilcox", direction="up")
markers <- readRDS("../../../PrincipalSTA-supp/markers_imputation.rds")
top.markers <- mapply(function(tbl,nm){
  cur.tbl <- tbl[1:20,c("p.value", "FDR", "summary.AUC")]
  data.frame(cur.tbl, 
             celltype = nm, 
             genes = rownames(cur.tbl),
             row.names = NULL)
}, tbl = markers, nm = names(markers), SIMPLIFY = FALSE)
top.markers <- do.call("rbind", top.markers)
```

## Harmony Integration

Besides physically aligning both datasets, we can integrate them on 
a transcriptional level; here, using `r BiocStyle::Biocpkg("harmony")`.

To this end, we first consolidate the Visium and Xenium data into one object:

```{r obj}
# convert to single cell
spe_single <- as(spe, "SingleCellExperiment")

# get common genes
gs <- intersect(rownames(spe_single), rownames(sce))
cd <- intersect(names(colData(spe_single)), names(colData(sce)))
lys <- list(spe_single=spe_single, sce=sce)
lys <- lapply(lys, \(scRNA) {
    scRNA <- scRNA[gs,]
    rowData(scRNA) <- rowData(scRNA)[gs,]
    colData(scRNA) <- colData(scRNA)[cd]
    assay(scRNA) <- as(assay(scRNA), "dgCMatrix")
    return(scRNA)
})

# combine datasets
obj <- SingleCellExperiment(assays = list(counts = cbind(assay(lys[[1]]), assay(lys[[2]]))),
                            colData = rbind(colData(lys[[1]]), colData(lys[[2]])),
                            rowData = rowData(lys[[1]]))
obj$col_counts <- colSums(assay(obj))
```

We now initiate the harmony integration:

```{r int}
# minimal filtering
obj <- obj[, obj$col_counts > 20]
# library size normalization 
obj <- logNormCounts(obj)
# principal component analysis
# (w/o additional feature selection)
obj <- runPCA(obj)
# 'harmony' integration
pcs <- RunHarmony(
    data_mat=reducedDim(obj, "PCA"), 
    meta_data=obj$sample_id, 
    verbose=FALSE)
reducedDim(obj, "PCA") <- pcs
# dimensionality reduction
map <- calculateUMAP(t(pcs))
reducedDim(obj, "UMAP") <- map
```

let us now visualize the embedding of both Xenium and chromium cells to assess the mixing

```{r plt-map, fig.width=3.5, fig.height=3}
plotUMAP(obj, colour_by="sample_id", point_size=0.1) +
    guides(col=guide_legend(override.aes=list(alpha=1, size=2))) +
    theme_void() + theme(aspect.ratio=1, legend.key.size=unit(0, "pt"))
```
We can also check if annotations of the chromium cells are associated with the embedding of cells when the embedding was trained jointly on Xenium and chromium cells.

```{r plt-map-subset, fig.width=3.5, fig.height=3}
obj_subset <- obj[,obj$sample_id == "Janesick_Chromium"]
annotations <- setNames(sce$Annotation, colnames(sce))
obj_subset$Annotation <- annotations[colnames(obj_subset)]
plotUMAP(obj_subset, colour_by="Annotation", point_size=0.1) +
    guides(col=guide_legend(override.aes=list(alpha=1, size=2))) +
    theme_void() + theme(aspect.ratio=1, legend.key.size=unit(0, "pt"))
```

## Imputation

Now that we show there is a correction of batch effects in the PC space. We can transfer counts of single cell data to the Xenium dataset by detecting kNNs of xenium cells in PC space and aggregate missing genes over nearest neighbors.

```{r impute}
embeddings <- reducedDim(obj, "PCA")

# find kNN across reference and query
nNeighbors <- 20
reference_embedding <- embeddings[obj$sample_id == "JanesickBC_Xenium1",]
query_embedding <- embeddings[obj$sample_id == "Janesick_Chromium",]
knn <- RANN::nn2(query_embedding,
                 reference_embedding,
                 k = nNeighbors)
# knn <- readRDS("../../../PrincipalSTA-supp/knn.rds")

# create weight matrix
# nn_graph <- knn$nn.idx
# nnedges <- data.table::melt(data.table::data.table(id = 1:nrow(nn_graph), nn_graph), id.vars = "id")
# nnedges <- as.vector(t(as.matrix(nnedges[,c("id", "value")])))
# weights <- igraph::make_empty_graph(n = nrow(nn_graph)) %>% 
#   add_edges(nn_graph) %>% 
#   as_adjacency_matrix()
# weights <- weights[,1:nrow(query_embedding)]
# weights <- weights * 1/nNeighbors

# create weight matrix alternative
# weights <- Matrix::Matrix(0, 
#                           nrow = nrow(reference_embedding), 
#                           ncol = nrow(knn$nn.idx), 
#                           sparse = TRUE)
# for (n in seq_len(nrow(knn$nn.idx))) {
#   print(n)
#   weights[knn$nn.idx[n, ], n] <- 1/nNeighbors
# }
# weights <- t(weights)
# weights <- weights[,1:nrow(query_embedding)]
weights <- readRDS("../../../PrincipalSTA-supp/weights.rds")

# impute by kNN
query_counts <- assay(sce)[unique(top.markers$genes),colnames(obj)[obj$sample_id == "Janesick_Chromium"]]
imputed_reference_counts <- query_counts %*% t(weights)
colnames(imputed_reference_counts) <- rownames(reference_embedding)
```

## Visualization

Create the imputed reference counts

```{r create_new_counts}
spe_imputed <- spe[,colnames(obj)[obj$sample_id == "JanesickBC_Xenium1"]]
altExp(spe_imputed, "Imputed") <- SummarizedExperiment(assays = list(counts = imputed_reference_counts))
spe_imputed.tmp <- swapAltExp(spe_imputed, name="Imputed")
spe_imputed_new <- SpatialExperiment(list(counts = assay(spe_imputed.tmp, "counts")), 
                                 colData = colData(spe_imputed),
                                 spatialCoords = spatialCoords(spe_imputed), 
                                 imgData = imgData(spe_imputed))
```

We can now compare transfered counts with the original counts. Lets start with ACTA2 which is included in both datasets.

```{r vis-imputed}
vis.data <- data.frame(spatialCoords(spe_imputed),
                       ACTA2_new = assay(spe_imputed_new, "counts")["ACTA2",],
                       ACTA2 = assay(spe_imputed, "counts")["ACTA2",])
g1 <- ggplot(vis.data, aes(x = x_centroid, y = y_centroid, color = ACTA2)) + 
  geom_point(size = 1) +
  scale_color_gradient2(low="grey", high="red", space ="Lab")
g2 <- ggplot(vis.data, aes(x = x_centroid, y = y_centroid, color = ACTA2_new)) + 
  geom_point(size = 1) +
  scale_color_gradient2(low="grey", high="red", space ="Lab")
g1 | g2
```

Now lets select a gene that is not available in the Xenium panel. Here, KRT17 is again a marker of ACTA2 myoepithelial cells that together with ACTA2

```{r vis-imputed}
vis.data <- data.frame(spatialCoords(spe_imputed),
                       ACTA2 = assay(spe_imputed, "counts")["ACTA2",],
                       KRT17 = assay(spe_imputed_new, "counts")["KRT17",])
g1 <- ggplot(vis.data, aes(x = x_centroid, y = y_centroid, color = ACTA2)) + 
  geom_point(size = 1) +
  scale_color_gradient2(low="grey", high="red", space ="Lab")
g2 <- ggplot(vis.data, aes(x = x_centroid, y = y_centroid, color = KRT17)) + 
  geom_point(size = 1) +
  scale_color_gradient2(low="grey", high="red", space ="Lab")
g1 | g2
```

## References {.unnumbered}
