# Clustering {#sec-crs-clustering}

## Preamble

### Introduction

In spatial transcriptomics data, we can apply clustering algorithms to identify spatial domains, which represent spatially defined regions consisting of relatively consistent gene expression profiles. 
For example, spatial domains may consist of regions containing cells from a single cell type or a consistent mixture of cell types.

Several alternative approaches exist for identifying spatial domains. One strategy involves applying standard clustering algorithms from single-cell workflows without incorporating spatial information. 
Alternatively, spatial information can be integrated at various stages of the workflow. 
For example, spatial data can inform preprocessing steps, such as selecting spatially variable genes (SVGs), followed by the application of either non-spatial or spatial clustering algorithms. 
Another widely used strategy involves generating a latent space that incorporates spatial coordinates, often leveraging Bayesian framework or neural networks.

As advancements in spatial transcriptomics techniques enhance tissue resolution, computational tradeoffs between various approaches and parameters become increasingly relevant.

It is also important to keep in mind that when we use clustering to define cell types and/or states, these can be defined at various resolutions (or even on a continuum). 
The optimal number of clusters depends on the biological context -- in particular, there is no "true" number of clusters, since this depends on the biological context (e.g. if we are comparing major cell populations vs. comparing rare subtypes), so the choice of the optimal number of clusters or resolutions requires some judgment and biological interpretation.

Once we have identified spatial domains, these can then be further investigated in additional downstream analyses.

### Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(scran)
library(scater)
library(igraph)
library(Banksy)
library(ggspavis)
library(pheatmap)
library(basilisk)
library(patchwork)
library(OSTA.data)
library(BayesSpace)
library(reticulate)
library(SpatialExperiment)
# set seed for random number generation
# in order to make results reproducible
set.seed(2025)
# load processed Xenium datasets
spe <- readRDS("img-spe_cl.rds")
```

## Clustering

### Non-spatial

As a baseline, we will perform a standard graph-based clustering approach developed for scRNA-seq data, using molecular features (gene expression) only.
Specifically, we construct a shared nearest neighbor (SNN) graph based on non-spatial PCs (`reducedDim` slot `"PCA_tx"`), and use the Leiden algorithm for community detection.

```{r clu-Leiden}
# build cellular shared nearest-neighbor (SNN) graph
g <- buildSNNGraph(spe, use.dimred="PCA_tx", type="jaccard", k=20)
# cluster using Leiden community detection algorithm
k <- cluster_leiden(g, objective_function="modularity", resolution=0.8)
table(spe$Leiden <- factor(k$membership))
```

### Spatially aware

Several methods to identify spatial domains in ST data have recently been developed, which each have various methodological and computational tradeoffs. 
A few example include `r BiocStyle::Biocpkg("BayesSpace")` [@Zhao2021-BayesSpace], `r BiocStyle::Biocpkg("Banksy")` [@Singhal2024-BANKSY] and `r BiocStyle::CRANpkg("PRECAST")` [@Liu2023-PRECAST] in R, as well as `r BiocStyle::Githubpkg("CellCharter")` [@Varrone2024-CellCharter] and `r BiocStyle::Githubpkg("SpaceFlow")` [@Ren2022-SpaceFlow] in Python.

Here, we demonstrate tools representative of different methodologies, namely, (probabilistic) `BayesSpace` for seq-based Visium, as well as (neighborhood-based) `Banksy` and (encoder-based) `CellCharter` for img-based Xenium data.

#### Probabilistic

Underlying `r BiocStyle::Biocpkg("BayesSpace")` [@Zhao2021-BayesSpace] is a Bayesian model that encourages neighboring spots to belong to the same cluster, and utilizes Markov Chain Monte Carlo (MCMC) for estimating model parameters.
[Default parameters are `nrep=50,000` MCMC iterations and a burn-in period of `burn.in=1,000`; the authors recommend running with at least 10,000 iterations. Here, we use fewer iterations for the sake of runtime, which will arguably compromise performance. Note that, since MCMC is stochastic, a seed for random number generation should be set in order to make results reproducible.]{.aside}

```{r BayesSpace, message=FALSE}
# load processed Visium dataset
vis <- readRDS("seq-spe_cl.rds")
# prepare data for 'BayesSpace'
# skipping PCA (already computed)
.vis <- spatialPreprocess(vis, skip.PCA=TRUE)
# perform spatial clustering with 'BayesSpace'
# using 'd=20' PCs and targeting 'q=10' clusters
.vis <- spatialCluster(.vis, q=10, d=20, nrep=1e3, burn.in=100) 
table(vis$BayesSpace <- factor(.vis$spatial.cluster))
```

#### Neighborhood-based

`r BiocStyle::Biocpkg("BANKSY")` [@Singhal2024-BANKSY] embeds cells in a product space of their own and their local neighborhood's (average) transcriptome, representing cell state and microenvironment. 
Here, the mixing parameter $\lambda\in[0,1]$ (argument `lambda` in `runBanksyPCA()`), controls the spatial component's weight; e.g., **when $\lambda=0$, `BANKSY` reduces to non-spatial clustering**.

```{r clu-Banksy, message=FALSE}
# perform 'Banksy' clustering
spe <- clusterBanksy(spe, dimred="PCA_sp", resolution=0.4)
# reformat cluster assignments
k <- grep("^clust", names(colData(spe)))
names(colData(spe))[k] <- "Banksy"
table(spe$Banksy <- factor(spe$Banksy))
```

#### Encoder-based

Another category of spatially-aware clustering method uses encoder architectures to generate a latent embedding incorporating both spatial and transcriptomics information, which is then clustered using standard algorithms. 
One typical example is `r BiocStyle::Githubpkg("CSOgroup/cellcharter")` [@Varrone2024-CellCharter], which uses a variational auto-encoder to generate a latental feature space and aggregate the features of each spot across its neighbors to preserve their spatial context. 

::: {.callout-note title="Other methods" collapse="true"}
Several other published methods have also adopted encoder architectures in different ways to perform spatially-aware clustering. 
`r BiocStyle::Githubpkg("zhanglabtools/STAGATE")` [@Dong2022-STAGATE] uses a graph attention autoencoder to integrate spatial and transcriptional data by modeling spatial neighborhoods as graphs, whereas `r BiocStyle::Githubpkg("Tang-Lab-super/PROST")` [@Liang2024-PROST] employs a self-attention transformer-based encoder to learn spatial patterns by jointly modeling gene expression and spatial coordinates. 
Graph-based encoder methods, such as `STAGATE`, may be particularly well-suited for imaging-based ST data, as their ability to model spatial neighborhoods as graphs aligns naturally with the continuous and fine-grained tissue spatial organization captured by different imaging platforms.
:::

```{r env, warning=FALSE}
# initialize 'basilisk' environment
env <- BasiliskEnvironment(
  envname="CellCharter", pkgname="OSTA",
  channels=c("conda-forge", "pytorch"),
  packages=c(
    "python=3.10.15",
    "pytorch=1.12.1",
    "torchvision=0.13.1",
    "torchaudio=0.12.1"),
  pip=c(
    "scvi-tools==0.20.3",
    "cellcharter==0.2.0",
    "anndata==0.10.9",
    "scanpy==1.10.4"))
# activate underlying conda environment
use_condaenv(obtainEnvironmentPath(env))
counts <- r_to_py(t(counts(spe)))
coords <- r_to_py(spatialCoords(spe))
```

```{python CellCharter}
import cellcharter as cc
import anndata as ad
import squidpy as sq
import pandas as pd
import numpy as np
import random
import scvi

adata = ad.AnnData(X = r.counts,
                   obsm = {"spatial":r.coords},
                   layers = {"counts": r.counts})

seed = 2025
random.seed(seed)
scvi.settings.seed = seed

# variational autoencoder for feature extraction
scvi.model.SCVI.setup_anndata(adata, layer="counts")
model = scvi.model.SCVI(adata, n_hidden=64)
model.train(early_stopping=True,
    # the parameters below aim to reduce runtime; 
    # in reality, it'd be better to use defaults
    max_epochs=70, batch_size=512, train_size=0.5, validation_size=0.2)

adata.obsm["X_scVI"] = model.get_latent_representation(adata).astype(np.float32)

# Getting neighborhood aggregation
sq.gr.spatial_neighbors(adata, coord_type="generic", delaunay=True, spatial_key="spatial", percentile=99)
cc.gr.aggregate_neighbors(adata, n_layers=3, use_rep="X_scVI", out_key="X_cellcharter")

# clustering by scanning a range of data number
mod = cc.tl.Cluster(n_clusters=14, random_state=seed)
mod.fit(adata, use_rep="X_cellcharter")

label_df = pd.DataFrame({"label": mod.predict(adata, use_rep="X_cellcharter")}) 
label_df[["label"]].value_counts()
```

```{r py2r}
# pull 'CellCharter' assignments from Python into R
spe$CellCharter <- unlist(py$label_df)
```

<!--
#### Other methods: Probability statistics-based methods 

* See more example in @sec-seq-clustering-de
* BASS
* DR.SC
-->

## Downstream

### Visualization

<!--Jieran:
* Visualization (Align the cluster labels before running it)
* Smoothness entropy checking
-->

Let's visualize the assignment obtains from non-spatial and spatially aware clustering:

```{r plot-xy-k-joint, message=FALSE, fig.width=10, fig.height=5}
#| code-fold: true
# TODO: 'ggspavis' is being a bit pedantic...
ks <- c("Label", "Leiden", "Banksy", "CellCharter")
spe$in_tissue <- 1; spe$x_centroid <- spe$y_centroid <- NULL
lapply(ks, \(.) {
    plt <- plotSpots(spe, annotate=.)
    plt$layers[[1]]$aes_params$stroke <- 0
    plt$layers[[1]]$aes_params$size <- 0.2
    plt
}) |>
    wrap_plots(nrow=2) &
    scale_color_manual(values=unname(pals::trubetskoy())) &
    theme(legend.key.size=unit(0, "lines"), legend.justification="left")
```

Especially in large tissues, the above type of plot makes it difficult to spot rare subpopulations, and might cause cells to overlap in regions with high cellular density.
This can be misleading, as we will tend to see only highly abundant subpopulations, or the cells plotted last and on top (i.e., later columns in the object).

To better distinguish between different subpopulations, we can instead generate separate spatial plots with one subpopulation highlighted at a time:

```{r plot-xy-k-split, fig.width=9, fig.height=6}
#| code-fold: true
lapply(levels(spe$Banksy), \(k) {
    # highlight cells assigned to cluster 'k'
    spe$foo <- spe$Banksy == k
    spe <- spe[, order(spe$foo)]
    plt <- plotSpots(spe, annotate="foo")
    plt$layers[[1]]$aes_params$stroke <- 0
    plt$layers[[1]]$aes_params$size <- 0.2
    plt + ggtitle(k)
}) |>
    wrap_plots(nrow=3) &
    scale_color_manual(values=c("lavender", "purple")) &
    theme(plot.title=element_text(hjust=0.5), legend.position="none")
```

<!--
### (Collapse) Effect of configurations on the results

* Cross-methods ARI validation

### (Collapse) Scalability of clusteirng methods on the results

* Using local machine to show the runtime availablility 
* Refer to benchmarking studies for potential scability issue
-->

::: {.callout-note title="PC regression" collapse="true"}

### PC regression

For any single-cell analysis where downstream tasks rely on PCs, it is useful to perform linear regression of (continuous or categorical) covariates of interest onto PCs. 
This quantifies the variance explained by the covariate and can help assess the extend of unwanted variation (due to, e.g., cell area) as opposed to subpopulations driving transcriptional differences. 
Here, we regress total counts, cell area, and cluster assignments from different methods against PCs:

```{r pcr}
pcs <- reducedDim(spe, "PCA_tx")
ids <- c("total_counts", "cell_area", ks)
pcr <- lapply(ids, \(id) {
    fit <- summary(lm(pcs ~ spe[[id]]))
    r2 <- sapply(fit, \(.) .$adj.r.squared)
    data.frame(id, pc=seq_along(r2), r2)
}) |> do.call(what=rbind)
```

Here, `Leiden` (transcription-only) and `Banksy` (spatially aware) clusterings perform similar in terms of capturing (spatially unaware) PCs; as to be expected, `Leiden` clusters do slightly better.
Had we set a higher value for $\lambda$ when running `Banksy`, results would diverge more; vice versa, using spatial PCs for regression would have `Leiden` clusters perform worse.
Also, note that we would expect results to converge for $\lambda=0$.

```{r plt-pcr, fig.width=5, fig.height=2.5}
#| code-fold: true
pcr$id <- factor(pcr$id, ids)
pal <- c("red", "magenta", "gold", "cyan", "blue", "black")
ggplot(pcr, aes(pc, r2, col=id)) +
    geom_line(show.legend=FALSE) + geom_point() +
    scale_color_manual("predictor", values=pal) +
    scale_x_continuous(breaks=c(1, seq(5, 20, 5))) +
    scale_y_continuous(limits=c(0, 1), breaks=seq(0, 1, 0.2)) +
    labs(x="principal component", y="coeff. of determination") +
    guides(col=guide_legend(override.aes=list(size=2))) +
    coord_cartesian(xlim=c(1, 20)) +
    theme_minimal() + theme(
        panel.grid.minor=element_blank(),
        legend.key.size=unit(0, "lines"))
```

:::

### DGE analysis

To help characterize (unsupervised) clusters, we want to identify 'markers' for each subpopulation, i.e., features whose expression is positively or negatively restricted to (a) specific subpopulation(s).
For details on identifying genes that are differentially expressed (DE) between groups of cells, we refer readers to [OSCA](https://bioconductor.org/books/3.19/OSCA.basic/marker-detection.html); a standard approach is given below, visualizing the average expression of exemplary markers across clusters:

```{r mgs}
# differential gene expression analysis
mgs <- findMarkers(spe, groups=spe$Leiden, direction="up")
# select for a few markers per cluster
top <- lapply(mgs, \(df) rownames(df)[df$Top <= 3])
length(top <- unique(unlist(top)))
```

```{r plt-mgs-hm, fig.width=10, fig.height=3}
# average expression by clusters
pbs <- aggregateAcrossCells(spe, 
    ids=spe$Leiden, subset.row=top, 
    use.assay.type="logcounts", statistics="mean")
# visualize averages z-scaled across clusters
pheatmap(mat=t(assay(pbs)), scale="column")
```

<!--

```{r DEG analysis, eval=FALSE, fig.width=8, fig.height=4}
#| code-fold: true
#| label: fig-charts
#| fig-cap: "Charts"
#| fig-subcap: 
#|   - "First"
#|   - "Second"
#| layout-ncol: 2

getTopDEGs <- function(spe, clustering) {
  mgs <- findMarkers(spe, groups=spe[[clustering]], direction="up")
  top <- lapply(mgs, \(.) rownames(.)[.$Top <= 1])
  top <- unique(unlist(top))
  top <- top[top %in% rownames(spe)]
  
  pbs <- aggregateAcrossCells(spe, 
    ids=spe[[clustering]], subset.row=top, 
    use.assay.type="logcounts", statistics="mean")
  
  return(pbs)
}
pbsLeiden <- getTopDEGs(spe=spe, clustering="Leiden")
pbsBanksy <- getTopDEGs(spe=spe, clustering="Banksy")

plotHeatmap(pbsLeiden, 
                    colour_columns_by="Leiden", fontsize=9,
                    features=rownames(pbsLeiden), center=TRUE, scale=TRUE,
                    cluster_cols=FALSE, cluster_rows=FALSE)

plotHeatmap(pbsBanksy, 
                    colour_columns_by="Banksy", fontsize=9,
                    features=rownames(pbsBanksy), center=TRUE, scale=TRUE,
                    cluster_cols=FALSE, cluster_rows=FALSE)
```

```{r plot-xy-expr, fig.width=4, fig.height=3, eval=FALSE, message=FALSE}
#| collapse: TRUE
es <- as.matrix(t(logcounts(spe)[rownames(pbsBanksy), ]))
fd <- data.frame(df, es, check.names=FALSE)
gene_plot <- ggplot(fd, aes(x_centroid, y_centroid, col=MZB1)) +
    coord_equal(expand=FALSE) +
    geom_point(shape=16, stroke=0, size=0.2) +
    scale_color_gradientn(colors=c("navy", "red", "gold", "ivory")) +
    theme_xy + theme(legend.key.height=unit(1, "lines"))

cluster_plot <- ggplot(df, aes(x_centroid, y_centroid, col=(Banksy_clusters ==3))) + 
  geom_point(shape=16, stroke=0, size=0.2) +
  scale_color_manual(values=c("navy", "cyan")) + 
  ggtitle("Banksy Cluster 3") + 
  coord_equal(expand=FALSE) +
  theme_void() + 
  theme(legend.position="none",
        plot.title=element_text(hjust=0.5),
        panel.background=element_rect(fill="black", color="white"))

gene_plot | cluster_plot
```

### Comparison of DEGs in different spatially-aware clustering cases (collapse)
-->

## Appendix

### Reviews {.unnumbered}

* [https://doi.org/10.1093/bfgp/elae040](@Wang2024) categorize ST clustering approaches into probability statistics-, graph neural network- and contrastive learning-based approaches, and review their advantages and limitations for clustering of data from seq- and img-based assays.

### Benchmarks {.unnumbered}

* [https://doi.org/10.1186/s13059-024-03361-0](@Hu2024) compare 16 graph- and statistic-based approaches, including `BayesSpace` and `BANKSY`, across 10 (mostly seq-based) datasets.
* [https://doi.org/10.1016/j.csbj.2023.11.055](@Liu2024) compare eight GNN-based approaches, including `SpaceFlow`, across four seq- and one img-based dataset.

### References {.unnumbered}
