# Visium HD workflow {#sec-img-workflow-visiumhd}

## Preamble

### Introduction
<!-- TODO -->

### Dependencies
```{r load-libs}
## QC 
library(SpotSweeper)
library(VisiumIO)

library(RANN)
library(here)
library(SpatialExperiment)
library(spatialLIBD)
library(escheR)
library(patchwork)
library(scran)
library(scater)
library(tidyr)
library(dplyr)
```

We retrieve VisiumHD dataset at two different resolutions of bin size with square of 8um and 16um. 
```{r cache-vhd, results="hold", message=FALSE, warning=FALSE}
id <- "VisiumHD_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
```

```{r load-obj8um}
# importing
obj8 <- TENxVisiumHD(
    spacerangerOut=td, 
    bin_size = "016", 
    images="lowres", 
    format="h5")
(spe8 <- import(obj8))
```

```{r check-coldata}
colnames(colData(spe))
#  [1] "barcode"                       "in_tissue"                    
#  [3] "array_row"                     "array_col"                    
#  [5] "sample_id"                     "sum"                          
#  [7] "detected"                      "subsets_mito_percent"         
#  [9] "subsets_mito_sum"              "sum_discard"                  
# [11] "detected_discard"              "subsets_mito_percent_discard" 
# [13] "sum_log"                       "sum_outliers"                 
# [15] "sum_z"                         "detected_log"                 
# [17] "detected_outliers"             "detected_z"                   
# [19] "subsets_mito_percent_log"      "subsets_mito_percent_outliers"
# [21] "subsets_mito_percent_z"        "BS_k10" 
```

```{r fixed-QC}
# ====== Add discards by fixed thresholds ========
spe$detected_threshold <- spe$detected < 500
spe$sum_threshold <- spe$sum < 500
spe$subsets_mito_percent_threshold <- spe$subsets_mito_percent >10
```


```{r miQC-flags}
# ====== Add discards by miQC ======
library(miQC)

model <- mixtureModel(spe)

png(here(plot_dir, "miQC_mixtureModel.png"), width=4, height=4, units="in", res=300)
plotModel(spe, model) + ggtitle("HD Breast Cancer")
dev.off()

png(here(plot_dir, "miQC_mixtureModelFiltering.png"), width=4, height=4, units="in", res=300)
plotFiltering(spe, model) + ggtitle("HD Breast Cancer")
dev.off()
```

```{r miQC-manual-flags}
# NOTE: miQC does not offer a way to simply flag spots, so I'm importing part of the filter function here to do so manually
library(flexmix)

posterior_cutoff = 0.75
keep_all_below_boundary = TRUE
enforce_left_cutoff = TRUE

metrics <- as.data.frame(colData(spe))

    if (is.null(model)) {
        warning("call 'mixtureModel' explicitly to get stable model features")
        model <- mixtureModel(spe)
    }

    intercept1 <- parameters(model, component = 1)[1]
    intercept2 <- parameters(model, component = 2)[1]
    if (intercept1 > intercept2) {
        compromised_dist <- 1
        intact_dist <- 2
    } else {
        intact_dist <- 1
        compromised_dist <- 2
    }

    post <- posterior(model)
    metrics$prob_compromised <- post[, compromised_dist]
    spe$prob_compromised <- metrics$prob_compromised
    metrics$keep <- metrics$prob_compromised <= posterior_cutoff

    if (sum(metrics$keep)==nrow(metrics)){
        stop("all cells passed posterior probability filtering. One 
              cause of this is the model selecting two near-identical
              distributions. Try rerunning mixtureModel() and/or 
              setting a different random seed.")
    }
    
    if (keep_all_below_boundary == TRUE) {
        predictions <- fitted(model)[, intact_dist]
        metrics$intact_prediction <- predictions
        metrics[metrics$subsets_mito_percent <
                    metrics$intact_prediction, ]$keep <- TRUE
    }

    if (enforce_left_cutoff == TRUE) {
        min_discard <- min(metrics[!metrics$keep, ]$subsets_mito_percent)
        min_index <- which(metrics$subsets_mito_percent == min_discard)[1]
        lib_complexity <- metrics[min_index, ]$detected
        metrics[metrics$detected <= lib_complexity &
                    metrics$subsets_mito_percent >= min_discard, ]$keep <- FALSE
    }

# add "outliers" to SPE
spe$miQC_keep <- !metrics$keep

```


```{r spotsweeper}
# ======= SpotSweeper =========
spe <- localOutliers(spe,  n_neighbors = 48, metric = "sum", direction = "lower", log = TRUE)
spe <- localOutliers(spe,  n_neighbors = 48, metric = "detected", direction = "lower", log = TRUE)
spe <- localOutliers(spe,  n_neighbors = 48, metric = "subsets_mito_percent", direction = "higher", log = TRUE)

# save
saveRDS(spe, here("processed-data", "VisiumHD", "human_breast", "VisiumHD_HumanBreast_016_spotsweeper.rds"))
```


Visualization with Loupe browswer, can select a region and subset & download. 
# same for Xenium chapter, mention Xenium explorer

bin2cell's package has ploting functions. 

### Supervised

For comparison, we annotate the Xenium data using a label transfer approach, 
`r BiocStyle::Biocpkg("SingleR")`, which relies on labeled scRNA-seq data to 
compute references profiles and transfers labels based on the rank correlation 
between observed (here, Xenium) and reference (scRNA-seq) expression profiles.

First, we retrieve a matching (Chromium) scRNA-seq dataset, which includes low- 
(`Level1`) and high-resolution (`Level2`) annotations of cells into 9 and 31 
subpopulations, respectively:

```{r load-sce, results="hold", message=FALSE, warning=FALSE}
# retrieve dataset from OSF repository
id <- "Chromium_HumanColon_Oliveria"
pa <- OSTA.data::.load_data(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# read into 'SingleCellExperiment'
sce <- read10xCounts(list.files(td, "h5$", full.names=TRUE))
cd <- read.csv(list.files(td, "cell_meta", full.names=TRUE))
colData(sce) <- cbind(colData(sce), cd[, -1])
table(sce$Level1) # tabulate low-res. labels
ncol(sce) # overall number of cells
```


## References {.unnumbered}


