## Differential Colocalisation {#sec-crs-multi-sample}

In this section we will explore tools to assess differential co-localisation between conditions with the two packages `r BiocStyle::Biocpkg('spicyR')` and `r BiocStyle::Biocpkg('spatialFDA')`. Both packages are based the comparison of known spatial statistics functions implemented in the package `r BiocStyle::CRANpkg('spatstat')`. 

```{r, setup, warning=FALSE, error=FALSE, message=FALSE}
library("spatialFDA")
library("spicyR")
library("refund")
library("dplyr")
library("stringr")
library("tidyr")
```

The use case is a dataset on type I diabetes (T1D) [@damondMapHumanType2019]. This dataset contains three conditions (Non-diabetic, onset and long-druation diabetes), several patients per condition and per patient several fields-of-view. This hierarchical set-up requires the modeller to account for correlation between fields-of-view if they stem e.g. from the same patient. Such a setup can be modelled with mixed-models.

```{r, load-data, warning=FALSE, error=FALSE}
# load data
spe <- .loadExample(full = TRUE)
# subset data to smaller example
spe <- subset(spe, ,patient_id %in% c(6089,6180,6126,6134,6228,6414))
# as factor
colData(spe)$cell_type <- as.factor(colData(spe)$cell_type) 
# as factor and relevel
colData(spe)$patient_stage <- as.factor(colData(spe)$patient_stage) |> relevel("Non-diabetic") 
```

### `spicyR`

First, we show the functionality of `r BiocStyle::Biocpkg('spicyR')`. This package condenses the information of a spatial statistics curve to a scalar value and models this continuous value as the response in a linear (mixed) effects model [@Canete2022]. 

```{r, spicyR, warning=FALSE, error=FALSE}
colData(spe)$cellType <- colData(spe)$cell_type
colData(spe)$cell_type <- as.factor(colData(spe)$cell_type) 
spicyTestPair <- spicy(
  spe,
  condition = "patient_stage",
  imageID = "image_number",
  window = 'square',
  cores = 4
)

out <- topPairs(spicyTestPair, n = 27)
out
```

We can plot the results of `r BiocStyle::Biocpkg('spicyR')` as a heatmap

```{r, warning=FALSE, error=FALSE}
signifPlot(
  spicyTestPair,
  breaks = c(-3, 3, 1),
)
```

### `spatialFDA`

Code adapted from [spatialFDA vignette](https://bioconductor.org/packages/devel/bioc/vignettes/spatialFDA/inst/doc/DiabetesIsletExample.html)

`r BiocStyle::Biocpkg('spatialFDA')` is a package that computes spatial statistics functions and compares these with tools from functional data analysis (FDA). We will consider the co-localisation of $\alpha$ cells in the islet and cytoxic T-cells which are important in the destruction of islet cells in T1D [@damondMapHumanType2019]. 

```{r, calcMetric, warning=FALSE, error=FALSE}
metricRes <- calcMetricPerFov(spe = spe, selection = c("alpha", "Tc"),
                              subsetby = "image_number", fun = "Gcross", 
                              marks = "cell_type",
                              rSeq = seq(0, 50, length.out = 50), 
                              by = c("patient_stage", "patient_id",
                                     "image_number"),
                              ncores = 1)
```

First, we will plot the functional box plot of the square-root transformed spatial statistics functions. [@sun2011functional; @ramsay2024fda].

```{r, visualise, warning=FALSE, error=FALSE}
metricRes$ID <- paste0(
    metricRes$patient_stage, "x", metricRes$patient_id,
    "x", metricRes$image_number
)
#removing field of views that have as a curve only zeros - these are cases where
#there is no cells of one type
metricRes <- metricRes %>% dplyr::group_by(ID) %>% dplyr::filter(sum(rs) >= 1)

#sqrt transform response
metricRes$rs <- sqrt(metricRes$rs)

collector <- plotFbPlot(metricRes, "r", "rs", "patient_stage")
```

The black line indicates the median curve, the purple region is the 50% central region and the red lines indicate outliers. 

From these plots we not that onset functions seem to qualitatively have a higher median curve than non-diabetic or long-duration diabetes functions.

Another exploratory technique is functional PCA (FPCA). FPCA is similar to standard PCA in that it decomposes the functional data into the main modes of variation. Here, we decompose into eigenfunctions [@ramsayPrincipalComponentsAnalysis2005; @refund2024].  

```{r, fPCA, warning=FALSE, error=FALSE}
# filter out all rows that have a constant zero part - all r<10
metricRes <- metricRes %>% filter(r > 10)

# prepare dataframe from calcMetricRes to be in the correct format for pffr
dat <- prepData(metricRes, "r", "rs")

# create meta info of the IDs
splitData <- dat$ID %>%
  str_replace("-","_") %>%
  str_split_fixed("x", 3) %>% 
  data.frame(stringsAsFactors = TRUE) %>%
  setNames(c("condition", "patient_id", "imageId")) %>%
  mutate(condition = relevel(condition,"Non_diabetic"))
dat <- cbind(dat, splitData)

# drop rows with NA
dat <- dat |> drop_na()
# calculate the fPCA
pca <- functionalPCA(dat = dat, r = metricRes$r |> unique(), pve = 0.995)
evalues <- pca$evalues
efunctions <- pca$efunctions

plotFpca(dat = dat, res = pca, colourby = "condition")
```

From the FPCA we note that hte variability in onset diabetes curves is larger than the other two categories.

Next, we will show how FDA techniques can be used to make inference on the spatial statistics curves. We use a generalised additive mixed model with a functional response [@scheiplFunctionalAdditiveMixed2015; @scheiplGeneralizedFunctionalAdditive2016; @refund2024]. 

```{r, fGAM, warning=FALSE, error=FALSE}
# create a design matrix
mm <- model.matrix(~condition, data = dat)
colnames(mm)[1] <- "Intercept"
mm %>% head()

r <- metricRes$r |> unique()
# fit the model
mdl <- functionalGam(
    data = dat, x = r,
    designmat = mm, weights = dat$npoints,
    formula = formula(Y ~ 1 + conditionLong_duration +
                          conditionOnset + s(patient_id, bs = "re")),
    family = mgcv::scat(link = "log"),
    algorithm = "gam"
)
summary(mdl)

plotLs <- lapply(colnames(mm), plotMdl, mdl = mdl,
                 shift = mdl$coefficients[["(Intercept)"]])

patchwork::wrap_plots(plotLs, nrow = 3, axes = 'collect')
```

With the functional GAM we see not only the effect (increased or decreased co-localisation in comparison to the reference category) but also at which scale this effect is strong.

```{r, warning=FALSE, error=FALSE}
qqnorm(resid(mdl), pch = 16)
qqline(resid(mdl))
```

The Q-Q plot indicates an ok model fit, though not perfect. 


