# Processing {#sec-img-processing}

## Preamble

### Introduction

TODO

### Dependencies

```{r load-libs, message=FALSE, warning=FALSE}
library(dplyr)
library(Banksy)
library(scater)
library(patchwork)
library(SpatialExperiment)
# set seed for random number generation
# in order to make results reproducible
set.seed(20000229)
# load data from preceding 
# chapter (post quality control)
(spe <- readRDS("img-spe_qc.rds"))
```

## Normalization

Library size-based normalization as is typical for scRNA-seq data has been shown to be problematic for ST data, especially so for targeted panels underlying current commercial imaging-based ST platforms [@Atta2024,@Bhuva2024].
For lack of a better approach, we here use standard log-library size normalization.
We caution readers, however, to keep an eye out in the literature for attempts to provide a better strategy.

```{r log}
spe <- logNormCounts(spe)
```

## Feature selection

At this stage, we would typically perform selection of (e.g., highly variable) features; see @sec-seq-processing.
The dataset at hand, however, is targeted and relatively low-plex, so that features have already been selected by design (e.g., different targets will be included in immuno-oncology and neuroscience panels).

## Dimension Reduction

### Non-spatial

As a baseline, we will perform principal component analysis (PCA), which underlies many standard scRNA-seq analysis pipelines, such as (spatially unaware) graph-based clustering based on a shared nearest neighbor (SNN) graph and the Leiden or Louvain algorithm for community detection.

```{r dr-tx}
spe <- runPCA(spe, ncomponents=20)
```

### Spatially aware

`r BiocStyle::Biocpkg("BANKSY")` [@Singhal2024-BANKSY] essentially computes PCs that are weighted by embeds cells in a product space of their own and the local neighborhood transcriptome, representing cell state and microenvironment.
A key parameter for this method is $\lambda\in[0,1]$ (argument `lambda` in `runBanksyPCA()`), which controls the amount of spatial information to incorporate; specifically, **when $\lambda=0$, `BANKSY` reduces to non-spatial clustering**.

```{r dr-sp, message=FALSE}
spe <- computeBanksy(spe, assay_name="logcounts")
spe <- runBanksyPCA(spe, npcs=20, lambda=0.2)
```

## Exploratory

To not confuse different types of PCs, we rename `reducedDims` to end in `_sp` and `_tx` for spatially aware and unaware results, respectively.

```{r prep-data}
reducedDimNames(spe) <- c("PCA_tx", "PCA_sp")
```

A useful visualization is to color cells by their PCs in physical space. 
This will help highlight key drivers of transcriptional variability, e.g., between major biological compartments such as epithelia, immune and stromal cells.
Exemplary plots of PCs 1-3 are rendered below:

```{r plot-xy-pcs, collapse=TRUE, fig.width=11, fig.height=3}
#| code-fold: true
df <- data.frame(colData(spe), reducedDim(spe, "PCA_tx"))
lapply(paste0("PC", seq(3)), \(.) {
    fd <- df[order(abs(df[[.]])), ]
    ggplot(fd, aes(x_centroid, y_centroid, col=.data[[.]]))
}) |>
    wrap_plots(nrow=1) &
    geom_point(shape=16, stroke=0, size=0.2) &
    scale_color_gradientn(colors=pals::jet()) &
    coord_equal() & theme_void() & theme(
        legend.position="bottom",
        legend.key.width=unit(0.8, "lines"),
        legend.key.height=unit(0.4, "lines"))
```

::: {.callout-note title="PCs as RGB" collapse="true"}

A neat trick in this context is to rescale PC 1-3 values between 0 and 1, and interpret them as RGB values for visualization:

```{r pcs-to-rgb}
pcs <- reducedDim(spe, "PCA_tx")[, seq(3)]
pcs <- sweep(pcs, 2, colMins(pcs), `-`)
pcs <- sweep(pcs, 2, colMaxs(pcs), `/`)
rgb <- apply(pcs, 1, \(.) rgb(.[1], .[2], .[3]))
df <- data.frame(colData(spe), rgb)
```

E.g., we see that colors corresponding to opposing PC values are clearly separated in space, indicating distinct drivers of transcriptional variability.
In some regions, we see color mixtures, indicating cells of ambiguous transcriptional profiles (with respect to what is being captured by the 3 PCs considered here).

```{r plot-xy-rgb, fig.width=6, fig.height=4}
#| code-fold: true
ggplot(df, aes(x_centroid, y_centroid, col=rgb)) +
    geom_point(shape=16, stroke=0, size=0.4) +
    scale_color_identity() + coord_equal() + theme_void()
```

:::

## Clustering

### Non-spatial

As a baseline, we will perform graph-based clustering.
Specifically, we construct a shared nearest neighbor (SNN) graph based on non-spatial PCs (`reducedDim` slot `"PCA_tx"`), and use the Leiden algorithm for community detection.

```{r clu-Leiden}
# build cellular shared nearest-neighbor (SNN) graph
g <- buildSNNGraph(spe, use.dimred="PCA_tx", type="jaccard", k=20)
# cluster using Leiden community detection algorithm
k <- cluster_leiden(g, objective_function="modularity", resolution=0.8)
table(spe$Leiden <- factor(k$membership))
```

### Spatially aware

`r BiocStyle::Biocpkg("BANKSY")` [@Singhal2024-BANKSY] embeds cells in a product space of their own and their local neighborhood's (average) transcriptome, representing cell state and microenvironment. 
Here, the mixing parameter $\lambda\in[0,1]$ (argument `lambda` in `runBanksyPCA()`), controls the spatial component's weight; e.g., **when $\lambda=0$, `BANKSY` reduces to non-spatial clustering**.

```{r clu-Banksy, message=FALSE}
# retrieve 'lambda' used for dimension reduction
l <- metadata(spe)$BANKSY_params$lambda 
# perform 'Banksy' clustering
spe <- clusterBanksy(spe, dimred="PCA_sp", resolution=0.4)
# reformat cluster assignments
k <- grep("^clust", names(colData(spe)))
names(colData(spe))[k] <- "Banksy"
table(spe$Banksy <- factor(spe$Banksy))
```

## Visualization

Let's visualize the assignment obtains from non-spatial and spatially aware clustering:

```{r plot-xy-k-joint, message=FALSE, fig.width=12, fig.height=3}
#| code-fold: true
# TODO: 'ggspavis' is being a bit pedantic...
spe$in_tissue <- 1; spe$x_centroid <- spe$y_centroid <- NULL
lapply(c("Label", "Leiden", "Banksy"), \(.) {
    plt <- plotSpots(spe, annotate=.)
    plt$layers[[1]]$aes_params$stroke <- 0
    plt$layers[[1]]$aes_params$size <- 0.2
    plt
}) |>
    wrap_plots(nrow=1) &
    theme(legend.key.size=unit(0, "lines")) &
    scale_color_manual(values=unname(pals::trubetskoy()))
```

## Appendix

### Saving {.unnumbered}

```{r save-data}
saveRDS(spe, "img-spe_dr.rds")
```

### Benchmarks {.unnumbered}

### References {.unnumbered}
