# Imputation


## Introduction

The analyses demonstrated here will use 313-plex Xenium data (10x Genomics) of a human breast cancer biopsy section [@Janesick2023] and the 1k-plex CosMx data (Nanostring) of a mouse brain sample. 

## Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(scater)
library(harmony)
library(OSTA.data)
library(SpatialExperiment)
library(SpatialExperimentIO)
library(DropletUtils)
library(Matrix)
library(RANN)
```

## Data import

`r BiocStyle::Biocpkg("SpatialExperimentIO")` provides functions to import CosMx, Xenium, and MERSCOPE datasets in Bioconductor, starting from the output provided by the three platforms.

### Xenium

We now load Xenium data:

```{r, message=FALSE, warning=FALSE}
id <- "Xenium_HumanBreast1_Janesick"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
spe <- readXeniumSXE(td)
spe$sample_id <- "JanesickBC_Xenium1"
spe
```

### Single Cell

We now load the single cell data:

```{r, message=FALSE, warning=FALSE}
id <- "Chromium_HumanBreast_Janesick"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
h5 <- file.path(td, "filtered_feature_bc_matrix.h5")
sce <- read10xCounts(h5, col.names=TRUE)
metadata(sce) <- list()
sce$sample_id <- "Janesick_Chromium"
sce

# set gene symbols as feature names
rownames(sce) <- make.unique(rowData(sce)$Symbol)
```

## Harmony Integration

Besides physically aligning both datasets, we can integrate them on 
a transcriptional level; here, using `r BiocStyle::Biocpkg("harmony")`.

To this end, we first consolidate the Visium and Xenium data into one object:

```{r obj}
# convert to single cell
spe_single <- as(spe, "SingleCellExperiment")

# get common genes
gs <- intersect(rownames(spe_single), rownames(sce))
cd <- intersect(names(colData(spe_single)), names(colData(sce)))
lys <- list(spe_single=spe_single, sce=sce)
lys <- lapply(lys, \(scRNA) {
    scRNA <- scRNA[gs,]
    rowData(scRNA) <- rowData(scRNA)[gs,]
    colData(scRNA) <- colData(scRNA)[cd]
    assay(scRNA) <- as(assay(scRNA), "dgCMatrix")
    return(scRNA)
})

# combine datasets
obj <- SingleCellExperiment(assays = list(counts = cbind(assay(lys[[1]]), assay(lys[[2]]))),
                            colData = rbind(colData(lys[[1]]), colData(lys[[2]])),
                            rowData = rowData(lys[[1]]))
obj$col_counts <- colSums(assay(obj))
```

We now initiate the harmony integration:

```{r int}
# minimal filtering
obj <- obj[, obj$col_counts > 0]
# library size normalization 
obj <- logNormCounts(obj)
# principal component analysis
# (w/o additional feature selection)
obj <- runPCA(obj)
# 'harmony' integration
pcs <- RunHarmony(
    data_mat=reducedDim(obj, "PCA"), 
    meta_data=obj$sample_id, 
    verbose=FALSE)
reducedDim(obj, "PCA") <- pcs
# dimensionality reduction
map <- calculateUMAP(t(pcs))
reducedDim(obj, "UMAP") <- map
```

let us now visualize the embedding of both Xenium and chromium cells to assess the mixing

```{r plt-map, fig.width=3.5, fig.height=3}
plotUMAP(obj, colour_by="sample_id", point_size=0.1) +
    guides(col=guide_legend(override.aes=list(alpha=1, size=2))) +
    theme_void() + theme(aspect.ratio=1, legend.key.size=unit(0, "pt"))
```
## Imputation

Now that we show there is a correction of batch effects in the PC space. We can transfer counts of single cell data to the Xenium dataset by detecting kNNs of xenium cells in PC space and aggregate missing genes over nearest neighbors.

```{r impute}
embeddings <- reducedDim(obj, "PCA")

# find kNN across reference and query
nNeighbors <- 20
reference_embedding <- embeddings[obj$sample_id == "JanesickBC_Xenium1",]
query_embedding <- embeddings[obj$sample_id == "Janesick_Chromium",]
knn <- RANN::nn2(query_embedding, 
                 reference_embedding, 
                 k = nNeighbors)
weights <- Matrix::Matrix(0, 
                          nrow = nrow(reference_embedding), 
                          ncol = nrow(knn$nn.idx), 
                          sparse = TRUE)
for (n in seq_len(nrow(knn$nn.idx))) {
  weights[knn$nn.idx[n, ], n] <- 1/nNeighbors
}

# impute by kNN
query_counts <- assay(obj)[,obj$sample_id == "Janesick_Chromium"]
imputed_reference_counts <- query_counts %*% weights
colnames(imputed_reference_counts) <- rownames(reference_embedding)
rownames(imputed_reference_counts) <- colnames(reference_embedding)
```

## Visualization

## References {.unnumbered}
