# Differential colocalization {#sec-mult-diff-colocalization}

## Preamble

### Introduction

In this section, we will explore tools to assess differential colocalization between conditions with the two packages `r BiocStyle::Biocpkg("spicyR")` and `r BiocStyle::Biocpkg("spatialFDA")`. Both packages are based the comparison of known spatial statistics functions implemented in the package `r BiocStyle::CRANpkg("spatstat")`.


### Dependencies

```{r, setup, warning=FALSE, message=FALSE}
library(osfr)
library(dplyr)
library(tidyr)
library(refund)
library(spicyR)
library(stringr)
library(ggplot2)
library(patchwork)
library(spatialFDA)
library(SpatialExperiment)
```


### Load data

For this example, we use a dataset on type I diabetes (T1D) [@Damond2019-IMC-diabetes]. This dataset contains three conditions (non-diabetic, onset, and long-duration diabetes), several patients per condition, and several fields-of-view per patient. This hierarchical setup requires the modeler to account for correlation between fields-of-view, for example if they stem from the same patient. Such a setup can be modeled with mixed models (see below).

```{r, load-data-hide, warning=FALSE, eval=FALSE, echo=FALSE}
# load data from full dataset (not run due to computational limitations)
spe <- .loadExample(full = TRUE)
# subset data to smaller example
spe <- spe[, spe$patient_id %in% c(6089, 6180, 6126, 6134, 6228, 6414)]
# remove assays (for smaller object size)
assays(spe) <- list()
```

An assay-free version of the dataset is available through our OSF repository
(used here to reduce runtime as the following analyses do not require assay data):

```{r, load-data, message=FALSE}
# load data from OSF repository (for smaller download)
osf_repo <- osf_retrieve_node("https://osf.io/5n4q3/")
osf_files <- osf_ls_files(osf_repo, 
    path="zzz", n_max=Inf,
    pattern="Damond_noAssays.rds")
dir.create(td <- tempfile())
foo <- osf_download(osf_files, path=td)

# read into R & subset patient IDs (to reduce runtime)
spe <- readRDS(file.path(td, list.files(td, ".rds")))
ids <- c(6089, 6180, 6126, 6134, 6228, 6414)
(spe <- spe[, spe$patient_id %in% ids])
```


## spicyR

First, we show the functionality of `r BiocStyle::Biocpkg("spicyR")`. 
This package condenses the information of a spatial statistics curve 
to a scalar value and models this continuous value as the response 
in a linear (mixed) effects model [@Canete2022-spicyR]. 

By default, `spicyR` chooses the first level of the `condition` 
factor as the base, and the second as the comparison condition;
we first refactor patient stages accordingly:

```{r prep-data}
stages <- c("Non-diabetic", "Onset", "Long-duration")
spe$patient_stage <- factor(spe$patient_stage, stages)
spe$cell_type <- factor(spe$cell_type)
```

```{r, spicyR, message=FALSE, warning=FALSE}
# note: runtime ~5 min
spicyTestPair <- spicy(
    cells=spe, 
    condition="patient_stage", 
    imageID="image_name", 
    cellType="cell_type",
    window="square", 
    cores=4)
# extract most significant pairs
out <- topPairs(spicyTestPair, n=27)
head(out)
```

The results can be represented as a bubble plot using the `signifPlot` function.

```{r, spicyR-heatmap, warning=FALSE, message=FALSE, fig.width=8, fig.height=6}
signifPlot(spicyTestPair)
```

Here, we can observe that the most significant relationship occurs between 
cytotoxic T cells (Tc) and the beta cells of the pancreas. Tc and beta cells 
appear to be significantly more co-localized in patients with onset diabetes 
compared to non-diabetic patients, indicated by the positive coefficient.

We can use `spicyBoxPlot` to examine a cell type-cell type relationship closely.

```{r, spicyR-boxplot, warning=FALSE, fig.width=5, fig.asp=1}
spicyBoxPlot(spicyTestPair, from="Tc", to="beta")
```

We can investigate individual images to ensure our results reflect biologically 
meaningful relationships. We can extract the individual co-localization metrics 
from the `spicyTest` object using the `bind` function.

```{r spicyR-metadata}
# extract co-localization metrics
cd <- colData(spe) |>
    as.data.frame() |> 
    distinct(patient_id, patient_stage, image_name) |>
    dplyr::rename(imageID=image_name)
bind(spicyTest) |> 
    # merge with metadata
    merge(cd, by="imageID", all=FALSE) |>
    # select columns of interest
    select(c(imageID, Tc__beta, patient_id, patient_stage)) |>
    filter(imageID %in% c("L29", "O02", "E17", "G17"))
```

We can then visualize the spatial relationship between a pair of 
cell types in the original image using the `plotImage` function.

- The left image (L29) comes from a non-diabetic patient. 
We can see that beta and Tc cells appear to be dispersed, 
which is reflected in the co-localization metric.

- The right image (O02) comes from a patient with early onset diabetes. 
We can see that beta and Tc cells are far more co-localized, 
with Tc cells even infiltrating the pancreatic islets in some cases.

```{r spicyR-plotImage, fig.width=10, fig.height=4.5}
spe$cellType <- spe$cell_type # bug? setting 'cellType' doesn't work
plotImage(spe, imageToPlot="L29", from="Tc", to="beta", imageID="image_name") +
plotImage(spe, imageToPlot="O02", from="Tc", to="beta", imageID="image_name") +
    plot_layout(nrow=1) &
    coord_equal()
```

`spicyR` can also be used with custom distance metrics and to perform 
survival analysis, which are outlined in the [spicyR vignette](
https://www.bioconductor.org/packages/devel/bioc/vignettes/spicyR/inst/doc/spicyR.html).


## spatialFDA

The `r BiocStyle::Biocpkg("spatialFDA")` package computes spatial summaries 
as functions (e.g., Ripley's $K$ function) and compares them using tools from 
functional data analysis (FDA). We will consider the colocalization (spacing) 
of alpha cells in the islet and cytoxic T-cells, which are important in the 
destruction of islet cells in T1D [@Damond2019-IMC-diabetes].

This example is adapted from the [spatialFDA vignette](
https://bioconductor.org/packages/devel/bioc/vignettes/spatialFDA/inst/doc/DiabetesIsletExampleBrief.html).

```{r, spatialFDA-calc, results="hide", message = FALSE, warning = FALSE}
colData(spe)[["patient_stage"]] <- factor(colData(spe)[["patient_stage"]])
#relevel to have non-diabetic as the reference category
colData(spe)[["patient_stage"]] <- relevel(colData(spe)[["patient_stage"]],
"Non-diabetic")

#run the spatial statistics inference
resLs <- crossSpatialInference(
    spe, 
    selection = c("alpha", "Tc", "Th"),
    subsetby = "image_name", 
    fun = "Gcross", 
    marks = "cell_type",
    rSeq = seq(0, 50, length.out = 50), 
    correction = "rs",
    sample_id = "patient_id",
    transformation = "Fisher",
    eps = 1e-3,
    delta = "minNnDist",
    family = mgcv::scat(link = "log"),
    image_id = "image_name", 
    condition = "patient_stage",
    ncores = 1,
    algorithm = "bam"
)

names(resLs)
```

The convenience function `crossSpatialInference` calculates first the spatial statistics
functions and then compares these with a functional additive mixed model across all cell types indicated in `selection`. [@Baddeley2015-book-point-patterns; @Baddeley2005-spatstat; @Scheipl2015-additive; @Scheipl2016-generalized; @Goldsmith2024-refund-package]. The result is a list of lists containing the results across all cell types. 

We can plot the overall picture of the interactions by performing an $F$-test on all functional models and plotting both the $p$-value of the tests as well as the mean coefficient over the domain $r$.

```{r, spatialFDA-bubbleplot, warning=FALSE, message=FALSE, fig.width=8, fig.height=4, results='hide'}
p <- plotCrossHeatmap(resLs, coefficientsToPlot = c("conditionLong_duration(x)",
                                                    "conditionOnset(x)"), 
                      QCThreshold = 0, QCMetric = "medianMinIntensity")

p <- p + guides(shape = "none") + 
  facet_wrap(~factor(coefficient, 
                     levels = c("conditionOnset(x)", "conditionLong_duration(x)"), 
                     labels = c("conditionOnset(x)" = "Onset", 
                                "conditionLong_duration(x)" = "Long-Duration")))

p
```


We will focus on the cell type pair $\alpha$-cytotoxic T cells in more detail below.

```{r, spatialFDA-results-selection, warning=FALSE, message=FALSE, fig.width=4, fig.height=4}
res <- resLs$alpha_Tc
names(res)
```

Each cell type pair then contains the transformed functions (`metricRes`), the untransformed functions (`metricResRaw`), the design matrix used to fit the model (`designmat`), the actual functional model (`mdl`) and some QC metrics of the statistical inference (`curveFittingQC`).

First, we will plot the functional box plot of the transformed spatial statistics functions [@Sun2011-boxplots].

```{r, spatialFDA-plot, warning=FALSE, message=FALSE, fig.width=4, fig.height=4}
#| layout-ncol: 3
#| 
# extract the metric dataframe
metricRes <- res$metricRes

# make unique identifiers
metricRes$ID <- with(metricRes, paste(sep="x", 
    patient_stage, patient_id, image_name))

# functional boxplot of spatial statistics curves
collector <- plotFbPlot(metricRes, "r", "rs", "patient_stage")
```

The black line indicates the median curve, the purple region 
is the 50% central region, and the red lines indicate outliers.

From these plots, we note that the functions from 'Onset' 
patients seem to qualitatively have a higher median curve 
than those from 'Non-diabetic' or 'Long-duration' patients.

Another exploratory technique is functional PCA (FPCA). FPCA is similar to 
standard PCA in that it decomposes the functional data into the main modes of 
variation. Here, we decompose into eigenfunctions [@Goldsmith2024-refund-package].

```{r, spatialFDA-fPCA, fig.width=8, fig.height=3}
# prepare 'data.frame' from 'calcMetricRes' 
# to be in the correct format for FDA
dat <- prepData(metricRes, "r", "rs")

# create meta info of the IDs
splitData <- dat$ID |> 
    str_replace("-","_") |> 
    str_split_fixed("x", 3) |> 
    data.frame(stringsAsFactors=TRUE) |> 
    setNames(c("condition", "patient_id", "imageId")) |> 
    mutate(condition=relevel(condition,"Non_diabetic"))

dat <- cbind(dat, splitData) # join with results
dat <- drop_na(dat)          # drop rows with NA

# calculate FPCA & visualize biplot
pca <- functionalPCA(dat=dat, r=unique(metricRes$r), pve=0.995)
plotFpca(dat=dat, res=pca, colourby="condition") + labs(color="condition")
```

From the FPCA, we note that the variability in curves 
from 'Onset' patients is larger than the other two groups.

Next, we will show how FDA techniques can be used to make inferences from the functional spatial summaries. We use a generalized additive mixed model with a functional response [@Scheipl2015-additive; @Scheipl2016-generalized; @Goldsmith2024-refund-package]. In such models, some parameters are functional and some are constants (see below).

```{r, spatialFDA-fGAM, message=FALSE, warning=FALSE, results="hide", fig.width=6}
mdl <- res$mdl
mm <- res$designmat

summary(mdl)

lapply(colnames(mm), \(x) {
    i <- mdl$coefficients[[1]] 
    plotMdl(mdl, predictor=x, shift=i)
}) |> wrap_plots(nrow=3, axes="collect")
```

With the functional GAM, we see not only the effect (e.g., increased 
colocalization of alpha and cytotoxic T cells in 'Onset' patients compared
to 'Non-diabetic' patients) but also the strength of the effect at each scale.

The Q-Q plot indicates a reasonable model fit, although not perfect.

```{r, spatialFDA-qq, fig.width=5}
qqnorm(mdl$residuals, pch=16)
qqline(mdl$residuals, col="red")
```


## Appendix

### References {.unnumbered}
