# Workflow: Visium HD {#sec-seq-workflow-visium-hd}

## Introduction

Visium HD is a next-generation spatial transcriptomics technology developed by 10x Genomics, designed to provide single-cell resolution spatial gene expression data across entire tissue sections. Commercially available since 2024 [@Oliveira2024], Visium HD advanced data resolution from 55 Âµm (spot-based resolution) to bins near subcellular resolution (2 Âµm). In this workflow, we will demonstrate common analysis steps with a Visium HD dataset on colorectal cancer [@Oliveira2024]. 

## Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(OSTA.data)
library(VisiumIO)
library(SpatialExperiment)
library(ggspavis)
library(spacexr)
library(ggplot2)
library(patchwork)
library(DropletUtils)
library(BiocParallel)
library(scran)
library(scater)
library(scuttle)
library(pheatmap)
library(tidyr)
library(SpotSweeper)
library(Banksy)
library(Statial)
```

In this demo, we perform deconvolution on 16 Âµm bins and compare the concordance with result on 8 Âµm bins, provided by 10x Genomics.
[Here, we only demonstrate the analysis for patient 2. Other patients, such as P1 and P5, also have public data available in CRC condition, with Chromium, Visium HD, and Xenium platforms (see @sec-bkg-datasets). It would be interesting to investigate the differences in malignant cells/bins (e.g. differentially expressed genes, pathway analysis, etc.) between patients across different platforms.]{.aside}

```{r load-vhd8, message=FALSE, warning=FALSE}
# retrieve dataset from OSF repo
id <- "VisiumHD_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)

# read 8um bins into 'SpatialExperiment'
vhd8 <- VisiumIO::TENxVisiumHD(
  spacerangerOut=td,
  images="lowres",
  processing="filtered",
  bin_size="008",
  format="h5"
) |> VisiumIO::import()
rownames(vhd8) <- make.unique(rowData(vhd8)$Symbol)

# retrieve 008um bin annotations
gz <- "binned_outputs/square_008um/deconvolution.csv.gz"
df <- read.csv(file.path(td, gz), row.names=1)
df <- df[complete.cases(df), ]
cx <- intersect(colnames(vhd8), rownames(df))

# keep only Visium HD 8um with a provided annotation
vhd8 <- vhd8[, cx]

# add as metadata
colData(vhd8)[names(df)] <- df[cx, ]

vhd8
```

Next, we read in Visium HD data for 16 Âµm bins. There is no annotation result for 16 Âµm bin in the public data download. Given the relatively smaller number of bins compared to 8 Âµm, our downstream analysis will be performed on a subset area of 16 Âµm bins.

```{r load-vhd16, message=FALSE}
# read 16um bins into 'SpatialExperiment'
vhd16 <- VisiumIO::TENxVisiumHD(
  spacerangerOut=td,
  images="lowres",
  processing="filtered",
  bin_size="016",
  format="h5"
) |> VisiumIO::import()
# use symbols as feature names
gs <- rowData(vhd16)$Symbol
rownames(vhd16) <- make.unique(gs)
vhd16
```

```{r vhd816-spavis, fig.width=12, fig.height=6.5}
#| code-fold: true
# vhd8$loglibsize <- log1p(colSums(counts(vhd8)))
# vhd16$loglibsize <- log1p(colSums(counts(vhd16)))
# plotSpots(vhd8, 
#     annotate="loglibsize", pal="viridis", 
#     point_size=0) + ggtitle("8 Âµm log-library size") 
# p <- plotSpots(vhd16, 
#     annotate="loglibsize", pal="viridis", 
#     point_size=0) + ggtitle("16 Âµm log-library size")
# p$layers[[1]]$aes_params$shape <- 16
# p$layers[[1]]$aes_params$stroke <- 0
# p$layers[[1]]$aes_params$size <- 0.3
# p
```

Note that 8 and 16 Âµm binned Visium HD data share the same full resolution H&E image and scaling factor. 
For 16 Âµm bins, we will perform analysis steps on only 1/64 of the tissue area that was measured using Visium HD (to reduce demo runtime).

```{r plotvis-cove-zoom-8-16, warning=FALSE, message=FALSE, fig.width=12, fig.height=4}
#| code-fold: true
get_range <- function(spe){
  xy <- spatialCoords(spe)*scaleFactors(spe)
  xs <- range(xy[, 1]); ys <- nrow(imgRaster(vhd8))-range(xy[, 2])
  xmi <- xs[1] + 4*(xs[2]-xs[1])/8; xma <- xs[2] - 3*(xs[2]-xs[1])/8
  ymi <- ys[1] + 3*(ys[2]-ys[1])/8; yma <- ys[2] - 4*(ys[2]-ys[1])/8
  return(list(zoom=c(xmi, xma, ymi, yma), cove=c(xs, ys)))
}
vhd8r <- get_range(vhd8)
vhd16r <- get_range(vhd16)

# 8um bounding box, should be similar to 16um 
get_box <- function(cate, ltp, lwd, col){
  box <- geom_rect(
    xmin=vhd8r[[cate]][1], xmax=vhd8r[[cate]][2], 
    ymin=vhd8r[[cate]][3], ymax=vhd8r[[cate]][4],
    col=col, fill=NA, linetype=ltp, linewidth=lwd)
  return(box)
}
box <- get_box(cate="zoom", ltp=4, lwd=2/3, col="black")
box2 <- get_box(cate="cove", ltp=2, lwd=1/2, col="grey")

plotVisium(vhd8, spots=FALSE, point_size=.75) + box + box2 +
  theme(plot.title=element_text(hjust=0.5)) + 
  ggtitle("grey box: data coverage \n black box: zoomed region") + 
plotVisium(vhd8, point_size=0.9, zoom=TRUE) + xlim(vhd8r[["zoom"]][c(1, 2)]) + ylim(vhd8r[["zoom"]][c(4, 3)]) + 
  theme(plot.title=element_text(hjust=0.5)) + 
  ggtitle("8 Âµm bins in black box") + 
  plot_spacer() +
plotVisium(vhd16, point_size=1.8, zoom=TRUE) + xlim(vhd16r[["zoom"]][c(1, 2)]) + ylim(vhd16r[["zoom"]][c(4, 3)]) + 
  theme(plot.title=element_text(hjust=0.5)) + 
  ggtitle("16 Âµm bins in black box") + 
plot_layout(nrow=1, widths=c(1.7, 1, 0.05, 1)) & facet_null()
```

Subset the 16 Âµm object to bins in the black box.

```{r vhd16-subset-zoom, message=FALSE}
get_subset_vhd <- function(spe, spe_range){
  xy <- spatialCoords(spe)*scaleFactors(spe)
  y_no_rev <- nrow(imgRaster(spe)) - spe_range[["zoom"]][c(3,4)]
  vhd16sub <- spe[, xy[, 1] > spe_range[["zoom"]][1] & 
                    xy[, 1] < spe_range[["zoom"]][2] & 
                    xy[, 2] > y_no_rev[1] & 
                    xy[, 2] < y_no_rev[2]]
  return(vhd16sub)
}

vhd16sub <- get_subset_vhd(spe=vhd16, spe_range=vhd16r)
dim(vhd16sub)

# plotVisium(vhd16sub, point_size=1.8, zoom=TRUE) # sanity check
```

Now we are operating on `r ncol(vhd16sub)` 16 Âµm bins.

## Quality control

As detailed in @sec-seq-quality-control, we use `r BiocStyle::Biocpkg("SpotSweeper")` to perform quality control on the subsetted 16 Âµm data.

```{r qc-vhd16-local}
# calculate per-cell QC metrics
mt <- grepl("^MT-", rownames(vhd16sub))
vhd16sub <- addPerCellQCMetrics(vhd16sub, subsets=list(mt=mt))
# determine outliers based on 
# - low log-library size
# - few uniquely detected features
# - high mitochondrial count fraction
vhd16sub <- localOutliers(vhd16sub, metric="sum", direction="lower", log=TRUE)
vhd16sub <- localOutliers(vhd16sub, metric="detected", direction="lower", log=TRUE)
vhd16sub <- localOutliers(vhd16sub, metric="subsets_mt_percent", direction="higher", log=TRUE)
vhd16sub$discard <- 
    vhd16sub$sum_outliers | 
    vhd16sub$detected_outliers | 
    vhd16sub$subsets_mt_percent_outliers
# tabulate observations retained 
# vs. removed by any criterion 
table(vhd16sub$discard)
```

```{r plt-xy-qc, echo=FALSE, message=FALSE, fig.width=6, fig.height=2.5}
#| code-fold: true
plotSpots(vhd16sub, annotate="sum_log", point_size=0.2) + ggtitle("log-library size") +
plotSpots(vhd16sub, annotate="subsets_mt_percent", point_size=0.2) + ggtitle("% mitochondrial") +
plot_layout(nrow=1) & theme(
    legend.key.width=unit(0.5, "lines"),
    legend.key.height=unit(1, "lines")) &
    scale_color_gradientn(colors=pals::jet())
```

We can visualize the local outliers identified by SpotSweeper:
```{r qc-vis, message=FALSE, fig.width=7.5, fig.height=2.5}
#| code-fold: true
# table(vhd16sub$subsets_mt_percent_outliers) # NULL
plotSpots(vhd16sub, annotate="sum_outliers") + ggtitle("low_lib_size") +
plotSpots(vhd16sub, annotate="detected_outliers") + ggtitle("low_n_features") +
plotSpots(vhd16sub, annotate="discard") + ggtitle("discard") +
plot_layout(nrow=1, guides="collect") & theme(
    plot.title=element_text(hjust=0.5),
    legend.key.width=unit(0.5, "lines"),
    legend.key.height=unit(1, "lines")) &
    guides(col=guide_legend(override.aes=list(size=3))) & 
    scale_color_manual("discard", values=c("lavender", "purple"))
```

Subset the Visium HD 16 Âµm object to bins that passed QC:
```{r vhd16-qc-subset}
vhd16sub <- vhd16sub[, !vhd16sub$discard]
dim(vhd16sub)
```

## Clustering

First, we identify and subset to the highly/spatially variable genes in the target area:

```{r vhd16-lognorm-hvg}
vhd16sub <- logNormCounts(vhd16sub)
dec <- modelGeneVar(vhd16sub)
hvg <- getTopHVGs(dec, n=3000)

# restrict to HVGs; SVGs could also be used
.vhd16sub <- logNormCounts(vhd16sub[hvg, ])
```

As detailed in @sec-crs-clustering, `BANKSY` utilizes a pair of spatial kernels to capture gene expression variation, followed by dimension reduction and graph-based clustering to determine spatial domains.

```{r BANKSY, message=FALSE}
# use weak spatial information
l <- 0.2 
# first order neighbours in VisiumHD
k <- 8
a <- "logcounts"
xy <- c("array_row", "array_col")
# Run PCA on the BANKSY matrix
set.seed(2025)
.vhd16sub <- computeBanksy(.vhd16sub, assay_name=a, coord_names=xy, k_geom=k)
.vhd16sub <- runBanksyPCA(.vhd16sub, lambda=l, npcs=25, seed=2025)
# Run UMAP for marker gene visualization
set.seed(2025); .vhd16sub <- runUMAP(.vhd16sub)
reducedDim(vhd16sub, "UMAP") <- reducedDim(.vhd16sub, "UMAP")
# Perform BANKSY clustering
.vhd16sub<-clusterBanksy(.vhd16sub,resolution=1.2,lambda=l,npcs=25,seed=2025)
table(vhd16sub$BANKSY <- factor(.vhd16sub[[clusterNames(.vhd16sub)]]))
```

Next, we can perform differential expresison (DE) analysis and find markers for each cluster. Across all cells in each cluster, we obtain the average expression:
```{r deg}
# differential gene expression analysis
mgs <- findMarkers(vhd16sub, groups=vhd16sub$BANKSY, direction="up")
# select for a few markers per cluster
top <- lapply(mgs, \(df) rownames(df)[df$Top <= 3])
top <- unique(unlist(top))

# average expression by clusters
pbs <- aggregateAcrossCells(vhd16sub, 
    ids=vhd16sub$BANKSY, subset.row=top, 
    use.assay.type="logcounts", statistics="mean")
```

The marker genes in each cluster can be visualized with a heatmap:

```{r clus-gene-heatmap, message=FALSE, fig.width=10, fig.height=3.5}
#| code-fold: true
# visualize averages z-scaled across clusters
pheatmap(mat=t(assay(pbs)), scale="column", breaks=seq(-2, 2, length=101), 
         treeheight_row=5, treeheight_col=5)
```

The expression of marker genes in cluster 4, *MMP2* for instance, can be visualized in UMAP and spatially.
```{r spatial-marker-gene, warning=FALSE, message=FALSE, fig.width=6.5, fig.height=6.5}
#| code-fold: true
vhd16sub$log_MMP2 <- logcounts(vhd16sub)["MMP2", ]
(plotUMAP(vhd16sub, colour_by="BANKSY", point_size=0.8) +
  theme(legend.position="none") + plot_spacer() + 
  plotSpots(vhd16sub, annotate="BANKSY", point_size=0.8) +
  plot_layout(nrow=1, widths=c(1, 0.1, 1)) &
  scale_color_manual(name="BANKSY", values=unname(pals::watlington())) &
  theme(aspect.ratio=1, legend.key.size=unit(2/3, "lines"))) /
  (plotUMAP(vhd16sub, colour_by="log_MMP2", point_size=0.8) + 
     theme(legend.position="none") + plot_spacer() + 
  plotSpots(vhd16sub, annotate="log_MMP2", point_size=0.8) + 
    theme(plot.title=element_blank()) + 
    plot_layout(nrow=1, widths=c(1, 0.1, 1)) &
  scale_color_viridis_c(option="viridis", name="log_MMP2") &
  theme(aspect.ratio=1))
```

## Deconvolution

### Single cell reference

We will perform deconvolution on 16 Âµm binned Visium HD. First, we retrieve a matching (Chromium) scRNA-seq dataset, which includes low- (`Level1`) and high-resolution (`Level2`) annotations of cells into 9 and 31 subpopulations, respectively. We filter the reference population to patient 2 only, to ensure similar transcriptional profile across technologies.

```{r load-sce, results="hold", message=FALSE, warning=FALSE}
# retrieve dataset from OSF repository
id <- "Chromium_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)

# read into `SingleCellExperiment`
fs <- list.files(td, full.names=TRUE)
h5 <- grep("h5$", fs, value=TRUE)
sce <- read10xCounts(h5, col.names=TRUE)

# add cell metadata
csv <- grep("csv$", fs, value=TRUE)
cd <- read.csv(csv, row.names=1)
colData(sce)[names(cd)] <- cd[colnames(sce), ]

# use gene symbols as feature names
rownames(sce) <- make.unique(rowData(sce)$Symbol)

# exclude cells deemed to be of low-quality
sce <- sce[, sce$QCFilter == "Keep"]

# tabulate subpopulations by disease status
sce$Group <- gsub("P[0-9]", "", sce$Patient)
sce$Patient <- gsub("(P[0-9]).*", "\\1", sce$Patient)
sce <- sce[, sce$Patient == "P2"]
sce
```

::: {.callout-note title="Single cell reference annotation level mapping" collapse="true"}
We can visualize the mapping between the Level 1 and Level 2 annotation in the single cell reference data with a proportion table:

```{r sce-heatmap-level12, fig.width=6.5, fig.height=3.5}
#| code-fold: true
pheatmap(prop.table(table(sce$Level1, sce$Level2), 2),
         cellwidth=10, cellheight=10, treeheight_row=5, treeheight_col=5)
```

Note that `"Proliferating Immune II"` spans evenly between `"B cells"` and `"Tumor"`, so we will make it a separate class `"Prolif. Immune"` in Level 1 annotation.

```{r sce-level1addprolifimmune}
sce$Level1 <- ifelse(sce$Level2 == "Proliferating Immune II", "Prolif. Immune", sce$Level1)
table(sce$Level1) # tabulate low-res. labels
```

There are some differences in the provided deconvolution label (cell type 1 and 2) at 8 Âµm bin and the Level 2 single cell reference label:

```{r diff-lab1-level2}
setdiff(unique(vhd8$DeconvolutionLabel1), unique(sce$Level2))
setdiff(unique(sce$Level2), unique(vhd8$DeconvolutionLabel1))
```

```{r diff-lab2-level2}
setdiff(unique(vhd8$DeconvolutionLabel2), unique(sce$Level2))
setdiff(unique(sce$Level2), unique(vhd8$DeconvolutionLabel2))
```

These extra classes in Visium HD 8um annotation will be merged to `"Fibroblast"`, `"Myeloid"`, `"Prolif. Immune"`, "`B cells`", and `"Tumor"` as Level 1 annotation, respectively.
:::

### Merge annotation levels for 8 Âµm

We can merge the provided deconvolution annotation `"DeconvolutionLabel1"` (most likely cell type for singlets or first cell type for doublets) and `"DeconvolutionLabel2"` (the second cell type for doublets) into a lower resolution by the following grouping:

```{r sc-mapping-vhd8-lab1}
df <- data.frame(prop.table(table(sce$Level1, sce$Level2), 2))
colnames(df) <- c("Level1", "Level2", "Freq")
df <- df[df$Freq > 0.4, ]
df <- df[order(df$Level2),]
df <- df[df$Level2 != "Proliferating Immune II", names(df) != "Freq"]
df_map <- data.frame(rbind(apply(df, 2, as.character), 
                           c("Fibroblast", "Proliferating Fibroblast"),
                           c("Myeloid", "Proliferating Macrophages"),
                           c("Myeloid", "cDC I"),
                           c("Tumor", "Tumor IV"),
                           c("Fibroblast", "Vascular Fibroblast"),
                           c("B cells", "Memory B"),
                           c("Prolif. Immune", "NK"),
                           c("Prolif. Immune", "Proliferating Immune II")))

vhd8$deconlab1level1 <- df_map$Level1[match(vhd8$DeconvolutionLabel1, df_map$Level2)]
vhd8$deconlab2level1 <- df_map$Level1[match(vhd8$DeconvolutionLabel2, df_map$Level2)]
table(vhd8$deconlab1level1)
```

```{r sc-mapping-vhd8-lab2}
table(vhd8$deconlab2level1)
```

According to `RCTD`, approximately 75% of the 8 Âµm bins are singlets.

```{r vhd8-singlets}
print(round(table(vhd8$DeconvolutionClass)["singlet"]/ncol(vhd8), 2))
```

Let us check the top 4 common doublet pairs in the `"doublet_certain"` class:

```{r vhd8-doublets}
dl <- data.frame(colData(vhd8)[vhd8$DeconvolutionClass == "doublet_certain", 
                               c("deconlab1level1", "deconlab2level1")])

combo <- unlist(apply(dl, 1, sort))
combo_wide <- data.frame(deconlab1level1=combo[seq(1, length(combo), by=2)],
                         deconlab2level1=combo[seq(2, length(combo), by=2)])
rownames(combo_wide) <- rownames(dl)
head(sort(table(paste0(combo_wide$deconlab1level1,
                       "_", combo_wide$deconlab2level1)), 
          decreasing=TRUE), 4)
```

We can see that Myeloid are commonly co-localized with Fibroblast and Tumor cells as doublets.

::: {.callout-note title="Subset 8um bins in black box" collapse="true"}
We can perform the same subsetting procedure to 8 Âµm bins and compare the difference in number of bins in the zoomed (black box) area.

```{r vhd8-subset-zoom}
# rename column to be consistent with 16 Âµm
vhd8$deconlevel1 <- vhd8$deconlab1level1
vhd8sub <- get_subset_vhd(spe=vhd8, spe_range=vhd8r)
dim(vhd8sub)
# plotVisium(vhd8sub, point_size=0.9, zoom=TRUE) # sanity check
```

As expected, the number of bins in 8 Âµm is 4 times the number of bins in 16 Âµm. We will visualize on the same zoomed region for both resolutions in any downstream analysis.
:::

### `RCTD` ("doublet" mode) for 16 Âµm bins

We also assume there are at most two cell types in a 16 Âµm bin as in 8 Âµm, and thus we would expect less singlet proportions.

```{r rctd, message=FALSE, warning=FALSE, results="hide"}
xy <- data.frame(spatialCoords(vhd16sub))
.vhd16sub <- SpatialRNA(xy, counts(vhd16sub))
ids <- factor(sce$Level1)
names(ids) <- colnames(sce)
ref <- Reference(counts(sce), ids)
res <- create.RCTD(.vhd16sub, ref, max_cores=10)
res <- run.RCTD(res, doublet_mode="doublet")
```

Weights inferred by `RCTD` should be normalized such that proportions of cell types sum to 1 in each spot:

```{r dec-res, warning=FALSE, message=FALSE}
# normalize weights to obtain proportions
ws <- normalize_weights(res@results$weights)

# add proportion estimates as spot metadata
ws <- data.frame(as.matrix(ws))
colData(vhd16sub)[names(ws)] <- ws[colnames(vhd16sub), ]
```

Next, we can visualize the deconvolution proportion in space for each cell type:
```{r rctd-vis, message=FALSE, fig.width=12, fig.height=4}
#| code-fold: true
lapply(names(ws), \(.) 
    plotSpots(vhd16sub, annotate=., point_size=0.05)) |>
    wrap_plots(nrow=2) & theme(
    legend.key.width=unit(0.5, "lines"),
    legend.key.height=unit(1, "lines")) &
    scale_color_gradientn(colors=pals::jet())
```

Obtain the majority vote class for 16 Âµm bins. 
```{r rctd-majvote, message=FALSE}
# replace the last "." to " " in cell types 
colnames(ws) <- sub("\\.(?=[^.]*$)", " ", colnames(ws), perl=TRUE)

# derive majority vote label
ids <- names(ws)[apply(ws, 1, which.max)]
names(ids) <- rownames(ws)
table(vhd16sub$deconlevel1 <- factor(ids[colnames(vhd16sub)]))

# remove bins with NA label due to not passing internal 'RCTD' QC
vhd16sub <- vhd16sub[, !is.na(vhd16sub$deconlevel1)]
```

::: {.callout-note title="Majorited voted cell type frequency difference between 8 and 16 Âµm bins" collapse="true"}
We can also check the proportional frequency difference between 8 and 16 Âµm bins in the zoomed area. 
```{r dens-decon-ct-diff, fig.width=5, fig.height=2.5}
#| code-fold: true
n8 <- table(vhd8sub$deconlevel1)
n16 <- table(vhd16sub$deconlevel1)
plt_df <- data.frame(x=n8/n16)

ggplot(plt_df, aes(x=x.Freq)) +
  geom_density() + 
  scale_x_continuous(breaks=seq(0, max(plt_df$x.Freq), by=2)) + 
  xlab("Cell type frequency ratio") + 
  ggtitle("Distribution of cell type frequency ratio\nbetween 8 and 16 Âµm bins") + 
  theme_classic() +
  theme(plot.title=element_text(hjust=0.5))
```
Most cell types appear around 3-4 times more in the 8 compared to 16 Âµm bins, which is expected given the bin size difference. 
:::

### Deconvolution results at different resolutions

Visualize deconvolution result on 8 and 16 Âµm bins in the zoomed area.

```{r plt-vhd, warning=FALSE, fig.width=8, fig.height=3.6}
#| code-fold: true
plotVisium(vhd8sub, annotate="deconlevel1", zoom =TRUE, point_size=0.8) +
  ggtitle("8 Âµm") + plot_spacer() + 
  plotVisium(vhd16sub, 
             annotate="deconlevel1", zoom=TRUE, point_size=1.6) +
  ggtitle("16 Âµm") +
  plot_layout(nrow=1, guides="collect", widths=c(1, 0.1, 1)) &
  guides(col=guide_legend(override.aes=list(size=3))) &
  theme(legend.key.width=unit(0.5, "lines"),
        plot.title=element_text(hjust=0.5),
    legend.key.height=unit(1, "lines")) & 
  scale_fill_manual(values=unname(pals::trubetskoy())) & 
  facet_null()
```

## Concordance of clustering and deconvolution on 16 Âµm

We can visualize the results spatially:

```{r concordance-spatial, fig.width=8, fig.height=3.6}
#| code-fold: true
plotVisium(vhd16sub, annotate="BANKSY", zoom=TRUE, point_size=1.6,
           pal=unname(pals::watlington())) +
plotVisium(vhd16sub, annotate="deconlevel1", zoom=TRUE, point_size=1.6,
           pal=unname(pals::trubetskoy())) & facet_null()
```

Their concordance can be visualized with a heatmap:

```{r concordance-heatmap, fig.width=2.5, fig.height=3}
#| code-fold: true
fq <- prop.table(table(vhd16sub$BANKSY, vhd16sub$deconlevel1), 2)
pheatmap(fq, cellwidth=10, cellheight=10, treeheight_row=5, treeheight_col=5)
```

From the above heatmap, cluster 1 is mostly immune cells with endothelial; cluster 9 matches B cell bins; cluster 7 and 8 are likely intestinal epithelial and smooth muscle, respectively; fibroblast is common in cluster 4; tumor spans cluster 2, 5, and 6. 

## Neighbourhood analysis

Here, we want to quantify the abundance between *T cells*, *B cells*, or *Myeloid* bins to *Fibroblast* or *Tumor* bins in 16 Âµm zoomed region.
We use `getAbundances()` from `r BiocStyle::Biocpkg("Statial")` to calculate. 

For each bin, we compute the relative abundance of other cell types within a radius of 200 px;
the result is stored as `reducedDim` slot `"abundances"`.

```{r statial-vhd16sub}
xy <- data.frame(spatialCoords(vhd16sub))
colData(vhd16sub)[names(xy)] <- xy
vhd16sub <- getAbundances(vhd16sub, 
    cellType="deconlevel1", 
    imageID="sample_id", 
    spatialCoords=names(xy),
    r=200, nCores=4)
```

Let's have a look at the abundance and distance results, also including deconvolution results:

```{r statial-vhd16sub-abd}
abd <- reducedDim(vhd16sub, "abundances")
abd$CT <- vhd16sub$deconlevel1
abd[1:5, 1:5]
```

Now, we filter to only *T cells*, *B cells* and *Myeloid* bins, and investigate -- within their neighbourhood -- the spatial proximity differences to *Fibroblast* or *Tumor* bins in terms of abundance.

```{r statial-vhd16sub-plt-df}
abd_plt_df <- abd[
    abd$CT %in% c("T cells", "B cells", "Myeloid"), 
    c("Tumor", "Fibroblast", "CT")] |>
    pivot_longer(
        cols=c("Tumor", "Fibroblast"),
        names_to="targetCT", 
        values_to="Abundance")
```

In this region, we observe more tumor than fibroblast bins around each of the immune bins.
However, note that in this zoomed region, we have nearly 3 times more tumor bins compared to fibroblast bins. 
Thus, we need to scale abundance values by the number of target cell type bins observed in the region. 

```{r statial-vhd16sub-plt, fig.width=6, fig.height=2.5}
#| code-fold: true
abd_plt_df$Abundance_scaled <- 
  ifelse(abd_plt_df$targetCT == "Tumor",
         abd_plt_df$Abundance/table(vhd16sub$deconlevel1)["Tumor"],
         abd_plt_df$Abundance/table(vhd16sub$deconlevel1)["Fibroblast"])

ggplot(abd_plt_df, aes(x=CT, y=Abundance, fill=targetCT)) + 
ggplot(abd_plt_df, aes(x=CT, y=Abundance_scaled, fill=targetCT)) + 
plot_layout(nrow=1, guides="collect") & geom_boxplot() & 
    scale_fill_manual(values=c("dodgerblue", "tomato")) & 
    theme_classic() & theme(legend.key.size=unit(0, "lines")) &
    guides(fill=guide_legend(override.aes=list(shape=21, size=2)))
```

After correction, we see that the abundance of fibroblast bins does not differ much from the abundance of tumor bins around each immune cell type in the zoomed region. 

The same analysis steps could be applied to the (unfiltered) 8 ÎĽm resolution data with the corresponding `RCTD` results provided by the authors. 

## Conclusion

In this workflow, we demonstrated how to perform a standard analysis pipeline on a subset of Visium HD data binned at 16 Âµm.
Most methods and visualization techqniues developed for Visium can be applied to Visium HD.
However, more sparsity is expected, as each data point (e.g. 2, 8, or 16 Âµm) has smaller area compared to Visium (55Âµm).
Each bin cannot be interpreted as a cell, but rather, a cell-segmentation-free approach is applied here.
Below are some resources and news for Visium HD to consider and incorporate into your pipeline:

- GitHub repository of `bin2cell` for H&E segmentation and obtaining cell boundaries: https://github.com/Teichlab/bin2cell

- [Analysis pipeline](https://www.10xgenomics.com/analysis-guides/segmentation-visium-hd) by 10x Genomics (2024 April). The analysis consists of three steps: run `StarDist` nuclei segmentation in Python, sum gene expression into nuclei based on segmentation mask, and downstream analysis with `scanpy` and `anndata`. 

- 10x Genomics [news](https://www.prnewswire.com/news-releases/10x-genomics-unveils-innovation-roadmap-at-agbt-general-meeting-advancing-leadership-in-large-scale-single-cell-and-spatial-analysis-302382827.html) on morphology-driven segmentation from H&E enabled in CytAssist for Visium HD, by assigning transcripts to individual cells.

- [Pipeline](https://github.com/Sanofi-Public/enact-pipeline) on cell typing for Visium HD in Python by Sanofi.

## References {.unnumbered}
