# Spot deconvolution {#sec-seq-spot-deconvolution}

## Introduction

Spot-level ST data (e.g. from the 10x Genomics Visium platform) can contain zero, one, or multiple cells per spot, depending on the spatial resolution of the platform and the tissue cell density. This characteristic of the data affects several steps in analysis workflows, including quality control (@sec-seq-quality-control) and clustering (@sec-seq-clustering).

Note that this is also a characteristic of ST data that is distinct from single-cell RNA sequencing data, so here we cannot easily apply existing methods from single-cell workflows.

In this section, we will demonstrate methods to deconvolve cell types per spot.

## Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(RColorBrewer)
library(SpatialExperiment)
library(STexampleData)
library(ggspavis)

# devtools::install_github("dmcable/spacexr", build_vignettes = FALSE)
library(spacexr)

# BiocManager::install("STdeconvolve")
library(STdeconvolve)

resultpath <- "~/Downloads"


# basic theme for spatial plots
theme_xy <- list(
    coord_equal(expand=FALSE), 
    theme_void(), theme(
        plot.margin=margin(l=5),
        legend.key=element_blank(),
        panel.background=element_rect(fill="black")))
```


## Number of cells per spot

The following figure provides an overview of why deconvolution is needed for Visium data. The number of cells per spot is a known metadata stored in a column in `colData()` in the DLPFC human prefrontal cortex (TODO: cite DLPFC) `SpatialExperiment` of object. We use a visualization function from [ggspavis](https://bioconductor.org/packages/ggspavis) to generate the plot.

```{r, message=FALSE, results='hide'}
vis_dlpfc <- STexampleData::Visium_humanDLPFC()
head(colData(vis_dlpfc))
```

We see that spots in this human brain Visium dataset contain around 0-27 cells, with a median of 4. Therefore, it is plausible that some spots contain multiple cell types, and spot-level deconvolution could improve downstream analyses by deconvolving these cell types.

```{r, message=FALSE, warnings=FALSE, fig.width=5, fig.height=4}
# plot number of cells per spot
plotSpotQC(vis_dlpfc, plot_type = "histogram", x_metric = "cell_count") + 
  geom_vline(xintercept = median(na.omit(vis_dlpfc$cell_count)), lty = "dashed") +
  xlab("number of cells") + 
  ggtitle("Number of cells per spot")
```

In this example of consecutive slices of breast cancer data (TODO: Janesick), we perform Visium cell-type deconvolution with Chromium reference.


## Load previously saved data

We start by loading the previously saved data object(s).

```{r, message=FALSE, results='hide'}
vis_dlpfc <- STexampleData::Visium_humanDLPFC()
```

# RCTD

## Run RCTD

First, we create the reference single-cell matrix and the `spatialRNA` objects needed for running RCTD. We

```{r}
chrom <- readRDS(file.path())

# Reference Chromium, unnormalized counts
chrom_mat <- GetAssayData(chrom, slot = "counts")
cell_types <- as.factor(chrom$Annotation); names(cell_types) <- colnames(chrom)

ref <- Reference(chrom_mat, cell_types)

# Visium to be deconvolved, unnormalized counts
coords <- GetTissueCoordinates(vis)
counts <- GetAssayData(vis, slot = "counts")

puck <- SpatialRNA(coords, counts)
```

We specify "full" mode here to indicate there can be many cells in a spot, and save RCTD results.

```{r, warning=FALSE, eval = FALSE}
# DO NOT RUN - TAKING TOO LONG
myRCTD <- create.RCTD(puck, ref, max_cores = 4)
myRCTD <- run.RCTD(myRCTD, doublet_mode = 'full')
# saveRDS(myRCTD, "./intermediate_data/vis_RCTD.rds")
```

Organize cell-type deconvolution results, the weights should be normalized so the proportion of each cell type in a spot sums up to 1.

```{r}
myRCTD <- readRDS(file.path(resultpath, "vis_RCTD.rds"))
results <- myRCTD@results
norm_weights <- normalize_weights(results$weights)
cell_type_names <- myRCTD@cell_type_info$info[[2]]
spatialRNA <- myRCTD@spatialRNA
barcodes <- colnames(myRCTD@spatialRNA@counts)
```

## Plot cell-type deconvolution

Plot cell-type deconvolution result. We first check the deconvolution result for one cell-type. Here we set the intensity range based on the probability score range of this cell-type across all spots.

```{r, message=FALSE, fig.width=5, fig.height=4.5}

```

10X Genomics implemented a speed-up verison of RCTD in a pull request. It can be installed via TODO. This version has been in the preprint TODO: Oliveria, 2024. 


# STdeconvolve
Another deconvolution tool is 

## Run STdeconvolve
```{r}

```

Other available tools include: 

- R tools:
  - [CARD](https://github.com/YMa-lab/CARD), 
- Python tools:
  - Cell2Location


# Appendix

## Saving {.unnumbered}

## Benchmarks {.unnumbered}

http://dx.doi.org/10.1093/bioinformatics/btz363
http://dx.doi.org/10.1038/s41467-020-19015-1


## References {.unnumbered}
