# Workflow: Visium HD {#sec-seq-workflow-visium-hd}

## Introduction

Visium HD is a next-generation spatial transcriptomics technology developed by 10x Genomics, designed to provide full transcriptome spatial gene expression data across entire tissue sections. Commercially available since 2024, Visium HD advanced the platform's resolution from 55 µm spots to bins of subcellular resolution (2 µm). In this workflow, we will demonstrate common analysis steps with a Visium HD dataset on colorectal cancer [@Oliveira2024].

## Dependencies {#sec-seq-workflow-visium-hd-load-data}

```{r deps, message=FALSE, warning=FALSE}
library(Banksy)
library(BiocParallel)
library(dplyr)
library(DropletUtils)
library(ggplot2)
library(ggspavis)
library(igraph)
library(magick)
library(OSTA.data)
library(patchwork)
library(pheatmap)
library(scater)
library(scran)
library(scuttle)
library(sf)
library(spacexr)
library(SpatialExperiment)
library(SpotSweeper)
library(Statial)
library(tidyr)
library(tidyverse)
library(VisiumIO)
```

In this demo, we perform deconvolution on 16 µm bins and compare the concordance with results on 8 µm bins, provided by 10x Genomics. [Here, we only demonstrate the analysis for patient 2 (P2). Other patients, such as P1 and P5, also have public Chromium, Visium HD, and Xenium data available for the CRC group (see @sec-bkg-datasets). It would be interesting to investigate the differences in malignant cells/bins (e.g., DEGs, pathways) between patients across different platforms.]{.aside}

```{r load-vhd8, message=FALSE, warning=FALSE}
# retrieve dataset from OSF repo
id <- "VisiumHD_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# read 8um bins into 'SpatialExperiment'
vhd8 <- TENxVisiumHD(
    spacerangerOut=td, 
    processing="filtered",
    format="h5", 
    images="lowres", 
    bin_size="008") |>
    import()
# use gene symbols as feature names
gs <- rowData(vhd8)$Symbol
rownames(vhd8) <- make.unique(gs)
colnames(vhd8) <- vhd8$barcode # TODO: VisiumIO bug?
vhd8
```

These data come with bin-level annotations from deconvolution estimates by `RCTD`, implemented in `r BiocStyle::Biocpkg("spacexr")`. Specifically, two sets of labels are available that correspond to the most and second most frequent of 38 cell types per bin, respectively:

```{r load-anno, message=FALSE, warning=FALSE}
# retrieve annotations
gz <- "binned_outputs/square_008um/deconvolution.csv.gz"
df <- read.csv(file.path(td, gz), row.names=2)
head(df <- df[complete.cases(df), -1])
# keep only annotated bins
vhd8 <- vhd8[, rownames(df)]
names(df) <- c("DeconClass", "DeconLabel1", "DeconLabel2")
colData(vhd8)[names(df)] <- df
```

```{r plt-lab-xy, fig.width=12, fig.height=5, message=FALSE}
#| code-fold: true
lab <- grep("Label", names(cd <- colData(vhd8)), value=TRUE)
pal <- hcl.colors(length(unique(unlist(cd[lab]))), "Spectral")
lapply(lab, \(.) {
  plotCoords(vhd8, annotate=., point_size=0.2, point_shape = 15) + ggtitle(.)
}) |>
  wrap_plots(nrow=1, guides="collect") &
  scale_color_manual(NULL, values=pal) &
  theme(legend.key.size=unit(0, "lines"))
```

Another resolution of VisiumHD with 16 µm bin can be read in as follow:

```{r load-vhd16, message=FALSE}
# read 16um bins into 'SpatialExperiment'
vhd16 <- TENxVisiumHD(
    spacerangerOut=td,
    processing="filtered",
    format="h5",
    images="lowres",
    bin_size="016") |>
    import()
# use symbols as feature names
gs <- rowData(vhd16)$Symbol
rownames(vhd16) <- make.unique(gs)
colnames(vhd16) <- vhd16$barcode # TODO: VisiumIO bug?
vhd16
```

::: {.callout-note title="Remove bins overlaying empty tissue" collapse="true"}
Some global filtering based on library size is necessary to remove bins that overlay empty tissue.

```{r vhd16-libsize}
vhd16$libsize <- colSums(counts(vhd16))
```

We can visualize such bins by zooming into an empty tissue region. On the left-hand side, one region shows a clear gap with low counts. Informed by H&E staining, bins located above tissue gaps should not be interpreted biologically, as the low counts in these bins are likely due to ambient RNA or transcript spillover. Therefore, they should be excluded from downstream analysis. 

Based on a [discussion](https://github.com/10XGenomics/HumanColonCancer_VisiumHD/issues/28) with the original author at 10x Genomics, in this dataset, 8 µm bins with a library size below 100 are removed. Since we expect a 4-fold increase in library size across all bins at 16 µm, we can set the filtering threshold to 400. 

```{r vhd16-empvis, fig.width=8, fig.height=2.5, message=FALSE, warning=FALSE}
#| code-fold: true
p1 <- plotVisium(vhd16, annotate="libsize", point_shape = 22,
           zoom = TRUE, show_axes = TRUE, point_size = 1.3) + 
    xlim(c(422, 442)) + ylim(c(318, 330)) + ggtitle("16 µm - original")

vhd16 <- vhd16[, vhd16$libsize > 400]

p2 <- plotVisium(vhd16, annotate="libsize", point_shape = 22,
           zoom = TRUE, show_axes = TRUE, point_size = 1.3) + 
    xlim(c(422, 442)) + ylim(c(318, 330)) + ggtitle("16 µm - post > 400 UMI QC")

p1 | p2
```

:::

Clustering results generated by Space Ranger are provided for both resolutions.

```{r load-clus, message=FALSE, warning=FALSE}
csv <- list.files(td, "clustering", recursive=TRUE, full.names=TRUE)
dfs <- lapply(csv, read.csv, row.names="barcode") 

colData(vhd8)$cluster <- factor(dfs[[1]][colnames(vhd8), "cluster"])
colData(vhd16)$cluster <- factor(dfs[[2]][colnames(vhd16), "cluster"])
```

We can visualize the given clustering results at 8 µm and 16 µm resolutions. Later, we will compare the clustering result at 16 µm with that from Banksy.

```{r clusvis, fig.width=12, fig.height=6, message=FALSE}
#| code-fold: true
(plotCoords(vhd8, annotate="cluster", point_size=0.05, point_shape = 15,
            pal=unname(pals::kelly())) + ggtitle("8 µm")) |
(plotCoords(vhd16[, !is.na(vhd16$cluster)], annotate="cluster", point_size=0.1, 
            point_shape = 15, pal=unname(pals::kelly())) + ggtitle("16 µm")) 
```

To keep runtimes low, our downstream analysis will be performed on a subset area of 16 µm bins (1/64 of the data coverage area), which we read in here; there are no annotations available for this resolution in the public domain. [Note that 8 and 16 µm binned Visium HD data share the same full-resolution H&E image and scaling factor.]{.aside}

```{r plt-xy-box, fig.width=12, fig.height=3.5, message=FALSE, warning=FALSE}
#| code-fold: true
.rng <- \(spe) {
    xy <- spatialCoords(spe)*scaleFactors(spe)
    xs <- range(xy[, 1]); ys <- nrow(imgRaster(spe))-range(xy[, 2])
    x1 <- xs[1] + 4*(xs[2]-xs[1])/8; x2 <- xs[2] - 3*(xs[2]-xs[1])/8
    y1 <- ys[1] + 3*(ys[2]-ys[1])/8; y2 <- ys[2] - 4*(ys[2]-ys[1])/8
    list(box=c(x1, x2, y1, y2), cov=c(xs, ys))
}
vhd8r <- .rng(vhd8)
# vhd16r <- .rng(vhd16) 
# use 8um bounding boxes to subset spes at both resolutions; 
# it is similar to 16um's box range 
.box <- \(roi, lty, lwd, col) {
    geom_rect(
        xmin=vhd8r[[roi]][1], xmax=vhd8r[[roi]][2], 
        ymin=vhd8r[[roi]][3], ymax=vhd8r[[roi]][4],
        col=col, fill=NA, linetype=lty, linewidth=lwd)
}
cov <- .box(roi="cov", lty=2, lwd=1/2, col="grey")
box <- .box(roi="box", lty=4, lwd=2/3, col="black")
# plotting
.lim <- \(spe) list(
    xlim(spe[["box"]][c(1, 2)]),
    ylim(spe[["box"]][c(4, 3)]))

plotVisium(vhd8, spots=FALSE, point_shape=22) + cov + box +
    ggtitle("grey: data coverage\n black: zoomed region") + 
plotVisium(vhd8, point_size=0.8, zoom=TRUE, point_shape=22) + .lim(vhd8r) +
    ggtitle("8 µm bins in black box") + 
plotVisium(vhd16, point_size=1.6, zoom=TRUE, point_shape=22) + .lim(vhd8r) +
    ggtitle("16 µm bins in black box") + 
plot_layout(nrow=1, widths=c(1.5, 1, 1)) & facet_null() &
    theme(plot.title=element_text(hjust=0.5, vjust=0.5))
```

Subset the 16 µm object to bins in the black box:

```{r vhd16-sub, message=FALSE}
#| code-fold: true
.sub <- function(spe, rng, roi="box") {
    xs <- rng[[roi]][c(1, 2)]
    ys <- nrow(imgRaster(spe)) - rng[[roi]][c(3,4)]
    xy <- spatialCoords(spe)*scaleFactors(spe)
    spe[, xy[, 1] > xs[1] & xy[, 1] < xs[2] & 
          xy[, 2] > ys[1] & xy[, 2] < ys[2] ]
}
.vhd16 <- .sub(spe=vhd16, rng=vhd8r)
dim(.vhd16)
```

From now on, we are operating on `r ncol(.vhd16)` 16 µm bins.

## Quality control

As detailed in @sec-seq-quality-control, we use `r BiocStyle::Biocpkg("SpotSweeper")` to perform quality control on the subsetted 16 µm data.

```{r qc-vhd16-local}
# calculate per-cell QC metrics
mt <- grepl("^MT-", rownames(.vhd16))
.vhd16 <- addPerCellQCMetrics(.vhd16, subsets=list(mt=mt))
# determine outliers based on 
# - low log-library size
# - few uniquely detected features
# - high mitochondrial count fraction
.vhd16 <- localOutliers(.vhd16, metric="sum", direction="lower", log=TRUE)
.vhd16 <- localOutliers(.vhd16, metric="detected", direction="lower", log=TRUE)
.vhd16 <- localOutliers(.vhd16, metric="subsets_mt_percent", direction="higher", log=TRUE)
.vhd16$discard <- 
    .vhd16$sum_outliers | 
    .vhd16$detected_outliers | 
    .vhd16$subsets_mt_percent_outliers
# tabulate number of bins retained 
# vs. removed by any criterion 
table(.vhd16$discard)
```

```{r plt-xy-qc, echo=FALSE, message=FALSE, fig.width=6, fig.height=2.5}
#| code-fold: true
plotCoords(.vhd16, annotate="sum_log", point_size=0.2, point_shape=15) + 
  ggtitle("log-library size") +
  plotCoords(.vhd16, annotate="subsets_mt_percent", point_size=0.2, point_shape=15) + 
  ggtitle("% mitochondrial") +
  plot_layout(nrow=1) & theme(
    legend.key.width=unit(0.5, "lines"),
    legend.key.height=unit(1, "lines")) &
  scale_color_gradientn(colors=pals::jet())
```

We can visualize the local outliers identified by `SpotSweeper` in space:

```{r qc-vis, fig.width=7.5, fig.height=2.5, message=FALSE}
#| code-fold: true
plotCoords(.vhd16, annotate="sum_outliers", point_shape=15) + 
  ggtitle("low_lib_size") +
  plotCoords(.vhd16, annotate="detected_outliers", point_shape=15) + 
  ggtitle("low_n_features") +
  plotCoords(.vhd16, annotate="discard", point_shape=15) + 
  ggtitle("discard") +
  plot_layout(nrow=1, guides="collect") & theme(
    plot.title=element_text(hjust=0.5),
    legend.key.size=unit(0, "lines")) &
  guides(col=guide_legend(override.aes=list(size=3))) & 
  scale_color_manual("discard", values=c("lavender", "purple"))
```

Lastly, we subset the Visium HD 16 µm object to bins that passed QC:

```{r vhd16-qc-subset}
.vhd16 <- .vhd16[, !.vhd16$discard]
dim(.vhd16)
```

## Clustering

First, we identify highly variable genes (HVGs) in the target area: [Note that spatially variable genes (SVGs) could be used instead; c.f. @sec-seq-processing.]{.aside}

```{r vhd16-hvg}
.vhd16 <- logNormCounts(.vhd16)
dec <- modelGeneVar(.vhd16)
hvg <- getTopHVGs(dec, n=3e3)
```

As detailed in @sec-crs-clustering, `r BiocStyle::Biocpkg("Banksy")` utilizes a pair of spatial kernels to capture gene expression variation, followed by dimension reduction and graph-based clustering to identify spatial domains.

```{r Banksy, message=FALSE}
# set seed for random number generation
# in order to make results reproducible
set.seed(112358)
# 'Banksy' parameter settings
k <- 8   # consider first order neighbors
l <- 0.2 # use little spatial information
a <- "logcounts"
xy <- c("array_row", "array_col")
# restrict to selected features
tmp <- .vhd16[hvg, ]
# compute spatially aware 'Banksy' PCs
tmp <- computeBanksy(tmp, assay_name=a, coord_names=xy, k_geom=k)
tmp <- runBanksyPCA(tmp, lambda=l, npcs=20)
reducedDim(.vhd16, "PCA") <- reducedDim(tmp)
## run UMAP (for visualization purposes only)
# .vhd16 <- runUMAP(.vhd16, dimred="PCA")
# build cellular shared nearest-neighbor (SNN) graph
g <- buildSNNGraph(.vhd16, use.dimred="PCA", type="jaccard", k=20)
# cluster using Leiden community detection algorithm
k <- cluster_leiden(g, objective_function="modularity", resolution=1.2)
table(.vhd16$Banksy <- factor(k$membership))
```

Next, we can perform differential gene expresison (DGE) analysis to identify markers for each cluster; c.f. @sec-crs-signatures. We also compute the cluster-wise mean expression of selected markers:

```{r mgs}
# differential gene expression analysis
mgs <- findMarkers(.vhd16, groups=.vhd16$Banksy, direction="up")
# select for a few markers per cluster
top <- lapply(mgs, \(df) rownames(df)[df$Top <= 2])
top <- unique(unlist(top))
# average expression by clusters
pbs <- aggregateAcrossCells(.vhd16, 
    ids=.vhd16$Banksy, subset.row=top, 
    use.assay.type="logcounts", statistics="mean")
```

The marker genes in each cluster can be visualized with a heatmap:

```{r plt-mgs-hm, message=FALSE, fig.width=8, fig.height=3}
#| code-fold: true
# visualize averages z-scaled across clusters
pheatmap(
    mat=t(assay(pbs)), scale="column", breaks=seq(-2, 2, length=101), 
    cellwidth=10, cellheight=10, treeheight_row=5, treeheight_col=5)
```

Or, we can visualize the bin-wise expression of selected markers in space:

```{r plt-mgs-xy, fig.width=10, fig.height=3, message=FALSE}
#| code-fold: true
gs <- c("MMP2", "PIGR", "IGHG1")
ps <- lapply(gs, \(.) plotCoords(.vhd16, annotate=., point_shape=15, 
                                 point_size=0.8, assay_name="logcounts"))
wrap_plots(ps, nrow=1) & theme(
  legend.key.width=unit(0.5, "lines"),
  legend.key.height=unit(1, "lines")) &
  scale_color_gradientn(colors=rev(hcl.colors(9, "Rocket")))
```

## Deconvolution

### Load single-cell reference

We will now perform deconvolution on the 16 µm binned Visium HD data. First, we retrieve matching (Chromium) scRNA-seq data, which includes low- (`Level1`) and high-resolution (`Level2`) annotations of cells into 9 respectively 31 clusters. In order to ensure similar transcriptional profile across technologies, we filter the reference population to patient 2 only.

```{r load-sce, results="hold", message=FALSE, warning=FALSE}
# retrieve dataset from OSF repository
id <- "Chromium_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# read into `SingleCellExperiment`
fs <- list.files(td, full.names=TRUE)
h5 <- grep("h5$", fs, value=TRUE)
sce <- read10xCounts(h5, col.names=TRUE)
# add cell metadata
csv <- grep("csv$", fs, value=TRUE)
cd <- read.csv(csv, row.names=1)
colData(sce)[names(cd)] <- cd[colnames(sce), ]
# use gene symbols as feature names
gs <- rowData(sce)$Symbol
rownames(sce) <- make.unique(gs)
# exclude cells deemed to be of low-quality
sce <- sce[, sce$QCFilter == "Keep"]
# subset cells from same patient
sce <- sce[, grepl("P2", sce$Patient)]
sce
```

::: {.callout-note title="Single-cell reference annotation mapping" collapse="true"}
We can visualize the mapping between low- and high-resolution annotations in the single-cell reference data with a contingency table:

```{r sce-heatmap-level12, fig.width=6.5, fig.height=3.5}
#| code-fold: true
fq <- prop.table(table(sce$Level1, sce$Level2), 2)
pheatmap(fq, cellwidth=10, cellheight=10, treeheight_row=5, treeheight_col=5)
```

Note that `"Proliferating Immune II"` spans evenly between `"B cells"` and `"Tumor"`, so we will make it a separate class, `"Prolif. Immune"` in the low-resolution annotations:

```{r sce-level1addprolifimmune}
sce$Level1 <- ifelse(grepl("Prolif", sce$Level2), "Prolif. Immune", sce$Level1)
table(sce$Level1) # tabulate low-res. labels
```

There are some differences in the provided deconvolution labels at 8 µm resolution and the high-resolution single-cell reference labels:

```{r diff-labs-lvl2}
# all single-cell reference labels 
# are present in the Visium HD data
setdiff(sce$Level2, vhd8$DeconLabel1)
setdiff(sce$Level2, vhd8$DeconLabel2)
# but not vice versa
setdiff(vhd8$DeconLabel1, sce$Level2)
setdiff(vhd8$DeconLabel2, sce$Level2)
```

Below, these extra classes in the Visium HD 8 μm bin annotations will be ignored.
:::

### Simplify 8 µm annotations

We can merge the provided deconvolution annotation `"DeconLabel1"` (most likely cell type for singlets, most dominant type for doublets) into a lower resolution using the single-cell reference annotations:

```{r vhd8-lv1}
i <- match(vhd8$DeconLabel1, sce$Level2)
j <- match(vhd8$DeconLabel2, sce$Level2)
vhd8$.DeconLabel1 <- sce$Level1[i]
vhd8$.DeconLabel2 <- sce$Level1[j]
vhd8 <- vhd8[, !is.na(vhd8$.DeconLabel1)]
table(vhd8$.DeconLabel1)
```

According to `RCTD`, approximately 75% of the 8 µm bins are singlets:

```{r vhd8-singlets}
round(100*mean(vhd8$DeconClass == "singlet"), 2)
```

Let us check the top-5 common doublet pairs according to the `"doublet_certain"` class:

```{r vhd8-doublets}
dbl <- vhd8$DeconClass == "doublet_certain"
lab <- c(".DeconLabel1", ".DeconLabel2")
df <- data.frame(colData(vhd8)[dbl, lab])
# sort as to ignore order
df <- apply(df, 1, sort)
df <- do.call(rbind, df)
names(df) <- lab
# count unique pairs
ij <- paste(df[, 1], df[, 2], sep=";")
head(sort(table(ij), decreasing=TRUE), 5)
```

We can see that myeloid are commonly co-localized with tumor and fibroblasts cells; most doublets are homotypic, i.e., they are (mostly) composed of one low-resolution type.

::: {.callout-note title="Subset 8um bins within black bounding box" collapse="true"}
We can perform the same subsetting procedure to 8 µm bins and compare the difference in the number of bins in the zoomed (black box) area.

```{r vhd8-subset-zoom}
# subset zoomed (black box) area
.vhd8 <- .sub(spe=vhd8, rng=vhd8r)
# compare number of bins between resolutions
(n <- ncol(.vhd8))
(m <- ncol(.vhd16))
round(n/m, 2)
```

As expected, the number of 8 µm bins is about 4 times the number of 16 µm bins. We will visualize the same subset for both bin sizes in any downstream analyses.
:::

### Run `RCTD` on 16 µm bins

We also assume there are at most two cell types in a 16 µm bin as in 8 µm, and thus we would expect a smaller proportion of singlets.

```{r RCTD, message=FALSE, warning=FALSE, results="hide"}
# downsample to at most 4,000 cells per cluster for sce
# (this is done only to keep runtime/memory low)
cs <- split(seq_len(ncol(sce)), sce$Level1)
cs <- lapply(cs, \(.) sample(., min(length(.), 4e3)))
ncol(.sce <- sce[, unlist(cs)])
rctd_data <- createRctd(.vhd16, .sce, cell_type_col="Level1")
(res <- runRctd(rctd_data, max_cores=4, rctd_mode="doublet"))
```

Weights inferred by `RCTD` correspond to proportions of cell types, such that they sum to 1 for a given observation (expect for rejected ones, which sum to 0):

```{r dec-res}
# counts rejected observations
ws <- assay(res, "weights")
table(colSums(ws) == 0)
# add proportion estimates as metadata
ws <- data.frame(t(as.matrix(ws)))
colData(.vhd16)[names(ws)] <- ws[colnames(.vhd16), ]
```

Next, we can spatially visualize the deconvolution proportions estimates:

```{r plt-dec-est, message=FALSE, fig.width=9, fig.height=4}
#| code-fold: true
lapply(names(ws), \(.) 
    plotCoords(.vhd16, annotate=., point_size=0.3, point_shape=15)) |>
    wrap_plots(nrow=2, guides="collect") & theme(
    legend.key.width=unit(0.5, "lines"),
    legend.key.height=unit(1, "lines")) &
    scale_color_gradientn(colors=rev(hcl.colors(9, "Rocket")))
```

We may also obtain the majority vote class for 16 µm bins: [This type of analysis procedure is underlying the annotations provided by the authors.]{.aside}

```{r dec-lab, message=FALSE}
# derive majority vote labels
ids <- names(ws)[apply(ws, 1, which.max)]
ids <- gsub("\\.([A-z])", " \\1", ids)
idx <- match(colnames(.vhd16), rownames(ws))
table(.vhd16$.DeconLabel1 <- factor(ids[idx]))
```

::: {.callout-note title="Majority voted cell type frequency difference between 8 and 16 µm bins" collapse="true"}
We can also check the proportional frequency difference between 8 and 16 µm bins in our region of interest:

```{r dec-816, fig.width=5, fig.height=2.5}
#| code-fold: true
n8 <- table(.vhd8$.DeconLabel1)
n16 <- table(.vhd16$.DeconLabel1)
df <- data.frame(x=as.numeric(n8/n16))
ggplot(df, aes(x)) + geom_density() + 
    xlab("Cell type frequency ratio") + 
    ggtitle("distribution of cell type frequency\nratio between 8 vs. 16 µm bins") + 
    scale_x_continuous(breaks=seq(0, max(df$x), by=2)) + 
    theme_classic() + theme(plot.title=element_text(hjust=0.5))
```

Most cell types appear around 3-4 times more in 8 compared to 16 µm bins, which is expected given the difference in bin size.
:::

### Deconvolution results at different resolutions

Let's visualize deconvolution result for 8 and 16 µm bins in the zoomed area:

```{r plt-dec-lab, fig.width=8, fig.height=3, message=FALSE}
#| code-fold: true
plotVisium(.vhd8, annotate=".DeconLabel1", zoom =TRUE, point_size=0.8, point_shape=22) + 
  ggtitle("8 µm") + plot_spacer() +
plotVisium(.vhd16, annotate=".DeconLabel1", zoom=TRUE, point_size=1.6, point_shape=22) + 
  ggtitle("16 µm") +
plot_layout(nrow=1, guides="collect", widths=c(1, 0.05, 1)) &
    guides(col=guide_legend(override.aes=list(size=2))) &
    scale_fill_manual(values=unname(pals::trubetskoy())) & 
    facet_null() & theme(
        plot.title=element_text(hjust=0.5),
        legend.key.size=unit(0, "lines")) 
```

We observe clear agreement in deconvolution at both 8 µm and 16 µm resolutions; however, the 8 µm resolution better captures fine structures, such as the string-like shape of endothelial and fibroblast bins extending into the upper left tumor region.

## Downstream

### Concordance of clustering & deconvolution

We can visualize the results spatially:

```{r plt-dec-clu-xy, fig.width=12, fig.height=3}
#| code-fold: true
plotVisium(.vhd16, annotate="cluster", zoom=TRUE, point_shape=22,
    point_size=1.6, pal=unname(pals::kelly())) + 
    plot_spacer() +
plotVisium(.vhd16, annotate="Banksy", zoom=TRUE, point_shape=22,
    point_size=1.6, pal=unname(pals::kelly())) + 
    plot_spacer() +
plotVisium(.vhd16, annotate=".DeconLabel1", zoom=TRUE, point_shape=22,
    point_size=1.6, pal=unname(pals::trubetskoy())) +
plot_layout(nrow=1, 
    widths=c(1, 0.05, 1, 0.05, 1)) & facet_null() &
    theme(legend.key.size=unit(0, "lines"))
```

We observe that Banksy delineates tissue borders very well, while deconvolution provides direct biological insight into the underlying clusters. Their concordance can be visualized using a heatmap:

```{r plt-dec-clu-hm, fig.width=2.5, fig.height=3.5}
#| code-fold: true
fq <- prop.table(table(.vhd16$Banksy, .vhd16$.DeconLabel1), 1)
pheatmap(fq, cellwidth=10, cellheight=10, treeheight_row=5, treeheight_col=5)
```

From the above heatmap, cluster 5 and 7 are mostly immune cells; cluster 9 matches myeloid bins; cluster 13 and 1 are likely intestinal epithelial and smooth muscle, respectively; fibroblast is common in cluster 10; tumor spans cluster 3, 11, and 12.

## Neighborhood analysis

Here, we want to quantify the abundance between *T cells*, *B cells*, or *Myeloid* bins to *Fibroblast* or *Tumor* bins in 8 µm zoomed region. We use `getAbundances()` from `r BiocStyle::Biocpkg("Statial")` to calculate, for each bin, the abundance of other cell types within a radius of 200 px; the resulting (bins $\times$ types) matrix is stored as `reducedDim` slot `"abundances"`.

```{r Statial}
.vhd8 <- .vhd8[, !is.na(.vhd8$.DeconLabel1)]
xy <- data.frame(spatialCoords(.vhd8))
colData(.vhd8)[names(xy)] <- xy
.vhd8 <- getAbundances(.vhd8, 
    cellType=".DeconLabel1", 
    imageID="sample_id", 
    spatialCoords=names(xy),
    r=200, nCores=4)
```

Let's have a look at the abundance and distance results, also including deconvolution results:

```{r Statial-df}
df <- reducedDim(.vhd8, "abundances")
df$CT <- .vhd8$.DeconLabel1
df[1:5, 1:5]
```

Now, we filter to only *T cells*, *B cells* and *Myeloid* bins, and investigate -- within their neighborhood -- the spatial proximity differences to *Fibroblast* or *Tumor* bins in terms of abundance.

```{r Statial-fd}
source <- df$CT %in% c("T cells", "B cells", "Myeloid")
target <- c("Tumor", "Fibroblast", "CT")
fd <- pivot_longer(
    df[source, target], cols=-CT,
    names_to="target", values_to="n")
head(fd)
```

In this region, we observe more fibroblast than tumor bins around each of the immune bins. However, note that in this zoomed region, we have nearly 1.4 times more fibroblast bins compared to tumor bins. Thus, we need to scale abundance values by the number of target cell type bins observed in the region.

```{r Statial-plt, fig.width=6, fig.height=2.5}
#| code-fold: true
n_tum <- table(.vhd8$.DeconLabel1)["Tumor"]
n_fib <- table(.vhd8$.DeconLabel1)["Fibroblast"]
fd$p <- ifelse(fd$target == "Tumor", fd$n/n_tum, fd$n/n_fib)
ggplot(fd, aes(x=CT, y=n, fill=target)) + 
    labs(y="# cells within 200px") +
ggplot(fd, aes(x=CT, y=p, fill=target)) + 
    labs(y="relative abundance") +
plot_layout(nrow=1, guides="collect") & 
    geom_boxplot(key_glyph="point") & 
    scale_fill_manual(values=c("yellow", "pink")) & 
    guides(fill=guide_legend(override.aes=list(shape=21, size=2))) &
    theme_classic() & theme(
        axis.title.x=element_blank(),
        legend.key.size=unit(0, "lines"))
```

After correction, we see amplified differences in the relative abundance of fibroblast bins compared to tumor bins around each immune cell type in the zoomed region.

We can also investigate the "cell-cell-marker" relationship based on the minimum distance between two cell types within a given radius. The matrix of minimum distance between all pairwise cell types is stored in the reduced dimension, similar to how the abundance matrix was previously stored. 
```{r vhd8-getdis}
.vhd8 <- getDistances(.vhd8, cellType=".DeconLabel1", imageID="sample_id",
                      spatialCoords=names(xy), maxDist = 200, nCores=4)
```

Here, we aim to determine whether the expression level of the tumor marker changes in proximity to fibroblast. Using FTH1 as an example, we observe a decrease in FTH1 expression in tumor bins as the distance to fibroblast increase. This trend could be visualized using a scatter plot. 
```{r vhd8-ccm, fig.width=11, fig.height=5, message=FALSE, warning=FALSE}
#| code-fold: true
p <- plotStateChanges(cells = .vhd8, type = "distances", imageID = "sample_id",
                      image = "sample01", cellType = ".DeconLabel1",
                      from = "Tumor", to = "Fibroblast", marker = "FTH1",
                      spatialCoords = c("array_col", "array_row"),
                      size = 1, shape = 19, interactive = FALSE,
                      plotModelFit = FALSE, method = "lm")

p$image + facet_null() | p$scatter + facet_null()
```

The same analysis steps could be applied to the (unfiltered) 16 μm resolution data with the corresponding deconvolution results.

## Cell-level analysis

-   10x Genomics [news](https://www.prnewswire.com/news-releases/10x-genomics-unveils-innovation-roadmap-at-agbt-general-meeting-advancing-leadership-in-large-scale-single-cell-and-spatial-analysis-302382827.html) on morphology-driven segmentation from H&E enabled in CytAssist for Visium HD, by assigning transcripts to individual cells. 
    - Space Ranger v4 now provides a segmented, cell-level count matrix under `segmented_outputs`, in parallel with the previous `binned_outputs` at 2, 8, and 16 µm, simplifying additional cell segmentation using `bin2cell` or `StarDist`. However, a comparative study of segmentation methods on Visium HD data has yet to be conducted. 

```{r load-other-vhdcell, message=FALSE, warning=FALSE}
# Read the GeoJSON file of cell segmentation
geo_data <- st_read("~/Desktop/cell_segmentations.geojson")
scalef <- rjson::fromJSON(file = "~/Desktop/scalefactors_json.json")
image <- image_read(file.path("~/Desktop/tissue_hires_image.png"))
info <- image_info(image)
```
First, we need to scale the coordinates to match the high resolution image. In addition, the y-axis coordinates should be flipped, as R uses a different origin than the image coordinate system. While `ggspavis` manages this internally for Visium, we will proceed with a manual plotting approach for Visium HD polygons (rather than cell centroid x and y) to maintain clarity.
```{r scale-flip-polygons, message=FALSE, warning=FALSE}
# Multiply all coordinates by the scaling factor of high resolution image
st_geometry(geo_data) <- st_geometry(geo_data) * scalef$tissue_hires_scalef

# Flip axis by flip y coordinates with image height  - y axis
st_geometry(geo_data) <- st_sfc( 
    lapply(st_geometry(geo_data), function(geom) {
      coords <- st_coordinates(geom)
      coords[, 2] <- info$height - coords[, 2]
      st_polygon(list(matrix(coords[, 1:2], ncol = 2)))
    }),
    crs = st_crs(geo_data)
)
```

We subset to the exact same region as before, yet with high resolution image for better visualization quality. 
```{r subset-polygons, message=FALSE, warning=FALSE}
# Get data coverage and subsetted bounding box
.rng <- \(geo_data) {
    cov <- st_bbox(geo_data)
    xs <- cov[c(1, 3)]; ys <- cov[c(2, 4)]
    x1 <- xs[1] + 4*(xs[2]-xs[1])/8; x2 <- xs[2] - 3*(xs[2]-xs[1])/8
    y1 <- ys[1] + 4*(ys[2]-ys[1])/8; y2 <- ys[2] - 3*(ys[2]-ys[1])/8
    list(box=c(x1, x2, y1, y2), cov=c(xs, ys))
}
geo_data_rng <- .rng(geo_data)

# Subset the GeoJSON to geometries that intersect with the bounding box
bbox_polygon <- st_as_sfc(st_bbox(geo_data_rng$box, crs = st_crs(geo_data)))
subset_data <- geo_data[st_intersects(geo_data, bbox_polygon, sparse = FALSE), ]
subset_data$cell_id <- as.character(subset_data$cell_id)
head(subset_data)
```
Plot cell polygons overlay H&E image for this subset region. Here is an overview of the subset of tissue section that we will visualize on:
```{r plt-subset-he-polygon, fig.width=12, fig.height=4, message=FALSE}
#| code-fold: true
# Plot raster
p0 <- ggplot() +
    annotation_raster(image, 
                      xmin = 0, xmax = info$width, ymin = 0, ymax = info$height) +
    coord_fixed(expand = FALSE,
                xlim = geo_data_rng$box[1:2], ylim = geo_data_rng$box[3:4]) 

# Add sf layer
p <- p0 +
    geom_sf(data = subset_data, fill = NA, color = "white", size = 0.3) + 
    theme(axis.text = element_blank(),
          axis.ticks = element_blank())

# Just polygons
p1 <- ggplot(subset_data) +
    geom_sf(fill = "lightblue", color = "white") +
    theme_void()

# Show plots
(p0 + ggtitle("H&E region") |
    p + ggtitle("Overlayed with cell boundaries") |
    p1 + ggtitle("Cells")) +
    plot_layout(nrow=1) & theme(plot.title = element_text(hjust = 0.5))
```
Note that the analysis steps in bin-wise section - e.g. clustering, calculating reduced dimensions, spatial statistics - can all be migrated to cell-level data here. To avoid redundancy, however, we will use the UMAP reduced dimension and the clustering result from SpaceRanger output for visualization. Note that in your actual analysis pipeline, you should always generate your own clustering and reduced dimension. 

We first read in the clustering result and merge it to the subset `sf` object. We also merge clustering result to the subset of UMAP. [Note that you always need to re-run normalization, PCA, to derive reduced dimensions after subset your data. Here we skipped these steps because our focus is on visualization. You can refer to the previous sections for proper analysis steps]{.aside}
```{r clus-merge-sf-umap-merge-clus, warning=FALSE, message=FALSE}
clus <- read.csv("~/Desktop/clusters.csv")
clus$cell_id <- gsub("^cellid_0*([1-9][0-9]*|0)-1$", "\\1", clus$Barcode)
clus$Cluster <- as.factor(clus$Cluster)

subset_data <- subset_data %>% 
   left_join(clus %>% select(cell_id, Cluster))

umap <- read.csv("~/Desktop/projection.csv")
umap$cell_id <- gsub("^cellid_0*([1-9][0-9]*|0)-1$", "\\1", clus$Barcode)
umap <- umap[umap$cell_id %in% subset_data$cell_id, ]

umap <- umap %>% 
   left_join(clus %>% select(cell_id, Cluster))
```

Now we can visualize clustering result spatially and on UMAP. 
```{r clus-spa-umap, fig.width=7, fig.height=3.5, warning=FALSE, message=FALSE}
#| code-fold: true
# helper
plotPolygons <- function(objsf, annotate = "Cluster"){
    ggplot(objsf) +
      geom_sf(aes(fill = get(annotate)), color = "grey", linewidth = 0.1) +
      labs(fill = annotate) + 
      theme_void() + 
      theme(axis.text = element_blank(),
            axis.ticks = element_blank())
}

pp <- plotPolygons(subset_data, annotate = "Cluster") + 
    scale_fill_manual(values = unname(pals::kelly())) + 
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5)) + 
    ggtitle("Cluster")
pu <- ggplot(umap, aes(x = UMAP.1, y = UMAP.2)) +
    geom_point(aes(color = Cluster), size = 0.5) +
    scale_color_manual(values = unname(pals::kelly())) +
    labs(color = "Cluster") + 
    guides(col=guide_legend(override.aes=list(size=3))) +
    theme_classic()
pp | pu
```

We can also visualize some marker genes, by reading in the the count data first
```{r load-counts-vhdcell, message=FALSE, warning=FALSE}
## VHD cell
countmat_file <- "~/Desktop/filtered_feature_cell_matrix.h5"
vhdcellsce <- DropletUtils::read10xCounts(countmat_file, col.names = TRUE)

# Fix colnames or cell_id in colData
vhdcellsce$cell_id <- colnames(vhdcellsce) <- 
    gsub("^cellid_0*([1-9][0-9]*|0)-1$", "\\1", colnames(vhdcellsce))

# Subset to region in sf
vhdcellsce <- vhdcellsce[, colnames(vhdcellsce) %in% subset_data$cell_id]

# Update gene and cell names
rownames(vhdcellsce) <- rowData(vhdcellsce)$Symbol
rownames(vhdcellsce) <- make.unique(rownames(vhdcellsce))

# Get log normalized counts for plotting
vhdcellsce <- logNormCounts(vhdcellsce)
```

We can merge in gene expression data of selected markers to the subset `sf` object.
```{r plt-mgs-xy, fig.width=10, fig.height=3, message=FALSE}
#| code-fold: true
gs <- c("MMP2", "PIGR", "IGHG1")
ge <- t(logcounts(vhdcellsce)[rownames(vhdcellsce) %in% gs, ]) %>%
    as.data.frame() %>% rownames_to_column(var = "cell_id")
# sanity check of ordering of cell ids
all(ge$cell_id == subset_data$cell_id)

subset_data <- subset_data %>% 
    left_join(ge)

ps <- lapply(gs, \(.) 
             plotPolygons(subset_data, annotate=.))
wrap_plots(ps, nrow=1) & theme(
    legend.key.width=unit(0.5, "lines"),
    legend.key.height=unit(1, "lines")) &
    scale_fill_gradientn(colors=rev(hcl.colors(9, "Rocket")))
```

For a more detailed cell-level analysis on Visium HD, please refer to the EuroBioC2025 OSTA workshop content available at <TODO>.

## Conclusion

In this workflow, we demonstrated how to perform a standard analysis pipeline on a subset of Visium HD data binned at 16 µm and at cell-level. Most methods and visualization techniques developed for Visium or Xenium can be applied to Visium HD. However, more sparsity is expected at bin-level, as each data point (e.g. 2, 8, or 16 µm) has smaller area compared to Visium (55 µm). Each bin cannot be interpreted as a cell, but rather, a segmentation-free approach is applied here. Below are some other resources for Visium HD could be incorporated into your pipeline:

-   [GitHub repository of `bin2cell`](https://github.com/Teichlab/bin2cell) for H&E segmentation and obtaining cell boundaries.

-   [Analysis pipeline](https://www.10xgenomics.com/analysis-guides/segmentation-visium-hd) by 10x Genomics (April 2024). The analysis consists of three steps: run `StarDist` nuclei segmentation in Python, sum gene expression into nuclei based on segmentation mask, and downstream analysis with `scanpy` and `anndata`.

-   [Pipeline](https://github.com/Sanofi-Public/enact-pipeline) on cell typing for Visium HD in Python by Sanofi.


## Appendix

### References {.unnumbered}
