# Processing {#sec-seq-processing}

## Preamble

### Introduction

We apply the methods implemented in the `r BiocStyle::Biocpkg("scater")` [@McCarthy2017] and `r BiocStyle::Biocpkg("scran")` [@Lun2016] packages, which were originally developed for single-cell RNA-seq data, making the assumption here that these methods can be applied to spatial transcriptomics data by treating spots as equivalent to cells.

### Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(scater)
library(scran)
library(SpatialExperiment)
(spe <- readRDS("seq-spe_qc.rds"))
```

## Normalization {#sec-seq-normalization}

### Non-spatial

We use the library size factors methodology, since this is the simplest approach, and can easily be applied to spatial transcriptomics data. Alternative approaches that are popular for single-cell RNA-seq data, including normalization by deconvolution, are more difficulty to justify in the context of spot-based spatial transcriptomics data since (i) spots may contain multiple cells from more than one cell type, and (ii) datasets can contain multiple samples (e.g. multiple Visium slides, resulting in sample-specific clustering).
[Library size-based normalization was originally developed for scRNA-seq data, and makes the assumption that spots are equivalent to cells; this has been shown to be problematic because *"library size confounds [spatial] biology"* [@Bhuva2024].]{.aside}

```{r log, message=FALSE, fig.height=3.5}
# calculate library size factors
spe <- computeLibraryFactors(spe)
summary(sizeFactors(spe))
hist(sizeFactors(spe), breaks=20)
# calculate 'logcounts'; these will be added as an additional 'assay'
spe <- logNormCounts(spe)
assayNames(spe)
```

### Spatially aware

`r BiocStyle::Biocpkg("SpaNorm")` [@Salim2024-SpaNorm] was developed for spatial transcriptomics data specifically, and uses a gene-wise model (e.g., negative binomial) in which variation is decomposed into a library size-dependent (technical) and -indepedent (biological) component.

## Feature selection {#sec-seq-feature-selection}

Here, we apply feature selection methods to identify highly variable genes (HVGs), which will be used as input for downstream analyses such as dimension reduction.
At this stage, we can also apply methods to identify spatially variable genes (SVGs), either instead of or complementary to HVGs; see @sec-crs-spatially-variable.

We use methods from `r BiocStyle::Biocpkg("scran")` [@Lun2016] to identify a set of top highly variable genes (HVGs), which can be used to define major cell types.

[Note that HVGs are defined based only on molecular features (i.e. gene expression), and do not take any spatial information into account.
If the biologically meaningful spatial information in this dataset mainly reflects spatial distributions of cell types, relying on HVGs for downstream analyses may be sufficient.
However, many datasets contain further spatial structure that is not captured in this way, which may be investigated using spatially-aware methods such as identifying spatially variable genes (SVGs).]{.aside}

To identify HVGs, we first remove mitochondrial genes, since these are very highly expressed in this dataset and are not of main biological interest.

```{r mt}
# identify mitochondrial genes
nm <- rowData(spe)$gene_name
mt <- grepl("^MT-", nm, ignore.case=TRUE)
table(mt)
# remove them
spe <- spe[!mt, ]
dim(spe)
```

Then, we apply methods from `r BiocStyle::Biocpkg("scran")`. This gives us a list of HVGs, which can be used for further downstream analyses. The parameter `prop` defines how many HVGs we want. For example `prop = 0.1` returns the top 10% of genes.

```{r hvg, message=FALSE, fig.width=6, fig.height=5}
# fit mean-variance relationship, decomposing 
# variance into technical & biological components
dec <- modelGeneVar(spe)
# select top HVGs (i.e., variance above trend)
length(sel <- getTopHVGs(dec, prop=0.1))
```

```{r plt-hvg, fig.width=5, fig.height=4}
#| code-fold: true
# visualize mean-variance relationship
par(mar=c(4,4,0,0))
fit <- metadata(dec)
plot(fit$mean, fit$var, cex=0.5, xlab="mean expression", ylab="variance")
points(dec[sel, "mean"], dec[sel, "total"], cex=0.5, col="dodgerblue")
curve(fit$trend(x), add=TRUE, lwd=2, col="tomato")
```

## Dimensionality reduction {#sec-seq-dimensionality-reduction}

Apply dimensionality reduction methods to visualize the data and to generate inputs for further downstream analyses.

### Principal component analysis (PCA)

Apply principal component analysis (PCA) to the set of top highly variable genes (HVGs) to reduce the dimensionality of the dataset, and retain the top 50 PCs for further downstream analyses.

This is done for two reasons: (i) to reduce noise due to random variation in expression of biologically uninteresting genes, which are assumed to have expression patterns that are independent of each other, and (ii) to improve computational efficiency during downstream analyses.

We use the computationally efficient implementation of PCA provided in the `r BiocStyle::Biocpkg("scater")` package [@McCarthy2017]. This implementation uses randomization, and therefore requires setting a random seed for reproducibility.

```{r pca}
spe <- runPCA(spe, subset_row=sel)
reducedDimNames(spe)
dim(reducedDim(spe, "PCA"))
```

### Uniform Manifold Approximation and Projection (UMAP)

We also run UMAP on the set of top 50 PCs, and retain the first two UMAP components, which will be used for visualization.

```{r umap}
# compute UMAP (by default, on 1st 50 PCs)
set.seed(123)
spe <- runUMAP(spe, dimred="PCA")
reducedDimNames(spe)
dim(reducedDim(spe, "UMAP"))
# update column names for easier plotting
colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)
```

## Visualization

```{r plt-pcs, fig.height=4}
plotReducedDim(spe, "PCA", color_by="sum") + coord_equal() + theme_minimal()
```

## Appendix

### Saving

```{r}
saveRDS(spe, "seq-spe_dr.rds")
```

### References {.unnumbered}
