# Visium HD workflow {#sec-seq-workflow-visiumhd}

## Introduction

Visium HD is a next-generation spatial transcriptomics technology developed by 10x Genomics, designed to provide single-cell resolution spatial gene expression data across entire tissue sections. Commercially available in 2024 [@Oliveira2024], Visium HD advanced data resolution from 55 µm (spot-based resolution) to bins near subcellular resolution (e.g. 2 µm). In this workflow, we will demonstrate common analysis steps with Visium HD CRC dataset [@Oliveira2024]. 

## Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(OSTA.data)
library(VisiumIO)
library(SpatialExperiment)
library(ggspavis)
library(spacexr)
library(ggplot2)
library(patchwork)
library(DropletUtils)
library(BiocParallel)
library(scran)
library(scater)
library(pheatmap)
library(tidyr)
library(SpotSweeper)
library(Banksy)
library(Statial)
```

In this example of Visium colorectal cancer (CRC) data [@Oliveira2024], we perform Visium cell type deconvolution on 16 µm bins and compare the concordance with the provided Visium deconvolution result on 8 µm bins by 10x Genomics. [Here we only demonstrate the analysis for patient 2. Other patients, such as P1 and P5, also have public data available in CRC condition, with Chromium, Visium HD, and Xenium platforms (link in @sec-bkg-datasets). It would be interesting to investigate the differences in malignant cells/bins (e.g. differentially expressed genes, pathway analysis, etc.) between patients in different platforms.]{.aside}

```{r load-vhd8, message = FALSE, warning=FALSE}
# retrieve dataset from OSF repo
id <- "VisiumHD_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)

# read 008um bins into 'SpatialExperiment'
vhd8 <- VisiumIO::TENxVisiumHD(
  spacerangerOut = td,
  images = "lowres",
  processing = "filtered",
  bin_size = "008",
  format = "h5"
) |> VisiumIO::import()
rownames(vhd8) <- make.unique(rowData(vhd8)$Symbol)

# retrieve 008um bin annotations 
df <- read.csv(file.path(td,"binned_outputs/square_008um/deconvolution.csv.gz"),
               row.names = 1)
df <- df[complete.cases(df), ]
cx <- intersect(colnames(vhd8), rownames(df))

# keep only Visium HD 008um with a provided annotation
vhd8 <- vhd8[, cx]

# add as metadata
colData(vhd8)[names(df)] <- df[cx, ]

vhd8
```

Read in Visium HD data for 16 µm bins. There is no annotation result for 16 $miu$m bin in the public data download. Given the relatively smaller number of bins compared to 8 $miu$m, our downstream analysis will be performed on a subset area of 16 µm bins.

```{r load-vhd16, message = FALSE}
# read 016um bins into 'SpatialExperiment'
vhd16 <- VisiumIO::TENxVisiumHD(
  spacerangerOut = td,
  images = "lowres",
  processing = "filtered",
  bin_size = "016",
  format = "h5"
) |> VisiumIO::import()
rownames(vhd16) <- make.unique(rowData(vhd16)$Symbol)
vhd16
```

```{r vhd816-spavis, fig.width=12, fig.height=6.5}
#| code-fold: true
vhd8$loglibsize <- log(colSums(counts(vhd8)) + 1)
vhd16$loglibsize <- log(colSums(counts(vhd16)) + 1)
plotSpots(vhd8, annotate = "loglibsize", pal = "viridis",
          point_size = 0.005) + ggtitle("8 µm log library size") |
plotSpots(vhd16, annotate = "loglibsize", pal = "viridis",
          point_size = 0.01) + ggtitle("16 µm log library size")
```

Note the Visium HD at 8 and 16 µm share the same full resolution H&E image and the scaling factor. Here we demonstrate analysis steps on a small area that is only 1/64 of the data coverage for 16 µm bins.

```{r plotvis-cove-zoom-8-16, warning=FALSE, message = FALSE, fig.width=12, fig.height=4}
#| code-fold: true
get_range <- function(spe){
  xy <- spatialCoords(spe)*scaleFactors(spe)
  xs <- range(xy[, 1]); ys <- nrow(imgRaster(vhd8))-range(xy[, 2])
  xmi <- xs[1] + 4*(xs[2]-xs[1])/8; xma <- xs[2] - 3*(xs[2]-xs[1])/8
  ymi <- ys[1] + 3*(ys[2]-ys[1])/8; yma <- ys[2] - 4*(ys[2]-ys[1])/8
  return(list(zoom = c(xmi, xma, ymi, yma), cove = c(xs, ys)))
}
vhd8r <- get_range(vhd8)
vhd16r <- get_range(vhd16)

# 8um bounding box, should be similar to 16um 
get_box <- function(cate, ltp, lwd, col){
  box <- geom_rect(
    xmin=vhd8r[[cate]][1], xmax=vhd8r[[cate]][2], 
    ymin=vhd8r[[cate]][3], ymax=vhd8r[[cate]][4],
    col=col, fill=NA, linetype=ltp, linewidth=lwd)
  return(box)
}
box <- get_box(cate="zoom", ltp=4, lwd=2/3, col="black")
box2 <- get_box(cate="cove", ltp=2, lwd=1/2, col="grey")

plotVisium(vhd8, spots=FALSE, point_size=.75) + box + box2 +
  theme(plot.title = element_text(hjust = 0.5)) + 
  ggtitle("grey box: data coverage \n black box: zoomed region") + 
plotVisium(vhd8, point_size=0.9, zoom = TRUE) + xlim(vhd8r[["zoom"]][c(1, 2)]) + ylim(vhd8r[["zoom"]][c(4, 3)]) + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  ggtitle("8 µm bins in black box") + 
  plot_spacer() +
plotVisium(vhd16, point_size=1.8, zoom = TRUE) + xlim(vhd16r[["zoom"]][c(1, 2)]) + ylim(vhd16r[["zoom"]][c(4, 3)]) + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  ggtitle("16 µm bins in black box") + 
plot_layout(nrow=1, widths = c(1.7, 1, 0.05, 1)) & facet_null()
```

Subset the 16 µm object to bins in the black box.

```{r vhd16-subset-zoom, message = FALSE}
get_subset_vhd <- function(spe, spe_range){
  xy <- spatialCoords(spe)*scaleFactors(spe)
  y_no_rev <- nrow(imgRaster(spe)) - spe_range[["zoom"]][c(3,4)]
  vhd16sub <- spe[, xy[, 1] > spe_range[["zoom"]][1] & 
                    xy[, 1] < spe_range[["zoom"]][2] & 
                    xy[, 2] > y_no_rev[1] & 
                    xy[, 2] < y_no_rev[2]]
  return(vhd16sub)
}

vhd16sub <- get_subset_vhd(spe = vhd16, spe_range = vhd16r)
dim(vhd16sub)

# plotVisium(vhd16sub, point_size = 1.8, zoom = TRUE) # sanity check
```

Now we are operating on 2,771 bins with 16 µm resolution.


## Quality control

As detailed in @sec-seq-quality-control, we use SpotSweeper to perform quality control on the subsetted 16 µm Visium HD. 
```{r qc-vhd16-local}
is.mito <- grepl("^MT-", rownames(vhd16sub))
vhd16sub <- scuttle::addPerCellQCMetrics(vhd16sub, subsets=list(Mito=is.mito))

# library size
vhd16sub <- localOutliers(vhd16sub, metric = "sum", direction = "lower", log = TRUE)
# unique genes
vhd16sub <- localOutliers(vhd16sub, metric = "detected", direction = "lower", log = TRUE)
# mitochondrial percent
vhd16sub <- localOutliers(vhd16sub, metric = "subsets_Mito_percent", direction = "higher", log = TRUE)
vhd16sub$discard <- vhd16sub$sum_outliers | vhd16sub$detected_outliers | vhd16sub$subsets_Mito_percent_outliers
```

```{r plt-xy-qc, echo=FALSE, message=FALSE, fig.width=6, fig.height=2.5}
#| code-fold: true
plotSpots(vhd16sub, annotate="sum_log", point_size = 0.2) + 
  ggtitle("log-library size") +
plotSpots(vhd16sub, annotate="subsets_Mito_percent", point_size = 0.2) + 
  ggtitle("% mitochondrial") +
plot_layout(nrow=1) & theme(
    legend.key.width=unit(0.5, "lines"),
    legend.key.height=unit(1, "lines")) &
    scale_color_gradientn(colors=pals::jet())
```

We can visualize the local outliers identified by SpotSweeper:
```{r qc-vis, message=FALSE, fig.width=7.5, fig.height=2.5}
#| code-fold: true
# table(vhd16sub$subsets_Mito_percent_outliers) # NULL
plotSpots(vhd16sub, annotate="sum_outliers") + ggtitle("low_lib_size") +
  plotSpots(vhd16sub, annotate="detected_outliers") + ggtitle("low_n_features") +
  plotSpots(vhd16sub, annotate="discard") + ggtitle("discard") +
  plot_layout(nrow=1, guides = "collect") &
  guides(col=guide_legend(override.aes=list(size=3))) & 
  theme(
    plot.title=element_text(hjust=0.5),
    legend.key.width=unit(0.5, "lines"),
    legend.key.height=unit(1, "lines")) &
  scale_color_manual("discard", values=c("lavender", "purple"))
```

Subset the Visium HD 16 µm object to bins that passed QC:
```{r vhd16-qc-subset}
vhd16sub <- vhd16sub[, !vhd16sub$discard]
dim(vhd16sub)
```

## Clustering

First, we identify and subset to the highly/spatially variable genes in the target area:

```{r vhd16-lognorm-hvg}
vhd16sub <- logNormCounts(vhd16sub)
dec <- modelGeneVar(vhd16sub)
hvg <- getTopHVGs(dec, n = 3000)

# restrict to HVGs; SVGs could also be used
.vhd16sub <- logNormCounts(vhd16sub[hvg, ])
```

As detailed in @sec-crs-clustering, `BANKSY` utilizes a pair of spatial kernels to capture gene expression variation, followed by dimension reduction and graph-based clustering to determine spatial domains.

```{r BANKSY, message=FALSE}
# use weak spatial information
l <- 0.2 
# first order neighbours in VisiumHD
k <- 8
a <- "logcounts"
xy <- c("array_row", "array_col")
# Run PCA on the BANKSY matrix
set.seed(2025)
.vhd16sub <- computeBanksy(.vhd16sub, assay_name=a, coord_names=xy, k_geom=k)
.vhd16sub <- runBanksyPCA(.vhd16sub, lambda=l, npcs=25, seed = 2025)
# Run UMAP for marker gene visualization
set.seed(2025); .vhd16sub <- runUMAP(.vhd16sub)
reducedDim(vhd16sub, "UMAP") <- reducedDim(.vhd16sub, "UMAP")
# Perform BANKSY clustering
.vhd16sub<-clusterBanksy(.vhd16sub,resolution=1.2,lambda=l,npcs=25,seed=2025)
table(vhd16sub$BANKSY <- factor(.vhd16sub[[clusterNames(.vhd16sub)]]))
```

Next, we can perform differential expresison (DE) analysis and find markers for each cluster. Across all cells in each cluster, we obtain the average expression:
```{r deg}
# differential gene expression analysis
mgs <- findMarkers(vhd16sub, groups=vhd16sub$BANKSY, direction="up")
# select for a few markers per cluster
top <- lapply(mgs, \(df) rownames(df)[df$Top <= 3])
top <- unique(unlist(top))

# average expression by clusters
pbs <- aggregateAcrossCells(vhd16sub, 
    ids=vhd16sub$BANKSY, subset.row=top, 
    use.assay.type="logcounts", statistics="mean")
```

The marker genes in each cluster can be visualized with a heatmap:

```{r clus-gene-heatmap, message=FALSE, fig.width=10, fig.height=3.5}
#| code-fold: true
# visualize averages z-scaled across clusters
pheatmap(mat=t(assay(pbs)), scale="column", breaks=seq(-2, 2, length=101), 
         treeheight_row=5, treeheight_col=5)
```

The expression of marker genes in cluster 4, *MMP2* for instance, can be visualized in UMAP and spatially.
```{r spatial-marker-gene, warning=FALSE, message=FALSE, fig.width=6.5, fig.height=6.5}
#| code-fold: true
vhd16sub$log_MMP2 <- logcounts(vhd16sub)["MMP2", ]
(plotUMAP(vhd16sub, colour_by = "BANKSY", point_size = 0.8) +
  theme(legend.position = "none") + plot_spacer() + 
  plotSpots(vhd16sub, annotate = "BANKSY", point_size = 0.8) +
  plot_layout(nrow=1, widths = c(1, 0.1, 1)) &
  scale_color_manual(name = "BANKSY", values = unname(pals::watlington())) &
  theme(aspect.ratio=1, legend.key.size=unit(2/3, "lines"))) /
  (plotUMAP(vhd16sub, colour_by = "log_MMP2", point_size = 0.8) + 
     theme(legend.position = "none") + plot_spacer() + 
  plotSpots(vhd16sub, annotate = "log_MMP2", point_size = 0.8) + 
    theme(plot.title = element_blank()) + 
    plot_layout(nrow=1, widths = c(1, 0.1, 1)) &
  scale_color_viridis_c(option = "viridis", name = "log_MMP2") &
  theme(aspect.ratio=1))
```

## Deconvolution

### Single cell reference

We will perform deconvolution on 16 µm binned Visium HD. First, we retrieve a matching (Chromium) scRNA-seq dataset, which includes low- (`Level1`) and high-resolution (`Level2`) annotations of cells into 9 and 31 subpopulations, respectively. We filter the reference population to patient 2 only, to ensure similar transcriptional profile across technologies.

```{r load-sce, results="hold", message=FALSE, warning=FALSE}
# retrieve dataset from OSF repository
id <- "Chromium_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)

# read into `SingleCellExperiment`
fs <- list.files(td, full.names=TRUE)
h5 <- grep("h5$", fs, value=TRUE)
sce <- read10xCounts(h5, col.names=TRUE)

# add cell metadata
csv <- grep("csv$", fs, value=TRUE)
cd <- read.csv(csv, row.names=1)
colData(sce)[names(cd)] <- cd[colnames(sce), ]

# use gene symbols as feature names
rownames(sce) <- make.unique(rowData(sce)$Symbol)

# exclude cells deemed to be of low-quality
sce <- sce[, sce$QCFilter == "Keep"]

# tabulate subpopulations by disease status
sce$Group <- gsub("P[0-9]", "", sce$Patient)
sce$Patient <- gsub("(P[0-9]).*", "\\1", sce$Patient)
sce <- sce[, sce$Patient == "P2"]
sce
```

::: {.callout-note title="Single cell reference annotation level mapping" collapse="true"}
We can visualize the mapping between the Level 1 and Level 2 annotation in the single cell reference data with a proportion table:

```{r sce-heatmap-level12, fig.width=6.5, fig.height=3.5}
#| code-fold: true
pheatmap(prop.table(table(sce$Level1, sce$Level2), 2),
         cellwidth=10, cellheight=10, treeheight_row=5, treeheight_col=5)
```

Note that `"Proliferating Immune II"` spans evenly between `"B cells"` and `"Tumor"`, so we will make it a separate class `"Prolif. Immune"` in Level 1 annotation.

```{r sce-level1addprolifimmune}
sce$Level1 <- ifelse(sce$Level2 == "Proliferating Immune II", "Prolif. Immune", sce$Level1)
table(sce$Level1) # tabulate low-res. labels
```

There are some differences in the provided deconvolution label (cell type 1 and 2) at 8 µm bin and the Level 2 single cell reference label:

```{r diff-lab1-level2}
setdiff(unique(vhd8$DeconvolutionLabel1), unique(sce$Level2))
setdiff(unique(sce$Level2), unique(vhd8$DeconvolutionLabel1))
```

```{r diff-lab2-level2}
setdiff(unique(vhd8$DeconvolutionLabel2), unique(sce$Level2))
setdiff(unique(sce$Level2), unique(vhd8$DeconvolutionLabel2))
```

These extra classes in Visium HD 8um annotation will be merged to `"Fibroblast"`, `"Myeloid"`, `"Prolif. Immune"`, "`B cells`", and `"Tumor"` as Level 1 annotation, respectively.
:::

### Merge annotation levels for 8 µm

We can merge the provided deconvolution annotation `"DeconvolutionLabel1"` (most likely cell type for singlets or first cell type for doublets) and `"DeconvolutionLabel2"` (the second cell type for doublets) into a lower resolution by the following grouping:

```{r sc-mapping-vhd8-lab1}
df <- data.frame(prop.table(table(sce$Level1, sce$Level2), 2))
colnames(df) <- c("Level1", "Level2", "Freq")
df <- df[df$Freq > 0.4, ]
df <- df[order(df$Level2),]
df <- df[df$Level2 != "Proliferating Immune II", names(df) != "Freq"]
df_map <- data.frame(rbind(apply(df, 2, as.character), 
                           c("Fibroblast", "Proliferating Fibroblast"),
                           c("Myeloid", "Proliferating Macrophages"),
                           c("Myeloid", "cDC I"),
                           c("Tumor", "Tumor IV"),
                           c("Fibroblast", "Vascular Fibroblast"),
                           c("B cells", "Memory B"),
                           c("Prolif. Immune", "NK"),
                           c("Prolif. Immune", "Proliferating Immune II")))

vhd8$deconlab1level1 <- df_map$Level1[match(vhd8$DeconvolutionLabel1, df_map$Level2)]
vhd8$deconlab2level1 <- df_map$Level1[match(vhd8$DeconvolutionLabel2, df_map$Level2)]
table(vhd8$deconlab1level1)
```

```{r sc-mapping-vhd8-lab2}
table(vhd8$deconlab2level1)
```

According to `RCTD`, approximately 75% of the 8 µm bins are singlets.

```{r vhd8-singlets}
print(round(table(vhd8$DeconvolutionClass)["singlet"]/ncol(vhd8), 2))
```

Let us check the top 4 common doublet pairs in the `"doublet_certain"` class:

```{r vhd8-doublets}
dl <- data.frame(colData(vhd8)[vhd8$DeconvolutionClass == "doublet_certain", 
                               c("deconlab1level1", "deconlab2level1")])

combo <- unlist(apply(dl, 1, sort))
combo_wide <- data.frame(deconlab1level1 = combo[seq(1, length(combo), by = 2)],
                         deconlab2level1 = combo[seq(2, length(combo), by = 2)])
rownames(combo_wide) <- rownames(dl)
head(sort(table(paste0(combo_wide$deconlab1level1,
                       "_", combo_wide$deconlab2level1)), 
          decreasing = TRUE), 4)
```

We can see that Myeloid are commonly co-localized with Fibroblast and Tumor cells as doublets.

::: {.callout-note title="Subset 8um bins in black box" collapse="true"}
We can perform the same subsetting procedure to 8 µm bins and compare the difference in number of bins in the zoomed (black box) area.

```{r vhd8-subset-zoom}
# rename column to be consistent with 16 µm
vhd8$deconlevel1 <- vhd8$deconlab1level1
vhd8sub <- get_subset_vhd(spe = vhd8, spe_range = vhd8r)
dim(vhd8sub)
# plotVisium(vhd8sub, point_size = 0.9, zoom = TRUE) # sanity check
```

As expected, the number of bins in 8 µm is 4 times the number of bins in 16 µm. We will visualize on the same zoomed region for both resolutions in any downstream analysis.
:::

### RCTD ("doublet" mode) for 16 µm bins

We also assume there are at most two cell types in a 16 µm bin as in 8 µm, and thus we would expect less singlet proportions.

```{r rctd, message=FALSE, warning=FALSE, results='hide'}
xy <- data.frame(spatialCoords(vhd16sub))
.vhd16sub <- SpatialRNA(xy, counts(vhd16sub))
ids <- factor(sce$Level1)
names(ids) <- colnames(sce)
ref <- Reference(counts(sce), ids)
res <- create.RCTD(.vhd16sub, ref, max_cores=10)
res <- run.RCTD(res, doublet_mode="doublet")
```

Weights inferred by `RCTD` should be normalized such that proportions of cell types sum to 1 in each spot:

```{r dec-res, warning=FALSE, message=FALSE}
# normalize weights to obtain proportions
ws <- normalize_weights(res@results$weights)

# add proportion estimates as spot metadata
ws <- data.frame(as.matrix(ws))
colData(vhd16sub)[names(ws)] <- ws[colnames(vhd16sub), ]
```

Next, we can visualize the deconvolution proportion in space for each cell type:
```{r rctd-vis, message=FALSE, fig.width=12, fig.height=4}
#| code-fold: true
lapply(names(ws), \(.) 
    plotSpots(vhd16sub, annotate=., point_size = 0.05)) |>
    wrap_plots(nrow=2) & theme(
    legend.key.width=unit(0.5, "lines"),
    legend.key.height=unit(1, "lines")) &
    scale_color_gradientn(colors=pals::jet())
```

Obtain the majority vote class for 16 µm bins. 
```{r rctd-majvote, message=FALSE}
# replace the last "." to " " in cell types 
colnames(ws) <- sub("\\.(?=[^.]*$)", " ", colnames(ws), perl = TRUE)

# derive majority vote label
ids <- names(ws)[apply(ws, 1, which.max)]
names(ids) <- rownames(ws)
table(vhd16sub$deconlevel1 <- factor(ids[colnames(vhd16sub)]))

# remove bins with NA label due to not passing RCTD internal QC
vhd16sub <- vhd16sub[, !is.na(vhd16sub$deconlevel1)]
```

::: {.callout-note title="Majorited voted cell type frequency difference between 8 and 16 µm bins" collapse="true"}
We can also check the proportional frequency difference between 8 and 16 µm bins in the zoomed area. 
```{r dens-decon-ct-diff, fig.width=5, fig.height=2.5}
#| code-fold: true
plt_df = data.frame(
  x =table(vhd8sub$deconlevel1)/table(vhd16sub$deconlevel1))
ggplot(plt_df, aes(x = x.Freq)) +
  geom_density() + 
  scale_x_continuous(breaks = seq(0, max(plt_df$x.Freq), by = 2)) + 
  xlab("Cell type frequency ratio") + 
  ggtitle("Distribution of cell type frequency ratio\nbetween 8 and 16 µm bins") + 
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```
Most cell types appear around 3-4 times more in the 8 compared to 16 µm bins, which is expected given the bin size difference. 
:::

### Deconvolution results at different resolutions

Visualize deconvolution result on 8 and 16 µm bins in the zoomed area.

```{r plt-vhd, warning=FALSE, fig.width=8, fig.height=3.6}
#| code-fold: true
plotVisium(vhd8sub, annotate="deconlevel1", zoom =TRUE, point_size=0.8) +
  ggtitle("8 µm") + plot_spacer() + 
  plotVisium(vhd16sub, 
             annotate="deconlevel1", zoom=TRUE, point_size=1.6) +
  ggtitle("16 µm") +
  plot_layout(nrow=1, guides = "collect", widths = c(1, 0.1, 1)) &
  guides(col=guide_legend(override.aes=list(size=3))) &
  theme(legend.key.width=unit(0.5, "lines"),
        plot.title = element_text(hjust = 0.5),
    legend.key.height=unit(1, "lines")) & 
  scale_fill_manual(values = unname(pals::trubetskoy())) & 
  facet_null()
```

## Concordance of clustering and deconvolution on 16 µm

We can visualize the results spatially:

```{r concordance-spatial, fig.width=8, fig.height=3.6}
#| code-fold: true
plotVisium(vhd16sub, annotate = "BANKSY", zoom = TRUE, point_size = 1.6,
           pal = unname(pals::watlington())) +
plotVisium(vhd16sub, annotate = "deconlevel1", zoom = TRUE, point_size = 1.6,
           pal = unname(pals::trubetskoy())) & facet_null()
```

Their concordance can be visualized with a heatmap:

```{r concordance-heatmap, fig.width=2.5, fig.height=3}
#| code-fold: true
pheatmap(prop.table(table(vhd16sub$BANKSY, vhd16sub$deconlevel1), 2),
         cellwidth=10, cellheight=10, treeheight_row=5, treeheight_col=5)
```
From the above heatmap, cluster 1 is mostly immune cells with endothelial; cluster 9 matches B cell bins; cluster 7 and 8 are likely intestinal epithelial and smooth muscle, respectively; fibroblast is common in cluster 4; tumor spans cluster 2, 5, and 6. 


## Neighbourhood analysis

Here, we want to quantify the abundance between *T cells*, *B cells*, or *Myeloid* bins to *Fibroblast* or *Tumor* bins in 16 µm zoomed region. We use the functions `getAbundances()`  from `library(Statial)` to calculate. 

For each bin, we compute the abundance of other cell types within a radius of 200 pixels. The result is stored as `abundances` in the reduced dimension. 
```{r statial-vhd16sub}
spcd <- data.frame(spatialCoords(vhd16sub))
colData(vhd16sub)[names(spcd)] <- spcd
vhd16sub <- getAbundances(vhd16sub, cellType="deconlevel1", imageID="sample_id", 
              spatialCoords = names(spcd), r = 200, nCores = 4)
```

Have a look at the abundances and distance results, and we also append the cell type column to the results. 

```{r statial-vhd16sub-abd}
abd <- reducedDim(vhd16sub, "abundances")
abd$CT <- vhd16sub$deconlevel1
abd[1:5, 1:5]
```
Now, we filter to only *T cells*, *B cells*, or *Myeloid* bins, and investigate within their neighbourhood, the spatial proximity differences to *Fibroblast* or *Tumor* in terms of abundance.

```{r statial-vhd16sub-plt-df}
abd_plt_df <- abd[abd$CT %in% c("T cells", "B cells", "Myeloid"), 
                  c("Tumor", "Fibroblast", "CT")] |>
  pivot_longer(cols = c("Tumor", "Fibroblast"), 
               names_to = "targetCT", values_to = "Abundance")
```

In this region, we observe more Tumor bins compared to Fibroblast bins around each of the immune cell bins. However, note that in this zoomed region, we have nearly 3 times more tumor bins compared to fibroblast bins. Thus, we need to scale the Abundance value by the number of total target cell type bins observed in the region. 

```{r statial-vhd16sub-plt, fig.width=6, fig.height=2.5}
#| code-fold: true
abd_plt_df$Abundance_scaled <- 
  ifelse(abd_plt_df$targetCT == "Tumor",
         abd_plt_df$Abundance/table(vhd16sub$deconlevel1)["Tumor"],
         abd_plt_df$Abundance/table(vhd16sub$deconlevel1)["Fibroblast"])

ggplot(abd_plt_df, aes(x=CT, y=Abundance, fill=targetCT)) + 
    geom_boxplot() + plot_spacer() + 
ggplot(abd_plt_df, aes(x=CT, y=Abundance_scaled, fill=targetCT)) + 
    geom_boxplot() + 
  plot_layout(nrow = 1, guides = "collect", widths = c(1, 0.1, 1)) & 
  scale_fill_manual(values = unname(c("#388E8E", "#8B4513"))) & 
  theme_classic()
```

After correction, we see that the abundance of Fibroblast bins does not differ much from the abundance of Tumor bins around each immune cell types in the zoomed region. 

The same analysis steps can be applied to 8 µm whole area with the provided cell type annotation from RCTD. 

## Conclusion

In this workflow, we demonstrated how to perform standard analysis pipeline on a subset of Visium HD data binned at 16 µm. Most methods and visualization techqniues developed for Visium can be applied to Visium HD. However, more sparsity is expected, as each data point (e.g. 2, 8, or 16 µm) has smaller area compared to Visium (55µm). Each bin cannot be interpreted as a cell, but rather, a cell-segmentation-free approach is applied here. Below are some resources and news for Visium HD to consider and incorporate into your pipeline:

-   Github repository of `bin2cell` for H&E segmentation and obtaining cell boundaries: https://github.com/Teichlab/bin2cell

-   [Analysis pipeline] (https://www.10xgenomics.com/analysis-guides/segmentation-visium-hd) by 10x Genomics (2024 April). The analysis consists of three steps: run `StarDist` nuclei segmentation in Python, sum gene expression into nuclei based on segmentation mask, and downstream analysis with `scanpy` and `anndata`. 

-   10x Genomics [news] (https://www.prnewswire.com/news-releases/10x-genomics-unveils-innovation-roadmap-at-agbt-general-meeting-advancing-leadership-in-large-scale-single-cell-and-spatial-analysis-302382827.html) on morphology-driven segmentation from H&E enabled in CytAssist for Visium HD, by assigning transcripts to individual cells.

-   [Pipeline](https://github.com/Sanofi-Public/enact-pipeline) on cell typing for Visium HD in Python by Sanofi.

## References {.unnumbered}
