# Normalization, feature selection, dimensionality reduction {#sec-seq-norm-feat-dimred}


## Preamble

### Introduction

This chapter demonstrates methods for several data processing steps -- normalization, feature selection, and dimensionality reduction -- that are required before downstream analysis methods can be applied.

We use methods from the `r BiocStyle::Biocpkg("scater")` [@McCarthy2017] and `r BiocStyle::Biocpkg("scran")` [@Lun2016] packages, which were originally developed for single-cell RNA sequencing data, with the (simplified) assumption that these methods can be applied to spot-based spatial transcriptomics data by treating spots as equivalent to cells.

In addition, we discuss alternative spatially-aware methods, which take into account spatial information in spatial transcriptomics data.


### Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(scater)
library(scran)
library(SpatialExperiment)
library(ggspavis)

# load previously saved data
(spe <- readRDS("seq-spe_load.rds"))
```


## Normalization {#sec-seq-normalization}

### Non-spatial

The simplest approach is to calculate log-transformed normalized counts ("logcounts") using library size factors, treating each spot as equivalent to a single cell. This approach is simple and fast, and generally provides a good baseline. However, in some datasets, the (over)-simplified assumption of each spot being equivalent to a single cell will not be appropriate. This example uses the `scater` [@McCarthy2017] and `scran` [@Lun2016] packages.

Note that some alternative non-spatial methods from single-cell workflows (e.g. normalization by deconvolution) are less appropriate for spot-based spatial transcriptomics data, since spots can contain multiple cells from multiple cell types, and datasets can contain multiple samples (e.g. multiple tissue sections, resulting in sample-specific clustering).

[Library size-based normalization was originally developed for scRNA-seq data, and makes the assumption that spots are equivalent to single cells; this has been shown to be problematic because *"library size confounds [spatial] biology"* [@Bhuva2024].]{.aside}

```{r log, message=FALSE, fig.height=3.5}
# calculate library size factors
spe <- computeLibraryFactors(spe)

summary(sizeFactors(spe))
hist(sizeFactors(spe), breaks = 20)

# calculate 'logcounts'; these will be added as an additional 'assay'
spe <- logNormCounts(spe)

assayNames(spe)
```


### Spatially-aware

`r BiocStyle::Biocpkg("SpaNorm")` [@Salim2024-SpaNorm] was developed for spatial transcriptomics data specifically, and uses a gene-wise model (e.g., negative binomial) in which variation is decomposed into a library size-dependent (technical) and -independent (biological) component.


## Feature selection {#sec-seq-feature-selection}

### Non-spatial: Highly variable genes (HVGs)

Here, we re-use feature selection methods from scRNA-seq workflows to identify a set of top 'highly variable genes' (HVGs). These will be used as the input for subsequent steps, including dimensionality reduction and clustering. In this example, we use the `scran` package [@Lun2016] to identify HVGs. As above, this makes the simplified assumption that spots can be treated as equivalent to single cells.

[Note that HVGs are defined based only on molecular features (i.e. gene expression), and do not take any spatial information into account.
If the biologically meaningful spatial information in this dataset mainly reflects spatial distributions of cell types (defined by gene expression), then relying on HVGs for downstream analyses may be sufficient.
However, many datasets contain further spatial structure that is not captured in this way, which may be investigated using spatially-aware methods such as identifying spatially variable genes (SVGs).]{.aside}

To identify HVGs, we first remove mitochondrial genes, since these are very highly expressed in this dataset and are not of main biological interest.

```{r mt}
# identify mitochondrial genes
nm <- rowData(spe)$gene_name
mt <- grepl("^MT-", nm, ignore.case = TRUE)
table(mt)

# remove
spe <- spe[!mt, ]
dim(spe)
```


Next, apply methods from `r BiocStyle::Biocpkg("scran")`. This gives a list of HVGs, which can be used as the input for subsequent steps. The parameter `prop` defines how many HVGs to select. For example `prop = 0.1` returns the top 10%.

```{r hvg, message=FALSE, fig.width=6, fig.height=5}
# fit mean-variance relationship, decomposing 
# variance into technical and biological components
dec <- modelGeneVar(spe)

# select top HVGs (i.e. variance above trend)
sel <- getTopHVGs(dec, prop = 0.1)
length(sel)
```

```{r plt-hvg, fig.width=5, fig.height=4}
#| code-fold: true
# visualize mean-variance relationship
par(mar = c(4, 4, 0, 0))
fit <- metadata(dec)
plot(fit$mean, fit$var, cex = 0.5, 
     xlab = "mean expression", ylab = "variance")
points(dec[sel, "mean"], dec[sel, "total"], 
       cex = 0.5, col = "dodgerblue")
curve(fit$trend(x), add = TRUE, lwd = 2, col = "tomato")
```


### Spatially-aware: Spatially variable genes (SVGs)

Alternatively, spatially-aware methods can instead be used to identify 'spatially variable genes' (SVGs), either instead of or complementary to HVGs.

A number of methods have been developed to identify SVGs. These methods take the spatial information in spatial transcriptomics measurements into account, which can provide a more biologically informative ranking of genes.

For more details and examples, see @sec-crs-spatially-variable.


## Dimensionality reduction {#sec-seq-dimensionality-reduction}

Dimensionality reduction methods are used to generate the inputs for clustering and further downstream analysis steps, as well as for visualization.


### Principal component analysis (PCA)

A standard approach is to apply principal component analysis (PCA) to the set of top HVGs, and retain the top 50 principal components (PCs) for subsequent steps.

This is done for two main reasons: (i) to reduce noise due to random variation in expression of biologically uninformative genes, which are assumed to have expression patterns independent of each other, and (ii) to improve computational efficiency.

We use the fast implementation of PCA provided in the `r BiocStyle::Biocpkg("scater")` package [@McCarthy2017]. This implementation uses randomization, and therefore requires setting a random seed for reproducibility.

```{r pca}
set.seed(123)
spe <- runPCA(spe, subset_row = sel)

reducedDimNames(spe)
dim(reducedDim(spe, "PCA"))
```



### Uniform Manifold Approximation and Projection (UMAP)

The UMAP algorithm is often used for visualization purposes (although interpretation of the UMAP dimensions is difficult and can sometimes be misleading).

Run UMAP on the set of top 50 PCs, retaining the first two UMAP components to be used for visualization purposes.

```{r umap}
# compute UMAP (by default, on top 50 PCs)
set.seed(123)
spe <- runUMAP(spe, dimred = "PCA")

reducedDimNames(spe)
dim(reducedDim(spe, "UMAP"))

# update column names for easier plotting
colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)
```


## Visualizations

Generate a PCA plot using plotting functions from the `r BiocStyle::Biocpkg("scater")` package [@McCarthy2017].

```{r plt-pcs, fig.height=4}
# temporary until QC chapter updated
spe <- addPerCellQC(spe)

plotReducedDim(spe, "PCA", color_by = "sum") + coord_equal() + theme_minimal()
```


Alternatively, use plotting functions from the [ggspavis](https://bioconductor.org/packages/ggspavis) package.

```{r, fig.width=4, fig.height=4}
# top 2 PCs
plotDimRed(spe, plot_type = "PCA")
```

```{r, fig.width=4, fig.height=4}
# top 2 UMAP dimensions
plotDimRed(spe, plot_type = "UMAP")
```


## Appendix

### Saving

Save data object for re-use in later chapters.

```{r}
saveRDS(spe, "seq-spe_dr.rds")
```


### References {.unnumbered}

