# Processing {#sec-seq-processing}

## Preamble

### Introduction

This chapter demonstrates methods for several data processing steps -- normalization, feature selection, and dimensionality reduction -- that are required before downstream analysis methods can be applied.

In addition, we include short demonstrations of subsequent steps, including clustering and identification of marker genes. More details on these downstream steps are included in later parts of the book.

We use methods from the `r BiocStyle::Biocpkg("scater")` [@McCarthy2017] and `r BiocStyle::Biocpkg("scran")` [@Lun2016] packages, which were originally developed for scRNA-seq data, with the (simplified) assumption that these methods can be applied to spot-based ST data by treating spots as equivalent to cells. In addition, we discuss alternative spatially-aware methods, and provide links to later parts of the book where these methods are described in more detail.


### Dependencies

```{r load-libs, message=FALSE, warning=FALSE}
library(SpatialExperiment)
library(STexampleData)
library(scater)
library(scran)
library(BayesSpace)
library(ggspavis)
library(patchwork)
library(pheatmap)
```


### Load data

Here, we import a 10x Genomics Visium dataset that will be used in several of the following chapters. This dataset has previously been preprocessed using data preprocessing procedures with tools outside R and saved in `SpatialExperiment` format, and is available for download from the `r BiocStyle::Biocpkg("STexampleData")` package.

This dataset consists of one sample (Visium capture area) from one donor, consisting of postmortem human brain tissue from the dorsolateral prefrontal cortex (DLPFC) brain region, measured with the 10x Genomics Visium platform. The dataset is described in @Maynard2021.

```{r load-data, message=FALSE}
spe <- Visium_humanDLPFC()

class(spe)
dim(spe)
```


## Normalization {#sec-seq-processing-norm}

### Non-spatial: Library size normalization

The simplest approach for normalization is referred to as "library size normalization", which consists of calculating log-transformed normalized counts ("logcounts") using library size factors, treating each spot as equivalent to a single cell. This approach is simple and fast, and generally provides a good baseline. This can be calculated using the `r BiocStyle::Biocpkg("scater")` [@McCarthy2017] and `r BiocStyle::Biocpkg("scran")` [@Lun2016] packages.

However, note that in some datasets, the simplified assumption of each spot being equivalent to a single cell is not appropriate, and can create problems during analysis. (*"Library size confounds biology in spatial transcriptomics data"* [@Bhuva2024].)

Some alternative methods from non-spatial single-cell workflows (e.g. normalization by deconvolution) are also less appropriate for spot-based ST data, since spots can contain multiple cells from one or more cell types, and datasets can contain multiple samples (e.g. multiple tissue sections), resulting in sample-specific clustering.

```{r log, message=FALSE, fig.height=3.5}
# calculate library size factors
spe <- computeLibraryFactors(spe)

summary(sizeFactors(spe))
hist(sizeFactors(spe), breaks = 20, main = "Histogram of size factors")

# calculate logcounts and store in new assay
spe <- logNormCounts(spe)

assayNames(spe)
```


### Spatially-aware

`r BiocStyle::Biocpkg("SpaNorm")` [@Salim2024-SpaNorm] was developed specifically for ST data, using a gene-wise model (e.g. negative binomial) in which variation is decomposed into a library size-dependent (technical) and -independent (biological) component.


## Feature selection {#sec-seq-processing-feature}

### Non-spatial: Highly variable genes (HVGs)

A simple approach for feature selection is to re-use methods from scRNA-seq workflows to identify a set of "highly variable genes" (HVGs). The set of HVGs will then be used as the input for subsequent steps, such as dimensionality reduction and clustering. In this example, we use the `r BiocStyle::Biocpkg("scran")` package [@Lun2016] to identify HVGs.

Note that HVGs are defined based only on molecular features (i.e. gene expression), and do not take any spatial information into account. If the spatial patterns in gene expression in a dataset mainly reflect spatial distributions of cell types (which are defined by gene expression), then relying on HVGs for downstream analyses may be sufficient. However, if there is further biologically meaningful spatial structure that is not captured in this way, spatially-aware methods may be needed instead.

To identify the set HVGs, we first remove mitochondrial genes, since these tend to be very highly expressed and are not of main biological interest.

```{r mt}
# identify mitochondrial genes
nm <- rowData(spe)$gene_name
mt <- grepl("^MT-", nm, ignore.case = TRUE)
table(mt)

# remove
spe <- spe[!mt, ]
nrow(spe)
```

Next, apply methods from `r BiocStyle::Biocpkg("scran")`. This gives a list of HVGs, which can be used as the input for subsequent steps. The parameter `prop` defines the proportion of HVGs to select -- for example, `prop = 0.1` returns the top 10%. (Alternatively, argument `n` can be used to selected a fixed number of top HVGs, say 1000 or 2000.)

```{r hvgs, message=FALSE, fig.width=6, fig.height=5}
# fit mean-variance relationship, decomposing variance into 
# technical and biological components
dec <- modelGeneVar(spe)

# select top HVGs
sel <- getTopHVGs(dec, prop = 0.1)

# number of HVGs selected
length(sel)
```

```{r plt-hvgs, fig.width=5, fig.height=4}
#| code-fold: true
# visualize mean-variance relationship
par(mar = c(4, 4, 0, 0))
fit <- metadata(dec)
plot(fit$mean, fit$var, cex = 0.5, 
     xlab = "mean expression", ylab = "variance")
points(dec[sel, "mean"], dec[sel, "total"], cex = 0.5, col = "dodgerblue")
curve(fit$trend(x), add = TRUE, lwd = 2, col = "tomato")
```


### Spatially-aware: Spatially variable genes (SVGs)

Alternatively, spatially-aware methods can be used to identify a set of "spatially variable genes" (SVGs), either instead of or complementary to HVGs.

A number of methods have been developed to identify SVGs. These methods take the spatial coordinates of ST measurements into account, which can generate a more biologically informative ranking of genes.

For more details and examples, see @sec-crs-spatially-variable.


## Dimensionality reduction {#sec-seq-processing-dimred}

A standard next step is to perform dimensionality reduction using principal component analysis (PCA), applied to the set of top HVGs. See @sec-crs-dimension-reduction for more details.

In addition, we will perform non-linear dimensionality reduction using the UMAP algorithm, applied to a subset of the top principal components (PCs) (default 50). Note that the UMAP algorithm is stochastic, so the overall visual shapes and layout of the resulting UMAP plot will change from one run to the next. Thus, a random seed should be set to ensure the visualization is reproducible.

<!-- to do: add some details / references on UMAP interpretability issues / difficulties / non-interpretability of UMAP "clusters" -->

```{r pca}
# using functions from scater package
spe <- runPCA(spe, subset_row = sel)
spe <- runUMAP(spe, dimred = "PCA")
```


## Clustering {#sec-seq-processing-clustering}

For more details on clustering, see @sec-crs-clustering.

Here, we will run a spatially-aware clustering algorithm, `r BiocStyle::Biocpkg("BayesSpace")` [@Zhao2021-BayesSpace], which was developed specifically for sequencing-based ST data.

```{r clu, message=FALSE}
# run BayesSpace clustering
.spe <- spatialPreprocess(spe, skip.PCA = TRUE)
.spe <- spatialCluster(.spe, nrep = 1000, burn.in = 100, q = 10, d = 20)

# cluster labels
table(spe$BayesSpace <- factor(.spe$spatial.cluster))
```


## Visualizations

We can visualize the cluster labels in x-y space, alongside the manually annotated reference labels (`ground_truth`) available for this dataset.

```{r plt-xy-clu, fig.width=7, fig.height=2.5, message=FALSE}
# using plotting functions from ggspavis package
# and formatting using patchwork package
plotCoords(spe, annotate = "ground_truth") + 
plotCoords(spe, annotate = "BayesSpace") + 
  plot_layout() & 
  theme(legend.key.size = unit(0, "lines")) & 
  scale_color_manual(values = unname(pals::trubetskoy()))
```

Inspecting the spatial distribution of the top PCs, we can observe that the main driver of expression variability is the distinction between white matter (WM) and non-WM brain layers.

```{r plot-xy-pcs, collapse=TRUE, fig.width=8, fig.height=3}
#| code-fold: true
pcs <- reducedDim(spe, "PCA")
pcs <- pcs[, seq_len(4)]
lapply(colnames(pcs), \(.) {
    spe[[.]] <- pcs[, .]
    plotCoords(spe, annotate = .)
}) |> 
    wrap_plots(nrow = 1) & 
    geom_point(shape = 16, stroke = 0, size = 0.2) & 
    scale_color_gradientn(colors = pals::jet()) & 
    coord_equal() & 
    theme_void() & 
    theme(legend.position = "bottom", 
          plot.title = element_text(hjust = 0.5), 
          legend.key.width = unit(0.8, "lines"), 
          legend.key.height = unit(0.4, "lines"))
```


## Marker genes

Having clustered our spots, we can test for genes that are differentially expressed (DE) between spatial domains. Here, we will use pairwise t-tests and specifically test for upregulation (as opposed to downregulation), i.e. expression should be higher in the cluster for which a gene is reported to be a marker. See @sec-crs-clustering for additional details.

Alternatively, we could also test for spatially variable genes (SVGs) within domains to identify genes with spatial expression patterns that vary independently of the spatial arrangement of domains. See @sec-crs-spatially-variable for additional details.

```{r mgs}
# using scran package
mgs <- findMarkers(spe, groups = spe$BayesSpace, direction = "up")
top <- lapply(mgs, \(df) rownames(df)[df$Top <= 3])
length(top <- unique(unlist(top)))
```

We can visualize selected markers as a heatmap where bins represent the average expression (here, log-transformed library size-normalized counts) of a given gene (= columns) in a given cluster (= rows).
[To visually amplify differences, we use `scale = "column"`. This will, for every gene, subtract the mean and divide by the standard deviation across per-cluster means, bringing genes of potentially very different expression levels to a comparable scale.]{.aside}

```{r plt-mgs-hm, fig.width=9, fig.height=3}
#| code-fold: true
pbs <- aggregateAcrossCells(spe, 
    ids = spe$BayesSpace, subset.row = top, 
    use.assay.type = "logcounts", statistics = "mean")

# use gene symbols as feature names
mtx <- t(assay(pbs))
colnames(mtx) <- rowData(pbs)$gene_name

# using pheatmap package
pheatmap(mat = mtx, scale = "column")
```

In addition, we can plot in x-y space, i.e. coloring spots by their expression of a given marker.

```{r plt-mgs-xy, fig.width=8, fig.height=5.5}
ps <- lapply(
    c("MBP", "PLP1", "NRGN", "SNAP25", "NEFL", "HPCAL1"), \(.) 
    plotCoords(spe, annotate = ., feature_names = "gene_name", 
               assay_name = "logcounts"))

wrap_plots(ps, nrow = 2) & 
  theme(legend.key.width = unit(0.4, "lines"), 
        legend.key.height = unit(0.8, "lines")) & 
  scale_color_gradientn(colors = rev(hcl.colors(9, "Rocket")))
```


## Appendix

### Save object {.unnumbered}

```{r save-data}
colLabels(spe) <- spe$BayesSpace
saveRDS(spe, "seq-spe_cl.rds")
```


### References {.unnumbered}
