# Workflow: Visium CRC {#sec-seq-workflow-visium-crc}

## Preamble

### Introduction

### Dependencies

```{r load-libs, message=FALSE, warning=FALSE}
library(osfr)
library(scran)
library(scater)
library(igraph)
library(scuttle)
library(VisiumIO)
library(ggspavis)
library(patchwork)
library(SpatialExperiment)
```

```{r load-data, message=FALSE, results="hide"}
# retrieve dataset from OSF repo
id <- "https://osf.io/5n4q3"
df <- osf_retrieve_node(id)
id <- "Visium_HumanColon_Oliveria"
fd <- osf_ls_files(df, id)
dir.create(td <- tempfile())
osf_download(fd, td, recurse=TRUE)
```

```{r spe}
# read into 'SpatialExperiment'
obj <- TENxVisium(
    spacerangerOut=td, 
    format="h5", 
    images="lowres")
(spe <- import(obj))
```

```{r plt-hne, echo=FALSE, fig.width=12, fig.height=4.5}
xy <- spatialCoords(spe)*scaleFactors(spe)
xs <- range(xy[, 1]); ys <- nrow(imgRaster(spe))-range(xy[, 2])
box <- geom_rect(
    xmin=xs[1], xmax=xs[2], ymin=ys[1], ymax=ys[2],
    col="black", fill=NA, linetype=2, linewidth=2/3)
plotVisium(spe, spots=FALSE, point_size=.75) + box +
plotVisium(spe, point_size=1.5, zoom=TRUE) +
plot_layout(nrow=1) & facet_null()
```

## Quality control

```{r add-qc}
# use gene symbols as feature names
rownames(spe) <- make.unique(rowData(spe)$Symbol)
# add per-cell quality control metrics
sub <- list(mt=grep("^MT-", rownames(spe)))
spe <- addPerCellQCMetrics(spe, subsets=sub)
```

```{r plt-xy-qc, echo=FALSE, message=FALSE, fig.width=7, fig.height=3}
spe$log_sum <- log1p(spe$sum)
plotSpots(spe, annotate="log_sum") + ggtitle("library size") +
plotSpots(spe, annotate="subsets_mt_percent") + ggtitle("% mitochondrial") +
plot_layout(nrow=1) & theme(
    legend.key.width=unit(0.5, "lines"),
    legend.key.height=unit(1, "lines")) &
    scale_color_gradientn(colors=pals::jet())
```

## Processing

```{r pro}
# log-library size normalization
spe <- logNormCounts(spe)
# highly variable feature selection
tbl <- modelGeneVar(spe)
sel <- getTopHVGs(tbl, n=2e3)
# principal component analysis
spe <- runPCA(spe, subset_row=sel)
# dimensionality reduction
spe <- runUMAP(spe, dimred="PCA")
```

## Clustering

```{r clu}
# build shared nearest-neighbor (SNN) graph
g <- buildSNNGraph(spe, use.dimred="PCA", type="jaccard")
# cluster via Leiden community detection algorithm
k <- cluster_leiden(g, objective_function="modularity", resolution=1)
table(spe$Leiden <- factor(k$membership))
```

## Exploratory

```{r plt-xy-Leiden, fig.width=3.5, fig.height=3}
plotSpots(spe, annotate="Leiden") +
    theme(legend.key.size=unit(0, "lines")) +
    scale_color_manual(values=unname(pals::trubetskoy()))
```

## Signatures

Rather than investigating single genes, we can also evaluate the expression of sets of genes (e.g., pathway signatures); e.g., malignant tissue may differ in metabolic activity such as glycolysis and fatty acid metabolism, or exhibit increased apoptosis (cell death) etc. 

`r BiocStyle::Biocpkg("AUCell")`, 

```{r}

```

## Appendix

### References {.unnumbered}