# Quality control {#sec-seq-quality-control}

## Preamble

Quality control (QC) procedures aim to remove low-quality spots or technical artifacts before further analysis. Low-quality spots can occur due to problems during library preparation or other experimental procedures. Examples include large proportions of dead cells due to cell damage during tissue preparation or low mRNA capture efficiency due to inefficient reverse transcription or PCR amplification. These spots are usually removed prior to further analysis, since otherwise they tend to create problems during downstream analyses such as clustering. For example, problematic spots that are not removed could show up as separate clusters, which may be misidentified as distinct cell types.

Low-quality spots can be identified according to several characteristics, including:

-   library size (i.e. total unique molecular identifier (UMI) counts per spot)

-   number of expressed features (i.e. number of genes with non-zero UMI counts per spot)

-   proportion of reads mapping to mitochondrial genes (a high proportion indicates cell damage)

-   number of cells per spot (unusually high values can indicate problems)

Low library size or low number of expressed features can indicate poor mRNA capture rates, e.g. due to cell damage and missing mRNAs, or low reaction efficiency. A high proportion of mitochondrial reads can be indicative of cell damage, e.g. partial cell lysis leading to leakage and missing cytoplasmic mRNAs, with the resulting reads therefore concentrated on the remaining mitochondrial mRNAs that are relatively protected inside the mitochondrial membrane. Unusually high numbers of cells per spot can indicate problems during cell segmentation.



### QC challenges to sequencing-based SRT data 
The first three characteristics listed above are the standard QC metrics used in scRNA-seq data, and they are often valid for spatial transcriptions as well. However, **caution must be taken in interpreting these metrics as they tend to be confounded by biology in sequencing-based SRTs** [@Bhuva2024, @Totty2024-SpotSweeper]. For example, in the brain, neuronal cell bodies reside in grey matter, whereas white matter is made up almost entirely neuronal processes. This naturally leads to regions of grey matter containing much higher number of genes detected and overall transcripts captured relative to white matter. Moreover, white matter tends to show a higher proportion of mitochondrial reads than grey matter. Without this biological context, one might mistakenly assume that the white matter regions are of low quality and should be removed before downstream analyses.

### Structure of the Chapter

In this chapter, we'll start with introducing some methods to idenitfy low-quality spots using various strategies, inclduing 1) standard methods developed for snRNA-seq via global thresholding, 2) more recent methods that aim to mediates bias in quality control attributed to the spatial confounding. Some diagnositic visualization and strategies are further introduced in [QC Diagnositc](#qc-diagnosis).

In the[advanced topics section](#advanced-topics), we'll visit some novel and unique challenges the quality control of SRT data, including idenitfying novel artefacts, i.e. regnional artifacts and quality control based on number of cells idenitfied in each spot. While these knowledges are every useful, the procedures are not necessarily necessary for every SRT samples, depending on the needs. 

For readers have minimium experience with scRNA-seq data analysis, we recommond to read the remaining of the chapters. For experienced readers, one can start from @spotsweeper_local_outlier section. 

To note, the current Chapter focuses on the quality control at the spot level. In some datasets, it may also be appropriate to apply quality control procedures or filtering at the gene level. For example, certain genes may be biologically irrelevant for downstream analyses. However, here we make a distinction between quality control and feature selection. Removing biologically uninteresting genes (such as mitochondrial genes) may also be considered as part of feature selection, since there is no underlying experimental procedure that has failed. Therefore, we will discuss gene-level filtering in the @seq-normalize-featselect-dimred chapter.


### Dependencies and prerequisite

Load packages for this chapter:

```{r, message=FALSE}
library(SpatialExperiment)
library(ggspavis)
library(scater)
library(scuttle)
library(SpotSweeper)
library(patchwork)
```

Load previously saved data object (see @sec-seq-load-data-save-objects).

```{r, message=FALSE, results='hide'}
spe <- readRDS("seq-spe_load.rds")
```

#### Calculate QC metrics using `scater`

We calculate the QC metrics described above with a combination of methods from the `scater` [@McCarthy2017] package (for metrics that are also used for scRNA-seq data, where we treat spots as equivalent to cells) and our own functions. The QC metrics from `scater` can be calculated and added to the `SpatialExperiment` object's column data using the `addPerCellQC` function. The `sum` column contains the total number of unique molecular identifiers (UMIs) for each spot, the `detected` column contains the number of unique genes detected per spot, and `subsets_mito_percent` contains the percentage (or proportion) of transcripts mapping to mitochondrial genes per spot.

First, we subset the object to keep only spots covered by the tissue section. The remaining spots are background spots, which we are not interested in as they will contain almost entirely mitochondrial genes and transcripts from cellular debris.

```{r}
# subset to keep only spots over tissue
spe <- spe[, colData(spe)$in_tissue == 1]
dim(spe)
```

```{r}
# identify mitochondrial genes
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
table(is_mito)
rowData(spe)$gene_name[is_mito]

# calculate per-spot QC metrics and store in colData
spe <- addPerCellQC(spe, subsets = list(mito = is_mito))
head(colData(spe))
```

## Identifying low-quality spots

### Global outlier detection with fixed threhsolds

Many commonly used QC methods currently applied to SRT were adapted from snRNA-seq workflows, such as **global** **outlier detection**. The simplest option for identifying potentially low-quality spots is to apply fixed thresholds to each QC metrics across the entire sample, and remove any spots that do not meet the thresholds for one or more metrics. For example, we might consider spots to be of low-qualty if they have library sizes below 600 UMI or mitochondrial propotion above 30%. These cutoffs are somewhat arbitrary and often require extensive knowledge of the tissue type and dataset at hand. Exploratory visualizations can be used to help select appropriate thresholds, which may differ depending on the dataset.

Here, we use visualizations to select thresholds for several QC metrics in our human DLPFC dataset: (i) library size, (ii) number of expressed genes, (iii) proportion of mitochondrial reads, and (iv) number of cells per spot.

```{r, fig.height=3.5, fig.width=9}
par(mfrow = c(1, 4))
hist(colData(spe)$sum, xlab = "sum", main = "UMIs per spot")
hist(colData(spe)$detected, xlab = "detected", main = "Genes per spot")
hist(colData(spe)$subsets_mito_percent, xlab = "percent mitochondrial", main = "Percent mito UMIs")
hist(colData(spe)$cell_count, xlab = "number of cells", main = "No. cells per spot")
```

The distributions are relatively smooth, and there are no obvious issues such as a spike at very low library sizes.

We can also plot QC metrics together, such as the library sizes or mitochondrial proportion against the number of cells per spot. This can help us determine an optimal threshold while also checking that we are not inadvertently removing a biologically meaningful group of spots. The horizontal line (argument `threshold`) shows our first guesses at possible filtering thresholds for library size and mitochondrial proportion.

```{r, fig.width=7, fig.height=3.75, message=FALSE}
# plot library size vs. number of cells per spot
p1 <- plotSpotQC(spe, plot_type = "scatter", 
           x_metric = "cell_count", y_metric = "sum", 
           y_threshold = 600) +
  ggtitle("Library size vs. cells per spot")

# plot library size vs. number of cells per spot
p2 <- plotSpotQC(spe, plot_type = "scatter", 
           x_metric = "cell_count", y_metric = "subsets_mito_percent", 
           y_threshold = 30) +
  ggtitle("Mito proportion vs. cells per spot")

p1 | p2
```

The plot shows that these filtering thresholds do not appear to select for any obvious biologically consistent group of spots and will not result in removing an excessive number of observations. We can further verify this by checking the exact number of spots excluded.

```{r}
# select QC threshold for library size
qc_lib_size <- colData(spe)$sum < 600
table(qc_lib_size)

# select QC threshold for library size
qc_detected <- colData(spe)$detected < 400
table(qc_detected)

# select QC threshold for mito proportion
qc_mito_prop <- colData(spe)$subsets_mito_percent> 30
table(qc_mito_prop)
```

Finally, we also check that the discarded spots do not have any obvious spatial pattern that correlates with known biological features. Otherwise, removing these spots could indicate that we have set the threshold too high, and are removing biologically informative spots.

```{r, fig.height=3.5, fig.width=10.5}
# add QC results to colData
spe$qc_lib_size <- qc_lib_size
spe$qc_detected <- qc_detected
spe$qc_mito_prop <- qc_mito_prop

# check spatial pattern of discarded spots
p1 <- plotSpotQC(spe, plot_type = "spot", 
           annotate = "qc_lib_size") +
  ggtitle("Library size (< 600 UMI)")

p2 <-  plotSpotQC(spe, plot_type = "spot", 
           annotate = "qc_detected") +
  ggtitle("Detected genes (< 400 Genes)")

p3 <-  plotSpotQC(spe, plot_type = "spot", 
           annotate = "qc_mito_prop") +
  ggtitle("Mito proportion (> 30%)")

p1 | p2 | p3
```

As an aside, here we can also illustrate what happens if we set the threshold too high. For example, if we set the threshold to 2000 UMI counts per spot -- which may also seem like a reasonable value based on the histogram and scatterplot -- then we see a possible spatial pattern in the discarded spots, matching known cortical layers. This illustrates the importance of interactively checking exploratory visualizations when choosing these thresholds. To illustrate this point, we can plot the manually annotated ("ground truth") DLPFC layers here for reference.

```{r, fig.height=3.5, fig.width=7}
# check spatial pattern of discarded spots if threshold is too high
qc_lib_size_2000 <- colData(spe)$sum < 2000
colData(spe)$qc_lib_size_2000 <- qc_lib_size_2000

# plot the spots flagged with the high threshold
p1 <- plotSpotQC(spe, plot_type = "spot", 
           annotate = "qc_lib_size_2000") +
  ggtitle("Library size (< 2000 UMI)")

# plot reference (manually annotated) layers
p2 <- plotSpots(spe, annotate = "ground_truth", 
          pal = "libd_layer_colors") +
  ggtitle("Manually annotated layers")

p1 | p2
```

Looking at the flagged spots versus the manually annotated layers, it's clear the a setting a library size threshold of 2000 UMIs is flagging more spots in layers 1 and 6 relative to other spatial domains. In other words, library size is confounded with biology, as has been previously demonstrated (@Bhuva2024, @Totty2024-SpotSweeper).

We can additionally use violin plots to visualize the distribution of the QC metrics, with outliers annotated. This can help us to determine if the thresholds we have chosen are appropriate or are excluding too many spots.

```{r, fig.height=3.75}

# library size and outliers
p1 <- plotSpotQC(spe, plot_type="violin", x_metric="sum", 
                 annotate="qc_lib_size", point_size=0.5) +
      xlab("Library size")
  

# detected genes and outliers
p2 <- plotSpotQC(spe, plot_type="violin", x_metric="detected", 
                 annotate="qc_detected", point_size=0.5) +
      xlab("Detected genes") 

# mito percent and outliers
p3 <- plotSpotQC(spe, plot_type="violin", x_metric="subsets_mito_percent", 
                 annotate="qc_mito_prop", point_size=0.5) +
      xlab("Mito proportion") 

# plot with patchwork
p1 | p2 | p3

```

<!--### Global outlier detection using adaptive thresholds

An alternative approach to using fixed thresholds is to use adaptive thresholds that adjusts based on the underling data distribution. These methods assume that the dataset is composed of mostly high-quality observations. The method we will show here is based on median absolute deviations (MAD) from the median value of each metric across all spots. This should ensure that \~99% of non-outlier values will be retained, so long as they follow a normal distribution

As shown above, the distribution for library and detected genes tend to follow a more log-normal distribution, whereas mitochondrial proportion typically follows a normal distribution. For this reason, we will conduct outlier detection using the `isOutlier` function from the `scuttle` package to find spots that are more than 3 MADs below the median of the log-transformed library size and detected genes. For mitochondrial percentage, we will find spots that are 3 MADs above the median of the mitochondrial percentage (without log-transformation).

```{r}
# find 3 MADs outliers
qc_lib_mads <- isOutlier(spe$sum, nmads=3, log=TRUE, type="lower")
qc_detected_mads <- isOutlier(spe$detected, nmads=3, log=TRUE, type="lower")
qc_mito_mads <- isOutlier(spe$subsets_mito_percent, nmads=3, log=FALSE, type="higher")

table(qc_lib_mads)
table(qc_detected_mads)
table(qc_mito_mads)
```

As before, we can use spot plots and violins to determine how well these adaptive thresholds worked.

```{r, fig.height=3.5, fig.width=10.5}
# add QC results to colData
spe$qc_lib_mads <- qc_lib_mads
spe$qc_detected_mads <- qc_detected_mads
spe$qc_mito_mads <- qc_mito_mads

# check spatial pattern of discarded spots
p1 <- plotSpotQC(spe, plot_type = "spot", 
           annotate = "qc_lib_mads") +
  ggtitle("Library size (< 3 MADs)")

p2 <-  plotSpotQC(spe, plot_type = "spot", 
           annotate = "qc_detected_mads") +
  ggtitle("Detected genes (< 3 MADs)")

p3 <-  plotSpotQC(spe, plot_type = "spot", 
           annotate = "qc_mito_mads") +
  ggtitle("Mito proportion (> 3 MADs)")

p1 | p2 | p3

```

```{r, fig.height=3.75}
# library size and outliers
p1 <- plotSpotQC(spe, plot_type="violin", x_metric="sum", 
                 annotate="qc_lib_mads", point_size=0.5) +
      xlab("Library size")
  

# detected genes and outliers
p2 <- plotSpotQC(spe, plot_type="violin", x_metric="detected", 
                 annotate="qc_detected_mads", point_size=0.5) +
      xlab("Detected genes") 

# mito percent and outliers
p3 <- plotSpotQC(spe, plot_type="violin", x_metric="subsets_mito_percent", 
                 annotate="qc_mito_mads", point_size=0.5) +
      xlab("Mito proportion") 

# plot with patchwork
p1 | p2 | p3

```

The main advantage of using adaptive thresholds is that it relieves the user of manually searching for and finding a reasonable fixed threshold (or worse, blindly picking a fixed threshold). However, looking at the spots plots, it appears that global outlier detection using a threshold of 3 MADs is beginning to show spatial patterning in this particular dataset. Users should always carefully inspect their QC results before moving to downstream analyses. Next, we'll look at methods developed to account for biology confound in QC metrics by incorporating spatial information.

-->

### Local outlier detection using SpotSweeper {#spotsweeper_local_outlier}

One of the key assumptions of outlier detection in single cell and spatial transcriptomics is that the QC metrics are independent of natural biology. Otherwise cell/spots with naturally low library size or higher mitochondrial ratio will be more likely to be removed as outliers. For snRNA-seq, this assumption is rarely or only mildly violated, resulting in negligible impacts on downstream analyses. However, as demonstrated above, this assumption is more commonly violated in sequencing-based SRT due to the biological heterogeneity in the tissue being sampled for each spot.

One strategy to address this issue is to look for outliers within their local biological neighborhood. Here we will implement **local outlier detection** using the `localOutliers` function from the [SpotSweeper](https://www.bioconductor.org/packages/release/bioc/html/SpotSweeper.html) package [@Totty2024-SpotSweeper]. This function detects local outliers by comparing the number of unique detected genes, total library size, and mitochondrial percentage of each spot to that of it's nearest neighbors. By default, `localOutliers` uses `k=36` nearest neighbors, which equates to third-order neighbors (i.e., three concentric rings of neighbors around each spot) in Visium's hexagonal spot arrangement. However, for sequencing-based methods that uses square grid arrangements (e.g., STOmics), third-order neighbors would be `k=48`.

The QC metrics for each spot undergo a modified local z-score transformation (two-tailed) based on the median and median absolute deviations of it's neighbors. Here we will use threshold of 3.3 z-scores (`cutoff=3.3`), motivated by the reasoning this corresponding to a p-value of .001. These methods assume a normal distribution, so we’ll use the log-transformed `sum` of the total counts and the log-transformed number of `detected` genes. We will not log-transform the mitochondrial percentage as it tends to follow a normal distribution.

```{r, message=FALSE}
# library size
spe <- localOutliers(spe, metric = "sum", cutoff=3.3, direction = "lower", log = TRUE)

# unique genes
spe <- localOutliers(spe, metric = "detected", cutoff=3.3, direction = "lower", log = TRUE)

# mitochondrial percent
spe <- localOutliers(spe, metric = "subsets_mito_percent", cutoff=3.3, direction = "higher",
                     log = FALSE)
```

Similar to `scater`'s `addPerCellQC` function, the `localOutliers` function adds several columns to the `colData` slot of the `SpatialExperiment` object. The `_outlier` column contains a logical vector indicating whether the spot is an outlier for the respective metric and the `_z` column returns the local z-transformed QC metric. If `log=TRUE`, an additional `_log` column will return the log-transformed metric.

We can then visually confirm that the local outliers detected indeed appear to be outliers by using spot plots of the QC metrics. To do this, we'll visualize the output log2-transformed data (or un-transformed mitochondrial proportion), next to the detected local outliers.

```{r, fig.height=7, fig.width=11}
# spot plot of log-transformed library size
p1 <- plotSpots(spe, annotate="sum_log") +
  ggtitle("log2(Library Size)")

p2 <- plotSpotQC(spe, plot_type="spot", in_tissue="in_tissue",
                 annotate="sum_outliers", point_size=0.2) +
        ggtitle("Local Outliers (Library Size)")


# spot plot of log-transformed detected genes
p3 <- plotSpots(spe, annotate="detected_log") +
  ggtitle("log2(Detected)")
  
p4 <- plotSpotQC(spe, plot_type="spot", in_tissue="in_tissue",
                 annotate="detected_outliers", point_size=0.2) +
        ggtitle("Local Outliers (Detected)")


# spot plot of mitochondrial proportion
p5 <- plotSpots(spe, annotate="subsets_mito_percent") +
  ggtitle("Mito Proportion")
  
p6 <- plotSpotQC(spe, plot_type="spot", in_tissue="in_tissue",
                 annotate="subsets_mito_percent_outliers", point_size=0.2) +
        ggtitle("Local Outliers (Mito Prop)")

# plotting together using patchwork
(p1 / p2) | (p3 / p4) | (p5 / p6)

```

Indeed, it is particularly evident in the log-transformed library size and detected genes that there are clear outliers in the bottom right-hand corner of the sample. We can additionally see that these spots were successfully identified in the bottom row. This "eye test" is a good diagnostic for confirming that local outliers are being accurately detected. We can alternatively visualize the spots that were detected as outliers by visualizing the z-transformed metrics for each spot using violin plots.

```{r, fig.height=3.75}

# z-transformed library size and outliers
p1 <- plotSpotQC(spe, plot_type="violin", x_metric="sum_z", 
                 annotate="sum_outliers", point_size=0.5) +
      xlab("sum_outliers")
  

# z-transformed detected genes and outliers
p2 <- plotSpotQC(spe, plot_type="violin", x_metric="detected_z", 
                 annotate="detected_outliers", point_size=0.5)+
      xlab("detected_outliers")

# z-transformed mito percent and outliers
p3 <- plotSpotQC(spe, plot_type="violin", x_metric="subsets_mito_percent_z", 
                 annotate="subsets_mito_percent_outliers", point_size=0.5)+
      xlab("mito_outliers")

# plot with patchwork
p1 | p2 | p3

```

The spots that were clearly local outliers in library size and number of genes detected display z-scores of up to -10. In other words, these spots deviate up to 10 standard deviations lower than their local mean.

### Local outlier detection using adaptive thresholds

Similar to the global examples above, and as alternative to the fixed threshold of 3.3 z-scores, we can use the `isOutlier` function from `scuttle` to determine outliers based on 3 MADs from the median of the local z-transformed QC metrics. The z-transformed data here already follows a normal distribution, so we set `log=FALSE`.

```{r}
# find 3 MADs outliers
sum_z_mads <- isOutlier(spe$sum_z, nmads=3, log=FALSE, type="lower")
detected_z_mads <- isOutlier(spe$detected_z, nmads=3, log=FALSE, type="lower")
subsets_mito_percent_z_mads <- isOutlier(spe$subsets_mito_percent_z, nmads=3, 
                                         log=FALSE, type="higher")

table(sum_z_mads)
table(detected_z_mads)
table(subsets_mito_percent_z_mads)
```

```{r, fig.height=7, fig.width=11}
# add QC results to colData
spe$sum_z_mads <- sum_z_mads
spe$detected_z_mads <- detected_z_mads
spe$subsets_mito_percent_z_mads <- subsets_mito_percent_z_mads

# spot plot of log-transformed library size
p1 <- plotSpots(spe, annotate="sum_log") +
  ggtitle("log2(Library Size)")

p2 <- plotSpotQC(spe, plot_type="spot", in_tissue="in_tissue",
                 annotate="sum_z_mads", point_size=0.2) +
        ggtitle("Local Outliers (Library Size)")


# spot plot of log-transformed detected genes
p3 <- plotSpots(spe, annotate="detected_log") +
  ggtitle("log2(Detected)")
  
p4 <- plotSpotQC(spe, plot_type="spot", in_tissue="in_tissue",
                 annotate="detected_z_mads", point_size=0.2) +
        ggtitle("Local Outliers (Detected)")


# spot plot of mitochondrial proportion
p5 <- plotSpots(spe, annotate="subsets_mito_percent") +
  ggtitle("Mito Proportion")
  
p6 <- plotSpotQC(spe, plot_type="spot", in_tissue="in_tissue",
                 annotate="subsets_mito_percent_z_mads", point_size=0.2) +
        ggtitle("Local Outliers (Mito Prop)")

# plotting together using patchwork
(p1 / p2) | (p3 / p4) | (p5 / p6)
```

```{r, fig.height=3.75}

# z-transformed library size and outliers
p1 <- plotSpotQC(spe, plot_type="violin", x_metric="sum_z", 
                 annotate="sum_z_mads", point_size=0.5) +
      xlab("sum_outliers")
  

# z-transformed detected genes and outliers
p2 <- plotSpotQC(spe, plot_type="violin", x_metric="detected_z", 
                 annotate="detected_z_mads", point_size=0.5)+
      xlab("detected_outliers")

# z-transformed mito percent and outliers
p3 <- plotSpotQC(spe, plot_type="violin", x_metric="subsets_mito_percent_z", 
                 annotate="subsets_mito_percent_z_mads", point_size=0.5)+
      xlab("mito_outliers")

# plot with patchwork
p1 | p2 | p3
```

We can see that using an adaptive threshold also works well, though the thresholds determined here are slightly lower than 3 z-scores, so more spots are excluded.

### Remove low-quality spots

Now that we have calculated several QC metrics and selected thresholds for each one, we can combine the sets of low-quality spots, and remove them from our object.

We also check again that the combined set of discarded spots does not correspond to any obvious biologically relevant group of spots.

```{r}
# number of discarded spots for each metric
apply(cbind(qc_lib_size, qc_detected, qc_mito_prop#, qc_cell_count
), 2, sum)

# combined set of discarded spots
discard <- qc_lib_size | qc_detected | qc_mito_prop#| qc_cell_count
table(discard)

# store in object
colData(spe)$discard <- discard
```

```{r, fig.height=3.5}
# check spatial pattern of combined set of discarded spots
plotSpotQC(spe, plot_type = "spot", 
           annotate = "discard")
```

```{r}
# remove combined set of low-quality spots
#spe <- spe[, !colData(spe)$discard]
#dim(spe)
```

### Removing local outliers

We can then combine all outliers into a single column called `local_outliers` in the `colData` of the `SpatialExperiment` object.

```{r}
# combine all outliers into "local_outliers" column
spe$local_outliers <- as.logical(spe$sum_outliers) |
    as.logical(spe$detected_outliers) |
    as.logical(spe$subsets_mito_percent_outliers)
```

## QC Diagnostic

### Bivariate QC plot



### Visually Identifying Hangnail Artifacts
Hangnail artifacts are best identified by plotting QC metrics for all samples, as they appear as regions with unusually smooth or low variance mitochondrial percentage. We can observe this in the example below as the oblong piece of tissue on the right side has an usually smooth or "smudged" appearance in the mitochondrial percentage (`subsets_mito_percent`) on the right, but not in the library size (`sum`) on the left.

#### Loading Data and Adding QC metrics

We start by loading an example Visium dataset from SpotSweeper containing a known "hangnail" artifact, and then calculat per-spot QC metrics, such as mitochondrial percentage, as shown above.

```{r, message=FALSE, results='hide'}
# load DLPFC artifact samples
data(DLPFC_artifact)
spe.hangnail <- DLPFC_artifact

# identify mitochondrial genes
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe.hangnail)$gene_name)
table(is_mito)
rowData(spe.hangnail)$gene_name[is_mito]

# calculate per-spot QC metrics and store in colData
spe.hangnail <- addPerCellQC(spe.hangnail, subsets = list(mito = is_mito))
head(colData(spe.hangnail))
```


```{r, fig.height=3.75}

p1 <- plotSpots(spe.hangnail, annotate="sum")
p2 <- plotSpots(spe.hangnail, annotate="subsets_mito_percent")

p1 | p2
```


## Advanced Topics {#spot-qc-adv-topic}
### Artifact detection
#### Hangnail artifact detection with SpotSweeper

When working with spatial transcriptomics datasets, ensuring high-quality data is essential for accurate downstream analyses. One common issue researchers face is the presence of techical artifacts. Here we will demonstrate how detect and remove hangnail artifacts from Visium datasets. Hangnails artifacts arise during tissue preparation, and most commonly occur when when smaller areas, such as distinct brain regions, are dissected from larger structures like whole human brain sections @Totty2024-SpotSweeper.

The use of high-power drills or other tools during dissection can cause mechanical damage, leading to areas with artificially low biological heterogeneity. This manifests as regions with **low variance in QC metrics** like mitochondrial percentage. Such artifacts can significantly impact downstream analyses, such as domain detection, if not properly identified and addressed @Totty2024-SpotSweeper.


#### Quantifying Local Variance in Mitochondrial Percentage

One strategy to detect this artifacts for subsequent removal is to qunaity this loval variance in mitochondrial percentage. Here we will do this using the `localVariance` function in `SpotSweeper`, specifying the QC metric (`subsets_mito_percent`) and the number of neighbors to consider. The default neighbor size of `n_neighbors=36` works well in our hands.

```{r, fig.height=3.75}
spe.hangnail <- localVariance(spe.hangnail, metric = "subsets_mito_percent", 
                              n_neighbors = 36, name = "local_mito_variance_k36"
                              )

plotSpots(spe.hangnail, annotate="local_mito_variance_k36", point_size=1)
```

Indeed, we can see that the hangnail artifact has a much lower local variance in mitochondrial percentage compared to the rest of the tissue section.

#### Classifying Artifacts Using Multiscale Local Variance

After visualizing local variance in mitochondrial percentage, we can move on to classify potential artifact regions using the `findArtifacts` function. This function calculates the local variance (`subsets_mito_ratio`) at multiple scales (`n_order`) while adjusting for mean-variance relationships to reliably detect hangnail artifacts. Here we use `n_order=5` to calculate the local variance using 1st-5th order neighbors. Visium spots are shaped as hexagonal grids, so we specify `shape="hexagonal"` to ensure the correct number of neighbors are used for each order. Use `shape="sqaure"` for data sets with square grid arrangements, such as STEREO-seq or VisiumHD.

```{r, fig.height=3.75}
spe.hangnail <- findArtifacts(spe.hangnail, mito_percent = "expr_chrM_ratio", 
                              mito_sum = "expr_chrM", n_order = 5, 
                              shape = "hexagonal", name = "artifact"
                              )

plotSpotQC(spe.hangnail, plot_type='spot', annotate="artifact")
```

#### Discarding detected artifacts

```{r,}
# removing hangnail artifacts prior to downstream analyses
spe.hangnail <- spe.hangnail[, !colData(spe.hangnail)$artifact]
```

### Number of cells per spot
The number of cells per spot depends on the tissue type and organism.

Here, we check for any outlier values that could indicate problems during cell segmentation.

```{r, fig.height=3.5}
# histogram of cell counts
hist(colData(spe)$cell_count, breaks = 20)

# distribution of cells per spot
tbl_cells_per_spot <- table(colData(spe)$cell_count)
```

We see a tail of very high values, which could indicate problems for these spots. These values are also visible on the scatterplots. Here, we again plot the number of expressed genes vs. cell count, with an added trend.

```{r, fig.width=3.75, fig.height=3.75}
# plot number of expressed genes vs. number of cells per spot
plotSpotQC(spe, plot_type = "scatter", 
           x_metric = "cell_count", y_metric = "detected", 
           x_threshold = 10)
```

In particular, we see that the spots with very high cell counts also have low numbers of expressed genes. This indicates that the experiments have failed for these spots, and they should be removed.

We select a threshold of 10 cells per spot. The number of spots above this threshold is relatively small, and there is a clear downward trend in the number of expressed genes above this threshold.

```{r}
# select QC threshold for number of cells per spot
qc_cell_count <- colData(spe)$cell_count > 10
table(qc_cell_count)

colData(spe)$qc_cell_count <- qc_cell_count
```

```{r, fig.height=3.5}
# check spatial pattern of discarded spots
plotSpotQC(spe, plot_type = "spot", 
           annotate = "qc_cell_count")
```

While there is a spatial pattern to the discarded spots, it does not appear to be correlated with the known biological features (cortical layers). The discarded spots are all on the edges of the tissue. It seems plausible that something has gone wrong with the cell segmentation on the edges of the images, so it makes sense to remove these spots.

```{r}
# number of discarded spots for each metric
apply(cbind(qc_lib_size, qc_detected, qc_mito_prop#, qc_cell_count
), 2, sum)

# combined set of discarded spots
discard <- qc_lib_size | qc_detected | qc_mito_prop#| qc_cell_count
table(discard)

# store in object
colData(spe)$discard <- discard
```

```{r, fig.height=3.5}
# check spatial pattern of combined set of discarded spots
plotSpotQC(spe, plot_type = "spot", 
           annotate = "discard")
```


#### Zero-cell and single-cell spots

A particular characteristic of Visium data is that spots can contain zero, one, or multiple cells.

We could also imagine other filtering procedures such as (i) removing spots with zero cells, or (ii) restricting the analysis to spots containing a single cell (which would make the data more similar to scRNA-seq).

However, this would discard a large amount of information. Below, we show the distribution of cells per spot again (up to the filtering threshold of 10 cells per spot from above).

```{r}
# distribution of cells per spot
tbl_cells_per_spot[1:13]

# as proportions
prop_cells_per_spot <- round(tbl_cells_per_spot / sum(tbl_cells_per_spot), 2)
prop_cells_per_spot[1:13]
```

Only 6% of spots contain a single cell. If we restricted the analysis to these spots only, we would be discarding most of the data.

Removing the spots containing zero cells (2% of spots) would also be problematic, since these spots can also contain biologically meaningful information. For example, in this brain dataset, the regions between cell bodies consists of neuropil (dense networks of axons and dendrites). In our paper [@Maynard2021], we explore the information contained in these neuropil spots.


## Appendix

### Saving {.unnumbered}

Save object(s) in `.rds` format for re-use within later chapters to speed up the build time of the book.

```{r}
# save object(s)
saveRDS(spe, file = "seq-spe_qc.rds") 
```

### Benchmarks {.unnumbered}

<!-- Any benchmark papers to list -->

### References {.unnumbered}
