```{r echo=FALSE}
desc <- packageDescription("OSTA")
```

<i style="color:#0087af">
OSTA v`r desc$Version`, 
compiled on `r format(Sys.time(), "%e %b %Y")`,
using Bioconductor v`r BiocManager::version()`
with R v`r paste(R.version$major, R.version$minor, sep=".")`.
</i>

::: {style="float:right; margin-left:5px; width:32%"}

<!-- get 'images/carousel/*.png' & display them alphabetically, 
flipping with a 3.333 second delay via 'inst/carousel.R' -->

```{r "carousel"}
#| warning: false
#| echo: false
source("carousel.R")
dir <- file.path("images", "carousel")
img <- list.files(dir, full.names=TRUE)
img <- lapply(img, \(.) list(image=.))
carousel("gallery-carousel", 3333, img)
```

:::

# Welcome {-}

This is the website for the online book **Orchestrating Spatial Transcriptomics Analysis with Bioconductor**.

This book provides reproducible examples and discussion on computational analysis workflows for spatial omics data using [Bioconductor](https://bioconductor.org/) in R. The book contains chapters describing individual analysis steps as well as extended workflows, each with examples including R code and datasets. In some examples, R code is also integrated with Python tools.

The book is organized into several parts, consisting of introductory materials, and analysis steps and workflows for the two main streams of spatial omics data: sequencing-based and imaging-based.

Additional materials on analysis workflows for single-cell (non-spatial) data, as well as further introductory materials on R and Bioconductor, can be found in the related book [Orchestrating Single-Cell Analysis with Bioconductor (OSCA)](https://bioconductor.org/books/OSCA/).
